---
title: "Pathway analysis - Network"
author: "Viviana Freire Zapata"
date: "2025"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: '2'
    collapsed: false
    highlight: tango
    css: style.css
editor_options: 
  chunk_output_type: console
---

# Data Pre-processing



# Importing libraries

```{r libraries, message=FALSE, warning=FALSE}
library(KEGGREST)
library(ggkegg)
library(igraph)
library(ggnetwork)
library(furrr)
library(ggh4x)
library(patchwork)
library(readxl)
library(rstatix)
library(tidyverse)
source('scripts/LC_metat_network/network_functions.R')
source('figure_scripts/functions_cdis_exploration.R')
source('figure_scripts/functions_cdis_norm_stats.R')
source('figure_scripts/functions_extra.R')
```


## Loading CID table - ready

```{r}

CID_table <- read_csv("scripts/LC_metat_network/input/CID_combined_table.csv")

```


## ko_pathways from metaG

```{r}

ko_pathways_universe <- read_csv("scripts/LC_metat_network/input/ko_pathways_UPDATE_10_23_24.csv")


```



## Metadata

```{r}
metadata <- read_csv('input/metadata_omics.csv') 

metadata_ready <- metadata %>% 
  mutate(month = factor(month, levels = c("May", "July_1", "July_2", "July_3",
                                          "August", "October", "May_22")))
```


## Taxonomy

```{r}
mag_taxonomy <- read_csv('input/mag_taxonomy_denovo.csv')

```


## MAGs annotations

```{r}

annotations <- read_csv("input/annot_with_cydbs_2.23.25.csv")

```

## geTMM

```{r}
geTMM <- read_csv("output/geTMM_normalized_matrix_2025.csv")

geTMM_long <- geTMM %>% 
  pivot_longer(!Geneid, names_to = "sampleid", values_to = "geTMM")

```


## Calculate expression per Month, ALL KOs ALL MAGs

```{r}
annot_ready <- annotations %>% 
  rename(Gene_id = `...1`) %>% 
  select(Gene_id, fasta, ko_id) %>% 
  left_join(geTMM_long, by = c('Gene_id' = 'Geneid')) %>% 
  filter(!is.na(sampleid),
         !is.na(ko_id)) %>% 
  mutate(sampleid = str_replace_all(sampleid, '-', '_')) %>% 
  inner_join(metadata_ready, by = c('sampleid' = 'treatment_code')) %>% 
  group_by(ko_id, sampleid, month) %>% 
  summarise(geTMM_sum = sum(geTMM)) %>%
  group_by(ko_id, month) %>% 
  summarise(getmm = mean(geTMM_sum)) 

```


# Creating networks

```{r}
getmm_month_table <- annot_ready %>% 
  mutate(ko_id = case_when(ko_id == 'K109441' ~ 'K10944',
                           ko_id == 'K109442' ~ 'K10944',
                           ko_id == 'K109451' ~ 'K10945',
                           ko_id == 'K109452' ~ 'K10945',
                           ko_id == 'K109461' ~ 'K10946',
                           ko_id == 'K109462' ~ 'K10946',
                           TRUE ~ ko_id)) %>% 
  group_by(ko_id, month) %>% 
  summarise(getmm = sum(getmm)) %>% 
  pivot_wider(names_from = month, values_from = getmm)

extra_kos <- annot_ready %>% 
  mutate(ko_id = case_when(ko_id == 'K109441' ~ 'K10944',
                           ko_id == 'K109442' ~ 'K10944',
                           ko_id == 'K109451' ~ 'K10945',
                           ko_id == 'K109452' ~ 'K10945',
                           ko_id == 'K109461' ~ 'K10946',
                           ko_id == 'K109462' ~ 'K10946',
                           TRUE ~ ko_id)) %>% 
  filter(!(ko_id %in% ko_pathways_universe$ko_id)) %>% 
  select(ko_id) %>% 
  distinct() %>% 
  pull()

extra_universe <- tibble(ko_id = extra_kos) %>% 
  mutate(name = NA,
         pathway = NA,
         reaction = NA,
         brite = NA,
         db_links = NA)

for(i in 1:nrow(extra_universe)){
  query <- keggGet(extra_universe$ko_id[i])
  
  extra_universe$name[i] <- ifelse(!is.null(query[[1]]$NAME),
                                   str_c(query[[1]]$NAME, collapse = ';'),
                                   NA)
  
  extra_universe$pathway[i] <- ifelse(!is.null(query[[1]]$PATHWAY),
                                      str_c(query[[1]]$PATHWAY, collapse = ';'),
                                      NA)
  
  extra_universe$reaction[i] <- ifelse(!is.null(query[[1]]$REACTION),
                                       str_c(query[[1]]$REACTION, collapse = ';'),
                                       NA)
  
  extra_universe$brite[i] <- ifelse(!is.null(query[[1]]$BRITE),
                                    str_c(query[[1]]$BRITE, collapse = ';'),
                                    NA)
  
  extra_universe$db_links[i] <- ifelse(!is.null(query[[1]]$DBLINKS),
                                       str_c(query[[1]]$DBLINKS, collapse = ';'),
                                       NA)
}

universe_ready <- rbind(ko_pathways_universe, extra_universe)



network_month <- create_kegg_network(compound_list = CID_table$kegg_cid,
                                     ko_list = universe_ready$ko_id,
                                     ko_abundances = getmm_month_table,
                                     pathways = c('map00920'),
                                     prevalence_per_reaction = 0.5,
                                     expand_search = 1,
                                     generate_plot = TRUE)


graph_df <- create_kegg_network(compound_list = CID_table$kegg_cid,
                                ko_list = universe_ready$ko_id,
                                ko_abundances = getmm_month_table,
                                pathways = c('map00910'),
                                prevalence_per_reaction = 0.5,
                                expand_search = 1,
                                generate_plot = FALSE)

# Exporting table to fill up by hand
#write_csv(graph_df, 'nitrogen_df.csv')

column_names <- names(network_month)

# Importing table with the nodes and edges stablished by hand
new_graph_df <- read_csv('scripts/LC_metat_network/input/nitrogen_df_ellipse_11-2.csv')

getmm_vals <- new_graph_df %>% 
  select(all_of(column_names))

min_getmm <- min(getmm_vals, na.rm = TRUE)
max_getmm <- max(getmm_vals, na.rm = TRUE)

net_plots <- purrr::map(column_names, function(id){
  temp <- new_graph_df %>% 
    # Removing nodes that are not needed
    filter(!(name %in% c('R07316', 'R03546', 'R10092',
                         'C00011', 'C01417', 'C01563',
                         'C00288'))) %>% 
    select(x, y, name, node_type, xend, yend, KEGG_name, 
           pathway, path_name, all_of(id), type) %>% 
    rename(getmm_rx = {{ id }}) 
  
  # Filtering out reactions that are not expressed  
  non_exp_rx <- temp %>% 
    filter(type == 'edge',
           is.na(getmm_rx)) %>% 
    pull(name) %>% 
    unique()
  
  nodes_temp <- temp %>% 
    filter(!(name %in% non_exp_rx)) %>% 
    filter(str_detect(node_type, 'metabolite')) %>% 
    mutate(xmin = x - 120,
           xmax = x + 120,
           ymin = y - 20,
           ymax = y + 20)
  
  plot <- temp %>% 
    filter(!(name %in% non_exp_rx)) %>% 
    ggplot(aes(x = x, y = y, xend = xend, yend = yend))+
    geom_edges(data = . %>% filter(is.na(getmm_rx)),
               color = 'gray20',
               linewidth = .5) +
    geom_edges(data = . %>% filter(!is.na(getmm_rx)),
               aes(color = getmm_rx,
                   linewidth = getmm_rx)) +
    scale_color_viridis_c(option = 'A', direction = -1,
                          limits = c(min_getmm, max_getmm)) +
    labs(color = 'geTMM') +
    ggnewscale::new_scale_color() +
    geom_nodes(data = . %>% filter(str_detect(node_type, 'Reaction')),
               aes(shape = node_type,
                   color = path_name,
                   size = node_type),
               fill = 'white') +
    geom_node_rect(data = nodes_temp,
                   aes(xmin = xmin,
                       xmax = xmax,
                       ymin = ymin,
                       ymax = ymax,
                       fill = node_type),
                   color = 'black',
                   linewidth = .2) +
    geom_nodetext(data = . %>% filter(node_type == 'Undetected metabolite'),
                  aes(label = KEGG_name),
                  size = 2) +
    geom_nodetext(data = . %>% filter(node_type == 'Detected metabolite'),
                  aes(label = KEGG_name),
                  fontface = 'bold',
                  size = 2) +
    geom_nodetext_repel(data = . %>% filter(str_detect(node_type, 'Reaction'),
                                            type == 'node'),
                        aes(label = name),
                        size = 1.7) +
    scale_fill_manual(values = c(colorspace::lighten('indianred2', amount = 0.5), 
                                 'white')) +
    scale_shape_manual(values = c('Detected metabolite' = 19, 
                                  'Undetected metabolite' = 21,
                                  'Reaction present' = 17,
                                  'Reaction absent' = 24)) +
    scale_size_manual(values = c('Detected metabolite' = 2, 
                                 'Undetected metabolite' = 2,
                                 'Reaction present' = 1,
                                 'Reaction absent' = 1)) +
    scale_linewidth_continuous(range = c(.5, 2)) +
    scale_x_continuous(limits = c(-80, 1700)) +
    scale_color_manual(values = 'black') +
    labs(title = id,
         shape = 'Reaction presence',
         fill = 'Metabolite presence') +
    guides(linewidth = 'none',
           size = 'none',
           color = 'none') +
    theme_void() +
    theme(legend.position = "bottom",
          legend.direction = "horizontal",
          legend.box = "horizontal",
          plot.title = element_text(hjust = 0.5, face = 'bold', size = 8),
          legend.spacing.x = unit(0.05, "cm"),
          legend.key.size = unit(0.3, "lines"),
          legend.key.width = unit(1, "lines"),
          legend.text = element_text(size = 5),
          legend.title = element_text(size = 4, face = "bold"),
          plot.margin = margin(0.05, 0.05, 0.05, 0.05, "lines"),
          panel.border = element_blank(),
          panel.spacing = unit(0.2, 'lines')
    )
  
  return(plot)
})

names(net_plots) <- column_names

## Selecting networks of interest to merge

joined_plots <- net_plots$May + net_plots$July_1 + net_plots$October +
  plot_layout(guides = 'collect', nrow = 1) &
  plot_annotation(title = "Metatranscriptomics - Metabolomics")&
  theme(legend.position = 'bottom',
        plot.title = element_text(face = 'bold', hjust = 0.5, size = 8))

net_legend <- ggpubr::get_legend(net_plots$May)

joined_plots


```

# Ammonium and nitrate plots

```{r}

soil_n <- read_xls("input/SPNa_21_22.xls", skip = 6) %>% 
  select(`Sample Location`, `Sample Description #2`, `NO3-N (ppm)`, `NH4-N (ppm)`, `Sum of N (ppm)`) %>% 
  mutate(`NO3-N (ppm)` = as.numeric(str_remove(`NO3-N (ppm)`, '<')),
         `NH4-N (ppm)` = as.numeric(str_remove(`NH4-N (ppm)`, '<')))

soil_n_ready <- soil_n %>% 
  left_join(metadata, by = c('Sample Description #2' = 'treatment_code')) %>% 
  select(!c(SampleID_RP, SampleID_HILIC, SampleID_metaT))



```


## Significant differences - Anova

# Run repeated measures ANOVA - NO3

```{r}

aov_NO3 <- anova_test(data = soil_n_ready, 
                      dv = `NO3-N (ppm)`, 
                      wid = site, 
                      within = month)

get_anova_table(aov_NO3)

# Post-hoc

pairwise_t_test(`NO3-N (ppm)` ~ month, paired = TRUE, 
                p.adjust.method = "fdr", 
                data = soil_n_ready)

```

# Run repeated measures ANOVA - NH4

```{r}

aov_NH4 <- anova_test(data = soil_n_ready, 
                      dv = `NH4-N (ppm)`, 
                      wid = site, 
                      within = month)

get_anova_table(aov_NH4)

# Post-hoc

pairwise_t_test(`NH4-N (ppm)` ~ month, paired = TRUE, 
                p.adjust.method = "fdr", 
                data = soil_n_ready)


```


## Inorganic N plot

```{r}

in_plot <- soil_n_ready %>% 
  mutate(month = factor(month, levels = c("May", "June", "July_1", "July_2",
                                          "July_3", "August",
                                          "September", "October","May_22")),
         NH4_scaled = `NH4-N (ppm)` * (max(`NO3-N (ppm)`, na.rm = TRUE) / max(`NH4-N (ppm)`, na.rm = TRUE))) %>%
  ggplot(aes(x = month)) +
  # NO3-N
  stat_summary(aes(y = `NO3-N (ppm)`), 
               fun = mean, 
               geom = "line", 
               color = "#1f77b4",
               size = 0.8) +
  stat_summary(aes(y = `NO3-N (ppm)`), 
               fun.data = mean_se, 
               geom = "errorbar", 
               width = 0.12, 
               color = "#1f77b4") +
  stat_summary(aes(y = `NO3-N (ppm)`), 
               fun = mean, 
               geom = "point", 
               size = 2, 
               color = "#1f77b4") +
  # NH4-N (right axis)
  stat_summary(aes(y = NH4_scaled), 
               fun = mean, 
               geom = "line", 
               color = "#ff7f0e",
               size = 0.8,
               linetype = "dashed") +
  stat_summary(aes(y = NH4_scaled),
               fun.data = mean_se,
               geom = 'errorbar',
               width = 0.12,
               color = "#ff7f0e") +
  stat_summary(aes(y = NH4_scaled), 
               fun = mean, 
               geom = "point", 
               size = 2, 
               color = "#ff7f0e",
               shape = 17) +
  scale_y_continuous(
    name = "NO₃⁻–N (ppm)",
    sec.axis = sec_axis(~ . * (max(soil_n_ready$`NH4-N (ppm)`, na.rm = TRUE) / 
                                 max(soil_n_ready$`NO3-N (ppm)`, na.rm = TRUE)),
                        name = "NH₄⁺–N (ppm)")) +
  labs(title = 'Inorganic nitrogen')+
  theme_classic(base_size = 6) +
  theme(
    axis.title.y.left = element_text(color = "#1f77b4", face = 'bold', size = 6),
    axis.title.y.right = element_text(color = "#ff7f0e", face = 'bold', size = 6),
    axis.text.y.left = element_text(color = "#1f77b4", size = 5),
    axis.text.y.right = element_text(color = "#ff7f0e", size = 5),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 5),
    axis.title.x = element_blank(),
    axis.line.y.right = element_line(color = "#ff7f0e"),
    axis.line.y.left = element_line(color = "#1f77b4"),
    panel.spacing = unit(0.2, 'lines'),
    plot.title = element_text(face = 'bold', hjust = 0.5, size = 8)
  )



in_plot

```


## Nitrogen related metabolites

# Loading data RP

```{r}
## Raw matrix

RP_df <- read_csv("scripts/LC-MS-MS/1.Pre-processing/output/RP_matrix_clean_2025.csv")


## Loading annotation

annot_rp <- read_csv("scripts/LC-MS-MS/3.Cleanning_Annotation/output/RP_final_annotation_11_20_25.csv")


```

# Loading data HILIC

```{r}
## Raw matrix

HILIC_df <- read_csv("scripts/LC-MS-MS/1.Pre-processing/output/HILIC_matrix_clean_2025.csv")


## Loading annotation

annot_hilic <- read_csv("scripts/LC-MS-MS/3.Cleanning_Annotation/output/HILIC_final_annotation_11_20_25.csv")


```


## Loading significant features

```{r}

sig_features_rp <- read_csv("scripts/LC-MS-MS/5.Metabolite_model/output/RP_sig_features_model_fwd_selected_RP.csv")


sig_features_hilic <- read_csv("scripts/LC-MS-MS/5.Metabolite_model/output/HILIC_sig_features_model_fwd_selected.csv")


```

## Filtering metabolites of interest

```{r}

vip_features_rp <- c('FT_1.419_230.09584', 'FT_1.112_147.07644', 'FT_1.822_277.10307',
                     'FT_2.050_247.12881', 'FT_9.013_261.14449', 'FT_9.313_261.14449',
                     'FT_1.375_276.11905', 'FT_1.420_246.14481', 'FT_1.700_249.10806',
                     'FT_7.543_311.12383', 'FT_6.857_247.12883', 'FT_3.668_182.08119',
                     'FT_1.106_134.04476', 'FT_1.135_120.06548', 'FT_1.080_106.04983',
                     'FT_3.874_132.10187', 'FT_3.589_132.10187', 'FT_6.845_166.08628',
                     'FT_1.774_118.08621', 'FT_1.200_118.08621', 'FT_1.106_90.05498',
                     'FT_1.311_191.10270', 'FT_1.183_176.10298', 'FT_1.295_173.09210',
                     'FT_1.202_166.05326', 'FT_1.412_522.20299', 'FT_1.401_689.21129',
                     'FT_1.354_222.09721', 'FT_1.323_522.20307', 'FT_1.413_522.20300',
                     'FT_1.411_527.15849', 'FT_2.047_204.12298', 'FT_1.745_204.12300')

vip_features_hilic <- c('FT_13.830_274.10456', 'FT_11.478_245.11431', 'FT_10.755_259.12997',
                        'FT_13.751_341.10903', 'FT_14.370_421.06602', 'FT_14.029_549.16735',
                        'FT_14.255_549.16737', 'FT_3.025_147.06635')

```



## Prepare matrix RP

```{r}

RP_ready <- RP_df %>% 
  select(FeatureID, contains("RP")) %>% 
  filter(FeatureID %in% annot_rp$FeatureID) %>% 
  column_to_rownames(var = "FeatureID")

```


# Metabolite Normalization

## Median normalization

```{r}

norm_matrix_rp <- median.norm(RP_ready, transform_data = TRUE)

```


## Filtering full matrix 

```{r}

norm_matrix_sig_rp <- norm_matrix_rp %>% 
  rownames_to_column(var = 'FeatureID') %>% 
  filter(FeatureID %in% vip_features_rp) %>% 
  pivot_longer(!FeatureID, names_to = 'SampleID', values_to = 'norm_int')

```


## Prepare matrix HILIC

```{r}

hilic_ready <- HILIC_df %>% 
  select(FeatureID, contains("HILIC")) %>% 
  filter(FeatureID %in% annot_hilic$FeatureID) %>% 
  column_to_rownames(var = "FeatureID")

```


# Metabolite Normalization

## Median normalization

```{r}

norm_matrix_hilic <- median.norm(hilic_ready, transform_data = TRUE)

```



## Filtering full matrix HILIC

```{r}

norm_matrix_sig_hilic <- norm_matrix_hilic %>% 
  rownames_to_column(var = 'FeatureID') %>% 
  filter(FeatureID %in% vip_features_hilic) %>% 
  pivot_longer(!FeatureID, names_to = 'SampleID', values_to = 'norm_int')


```

## Checking similar sig features between HILIC and RP


```{r}

combined_sig <- rbind(norm_matrix_sig_rp, norm_matrix_sig_hilic)


```


## Adding class info

## HILIC

```{r}

## Loading class of compounds

name_class_hilic <- read_csv("scripts/LC-MS-MS/4.Get_inchis/output/HILIC_classyfire_from_name.csv") %>% 
  filter(FeatureID %in% sig_features_hilic$FeatureID) %>% 
  select(FeatureID, Name, superclass, class, subclass, level_5 = `level 5`) %>% 
  mutate(LC_mode = 'HILIC')

canopus_class_hilic <- read_csv("scripts/LC-MS-MS/4.Get_inchis/output/HILIC_classyfire_canopus.csv") %>% 
  filter(FeatureID %in% sig_features_hilic$FeatureID) %>% 
  select(FeatureID, Name,
         superclass = `ClassyFire#superclass`,
         class = `ClassyFire#class`, 
         subclass = `ClassyFire#subclass`, 
         level_5 = `ClassyFire#level 5`) %>% 
  mutate(LC_mode = 'HILIC')

```

## RP

```{r}

name_class_rp <- read_csv("scripts/LC-MS-MS/4.Get_inchis/output/RP_classyfire_from_name_ready.csv") %>% 
  filter(FeatureID %in% sig_features_rp$FeatureID) %>% 
  select(FeatureID, Name, superclass, class, subclass, level_5 = `level 5`) %>% 
  mutate(LC_mode = 'RP')

canopus_class_rp <- read_csv("scripts/LC-MS-MS/4.Get_inchis/output/RP_classyfire_canopus_ready.csv") %>% 
  filter(FeatureID %in% sig_features_rp$FeatureID) %>% 
  select(FeatureID, Name,
         superclass = `ClassyFire#superclass`,
         class = `ClassyFire#class`, 
         subclass = `ClassyFire#subclass`, 
         level_5 = `ClassyFire#level 5`) %>% 
  mutate(LC_mode = 'RP')

```


## Combine class info in a single df

```{r}

class_info <- rbind(name_class_hilic, canopus_class_hilic, 
                    name_class_rp, canopus_class_rp)

```


## Joining CLASS info to df

```{r}

combined_sig_ready <- combined_sig %>% 
  left_join(class_info, by = c('FeatureID')) %>%
  mutate(Name = case_when(str_detect(Name, 'melitose') ~ 'D-Raffinose',
                          .default = Name),
         Name = str_remove(Name, "\\(\\+\\)-")) %>% 
  group_by(FeatureID) %>% 
  mutate(zscore = (norm_int - mean(norm_int)) / sd(norm_int)) 


```


## Heatmap

## Heatmap

```{r}

metadata_lc <- metadata %>% 
  mutate(sampleid_lc = str_remove_all(SampleID_RP, 'RP_')) %>% 
  filter(!is.na(sampleid_lc))

combined_sig_split <- combined_sig_ready %>% 
  mutate(SampleID = str_remove_all(SampleID, 'RP_'),
         SampleID = str_remove_all(SampleID, 'HILIC_')) %>% 
  left_join(metadata_lc, by = c('SampleID' = 'sampleid_lc')) %>% 
  filter(subclass == 'Amino acids, peptides, and analogues')

keep_n_compounds <- c('"Glu-Tyr', 'gamma-Glutamylglutamate', 'Gln-Glu',
                      'gamma-Glu-leu', 'Glu-Thr', 'Glutamylvaline', 'Gln-Val',
                      'L-Glutamine', 'gamma-Glutamylglutamine', '2,6-Diaminopimelic Acid',
                      'L-Citrulline', 'Sarcosine')


heat_vip_take2 <- combined_sig_split %>%
  filter(Name %in% keep_n_compounds) %>% 
  group_by(subclass, Name, month, LC_mode) %>% 
  summarise(month_int = mean(norm_int)) %>%
  group_by(subclass, Name, LC_mode) %>% 
  mutate(zscore = (month_int - mean(month_int))/ sd(month_int)) %>% 
  ungroup() %>% 
  mutate(month = factor(month, levels = c('May', 'June', 'July_1', 'July_2',
                                          'July_3', 'August', 'September', 'October')),
         subclass = if_else(
           subclass == 'Amino acids, peptides, and analogues',
           'Amino acids, peptides\nand analogues',
           subclass)) %>% 
  ggplot(aes(x = month,
             y = Name,
             fill = zscore)) +
  geom_tile() +
  facet_nested(rows = vars(subclass, LC_mode),
               space = 'free',
               scales = 'free',
               strip = strip_nested(
                 text_y = elem_list_text(
                   angle = c(90, 90, 90)
                 ),
                 size = 'variable'
               ),
               switch = 'y',
               nest_line = element_line(),
               solo_line = TRUE)+
  scale_fill_distiller(palette = 'RdBu', limits = c(-2.5, 2.5))+
  labs(x = 'Month',
       title = 'Metabolomics')+
  theme_bw() +
  theme(
          axis.text.x = element_text(angle = 45, hjust = 1, size = 5),
          axis.text.y = element_text(size = 5.5),
          axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          strip.text.x = element_text(face = 'bold', size = 6),
          strip.text.x.top = element_text(size = 6),
          strip.text.y.left = element_text(size = 6, angle = 90),
          strip.background = element_blank(),
          strip.background.x = element_blank(),
          strip.background.y = element_blank(),
          strip.placement = 'outside',
          panel.grid = element_blank(),
          panel.spacing = unit(0.2, 'lines'),       # reduce spacing between panels
          legend.position = 'bottom',
          legend.direction = 'horizontal',          # compact horizontally
          legend.text = element_text(size = 5),
          legend.title = element_text(size = 6),
          legend.key.size = unit(0.4, 'lines'),     # smaller keys
          legend.spacing.x = unit(0.2, 'lines'),    # horizontal spacing between items
          legend.spacing.y = unit(0.1, 'lines'),    # vertical spacing between items
          legend.box.spacing = unit(0.1, 'lines'),
          plot.title = element_text(face = 'bold', hjust = 0.5, size = 8)
  )


heat_vip_take2


```




## Save final plot

```{r}

## Save final plot

des <- "
AABBCC
FFFFFF
DDDEEE
"

final_plot <- net_plots$May + theme(legend.position = 'none') + 
  net_plots$July_1 + theme(legend.position = 'none') + 
  net_plots$October + theme(legend.position = 'none') + 
  in_plot + patchwork::free(heat_vip_take2) + net_legend +
  plot_layout(design = des, heights = c(2, .1, 1))+
  plot_annotation(tag_levels = list(c('A', '', '', 'B', 'C')),
                  title = "Metatranscriptomics - Metabolomics") &
  theme(plot.tag = element_text(size = 14),
        plot.title = element_text(face = 'bold', hjust = 0.5, size = 8))

final_plot

ggsave("output_figures/Figure_4.png", final_plot, dpi = 300,
       units = 'mm', 
       height = 180, 
       width = 180)

ggsave("output_figures/Figure_4.svg", final_plot, dpi = 300,
       units = 'mm', 
       height = 180, 
       width = 180)

```


## Amino acids


```{r}


graph_df_aa <- create_kegg_network(compound_list = CID_table$kegg_cid,
                                   ko_list = universe_ready$ko_id,
                                   ko_abundances = getmm_month_table,
                                   pathways = c('map01230'),
                                   prevalence_per_reaction = 0.5,
                                   expand_search = 0,
                                   generate_plot = TRUE)



## Selecting networks of interest to merge

joined_plots <- graph_df_aa$May + graph_df_aa$July_1 + graph_df_aa$October +
  plot_layout(guides = 'collect', nrow = 1) &
  plot_annotation(title = "Metatranscriptomics - Metabolomics")&
  theme(legend.position = 'bottom',
        legend.key.size = unit(0.4, "cm"),   # smaller boxes
        legend.text = element_text(size = 6), # smaller text
        legend.title = element_text(size = 6),
        plot.title = element_text(face = 'bold', hjust = 0.5, size = 8))


joined_plots




ggsave("output_figures/Supp_Figure_10.png", joined_plots, dpi = 300,
       units = 'mm', 
       height = 180, 
       width = 250)

ggsave("output_figures/Supp_Figure_10.svg", joined_plots, dpi = 300,
       units = 'mm', 
       height = 180, 
       width = 250)


```

