---
title: "Pathway analysis"
author: "Viviana Freire Zapata"
date: "2026"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: '2'
    collapsed: false
    highlight: tango
    css: style.css
editor_options: 
  chunk_output_type: console
---

# Data Pre-processing



# Importing libraries

```{r libraries, message=FALSE, warning=FALSE}

library(ggh4x)
library(patchwork)
library(taxonomizr)
library(KEGGREST)
library(tidyverse)

```


## Loading CID table - ready

```{r}

CID_table <- read_csv("scripts/LC_metat_network/input/CID_combined_table.csv") %>% 
  filter(!kegg_cid %in% c('D00098', 'D04334', 'D08673', 'D10610'))

```


## ko_pathways from metaG

```{r}

ko_pathways_universe <- read_csv("scripts/LC_metat_network/input/ko_pathways_UPDATE_10_23_24.csv")


```

## Metadata

```{r}
metadata <- read_csv('input/metadata_omics.csv') 

metadata_ready <- metadata %>% 
  mutate(month = factor(month, levels = c("May", "July_1", "July_2", "July_3",
                                          "August", "October", "May_22")))
```


## Taxonomy

```{r}
mag_taxonomy <- read_csv('input/mag_taxonomy_denovo.csv')

```


## MAGs annotations

```{r}

annotations <- read_csv("input/annot_with_cydbs_2.23.25.csv") %>% 
  mutate(ko_id = str_replace(ko_id, 'K109441', 'K10944'),
         ko_id = str_replace(ko_id, 'K109442', 'K10944'),
         ko_id = str_replace(ko_id, 'K109451', 'K10945'),
         ko_id = str_replace(ko_id, 'K109452', 'K10945'),
         ko_id = str_replace(ko_id, 'K109461', 'K10946'))

```

## geTMM

```{r}
geTMM <- read_csv("output/geTMM_normalized_matrix_2025.csv")

geTMM_long <- geTMM %>% 
  pivot_longer(!Geneid, names_to = "sampleid", values_to = "geTMM")

```


## Calculate expression per Month, ALL KOs ALL MAGs

```{r}

annot_ready <- annotations %>% 
  rename(Gene_id = `...1`) %>% 
  select(Gene_id, fasta, ko_id) %>% 
  left_join(geTMM_long, by = c('Gene_id' = 'Geneid')) %>% 
  filter(!is.na(sampleid),
         !is.na(ko_id)) %>% 
  mutate(sampleid = str_replace_all(sampleid, '-', '_')) %>% 
  inner_join(metadata_ready, by = c('sampleid' = 'treatment_code')) %>% 
  group_by(ko_id, sampleid, month) %>% 
  summarise(geTMM_sum = sum(geTMM)) %>%
  group_by(ko_id, month) %>% 
  summarise(getmm = mean(geTMM_sum)) 

```

## Loading KEGG tables

```{r}


## Module -- KO

module_ko <- keggLink('module', 'ko')

## Creating table with KO, RX, Compound

table_module <- tibble(module_id = module_ko,
                       ko_id = names(module_ko)) %>% 
  mutate(module_id = str_remove(module_id, 'md:'),
         ko_id = str_remove(ko_id, 'ko:'))


## Module -- Compound


module_compound <- keggLink('module', 'compound')

## Creating table with KO, RX, Compound

table_module_compound <- tibble(module_id = module_compound,
                                C_id = names(module_compound)) %>% 
  mutate(module_id = str_remove(module_id, 'md:'),
         C_id = str_remove(C_id, 'cpd:'))

```


## Joining my annotations

# Retrieving names from modules


```{r}


module_list <- keggList('module')

module_names <- data.frame(module_names = module_list,
                           module_id = names(module_list))

```


```{r}

mag_ko_unique <- annotations %>% 
  select(ko_id) %>% 
  distinct()

module_count_tb <- table_module %>% 
  group_by(module_id) %>% 
  mutate(kegg_count = n()) %>% 
  right_join(mag_ko_unique, by = 'ko_id') %>% 
  group_by(module_id, kegg_count) %>% 
  summarise(mag_ko_count = n()) %>% 
  inner_join(module_names, by = 'module_id')

## Compound

module_count_tb_compound <- table_module_compound %>% 
  group_by(module_id) %>% 
  mutate(kegg_count_cid = n()) %>% 
  right_join(CID_table, by = c('C_id' = 'kegg_cid')) %>% 
  group_by(module_id, kegg_count_cid) %>% 
  summarise(arid_cid_count = n()) %>% 
  inner_join(module_names, by = 'module_id')

## Final metaG based only

final_module_metaG <- full_join(module_count_tb, module_count_tb_compound,
                                by = c('module_id', 'module_names')) %>% 
  mutate(ko_pct = mag_ko_count / kegg_count * 100,
         cpd_pct = arid_cid_count / kegg_count_cid * 100) %>% 
  mutate(across(c(ko_pct, cpd_pct, mag_ko_count, arid_cid_count),
                ~ ifelse(is.na(.x), 0, .x))) %>% 
  select(module_id, module_names, kegg_count, mag_ko_count, ko_pct,
         kegg_count_cid, arid_cid_count, cpd_pct)

```

## Adding metaT

```{r}

## By month

metat_ex_ready <- annot_ready %>% 
  filter(!is.na(ko_id)) %>% 
  filter(getmm > 0) %>% 
  select(ko_id, month) %>% 
  distinct()

module_count_tb_metat <- table_module %>% 
  group_by(module_id) %>% 
  mutate(kegg_count = n()) %>% 
  right_join(metat_ex_ready, by = 'ko_id') %>% 
  group_by(module_id, kegg_count, month) %>% 
  summarise(ko_expression = n()) %>% 
  inner_join(module_names, by = 'module_id')

## Final table

final_table_omic <- final_module_metaG %>% 
  full_join(module_count_tb_metat) %>% 
  mutate(metat_pct = ko_expression / kegg_count * 100) %>% 
  mutate(across(c(metat_pct, ko_expression),
                ~ ifelse(is.na(.x), 0, .x)))

  
```

## Module classification

```{r}


ko_median <- median(final_module_metaG$ko_pct[final_module_metaG$ko_pct > 0], 
                    na.rm = TRUE)

cpd_median <- median(final_module_metaG$cpd_pct[final_module_metaG$cpd_pct > 0], 
                    na.rm = TRUE)

module_classified <- final_module_metaG %>%
  ungroup() %>% 
  
  # Remove rows with missing coverage
  filter(!is.na(ko_pct), !is.na(cpd_pct)) %>%
  
  # Classify modules
  mutate(
    gene_class = ifelse(ko_pct >= ko_median, "High_gene", "Low_gene"),
    met_class  = ifelse(cpd_pct >= cpd_median, "High_met", "Low_met"),
    concordance = case_when(
      gene_class == "High_gene" & met_class == "High_met" ~ "High-High (Concordant)",
      gene_class == "High_gene" & met_class == "Low_met"  ~ "HighGene-LowMet",
      gene_class == "Low_gene"  & met_class == "High_met" ~ "LowGene-HighMet",
      TRUE ~ "Low-Low"
    )
  )


```


## plot

```{r}

vip_list <- c('M00009', 'M00020', 'M00844', 'M00009', 
              'M00004', 'M00555', 'M00176', 'M00529', 'M00565', 'M00528')

scatter_modules <- module_classified %>%
  mutate(
    module_names = case_when(
      str_detect(module_names, "Citrate cycle") ~ "TCA",
      str_detect(module_names, "Nitrification, ammonia") ~ "Nitrification",
      str_detect(module_names, "Denitrification, nitrate") ~ "Denitrification",
      str_detect(module_names, "Pentose phosphate pathway") ~ "Pentose phosphate pathway",
      str_detect(module_names, "Assimilatory sulfate reduction") ~ "Assimilatory sulfate reduction",
      str_detect(module_names, "Trehalose biosynthesis") ~ "Trehalose biosynthesis",
      str_detect(module_names, "Serine biosynthesis") ~ "Serine biosynthesis",
      TRUE ~ module_names
    )
  ) %>% 
  select(module_id, module_names, ko_pct, cpd_pct, concordance) %>%
  distinct() %>%
  mutate(vip = ifelse(module_id %in% vip_list, module_names, NA)) %>%
  
  ggplot(aes(x = ko_pct,
             y = cpd_pct,
             color = concordance)) +
  geom_point(size = 2, alpha = 0.8) +
  ggrepel::geom_label_repel(
    aes(label = vip),
    size = 2,
    show.legend = FALSE,
    max.overlaps = Inf,
    nudge_x = 1,
    nudge_y = 10,
    segment.color = "grey50",
    segment.size = 0.3
  ) +
  scale_color_manual(values = ggpubr::get_palette(palette = 'Set2', 4))+
  labs(
    title = "KEGG Module vs Compound Coverage",
    x = "KO Percentage (%)",
    y = "Compound Percentage (%)"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 8),
    axis.title = element_text(face = "bold", size = 6),
    legend.position = "none",
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    axis.title.y = element_text(size = 7),
    panel.grid = element_blank(),
    plot.margin = margin(0, 0, 0, 0)
  )

scatter_modules


```

## MetaCyc

```{r}

arid_rx_meta <- read_csv("scripts/LC_metat_network/output/MetaCyc_arid_reaction.csv")


arid_path_meta <- read_csv("scripts/LC_metat_network/output/MetaCyc_arid_pathway.csv")

arid_path_metaG <- read_csv("scripts/LC_metat_network/output/MetaCyc_arid_pathway_counts.csv")

arid_comp_meta <- read_csv("scripts/LC_metat_network/output/MetaCyc_arid_compounds.csv")

```

## Calculate pathway vs compound coverage

```{r}

pathway <- read_tsv("input/MetaCyc_db_2026/All-pathways-of-MetaCyc.txt") %>% 
  mutate(`KEGG pathway` = str_extract_all(`KEGG pathway`, "(?i)(map|ko|rn)\\d{5}(?=[^0-9]*$)"))

pathway_name <- read_tsv("input/MetaCyc_db_2026/All-pathways-of-MetaCyc_path_name.txt") %>% 
  select(Pathways, `Common-Name`, `Engineered?`, Species)

reaction <- read_tsv("input/MetaCyc_db_2026/All-reactions-of-MetaCyc.txt")

compounds <- read_tsv("input/MetaCyc_db_2026/All-compounds-of-MetaCyc_extra.txt")

```

## Counts MetaCyc

```{r}

## Create pathway - reaction table

path_to_reaction <- pathway %>% 
  select(Pathways, `Reactions of pathway`, `KEGG pathway`) %>% 
  separate_rows(`Reactions of pathway`, sep = " // ") %>% 
  distinct()

path_rx_count <- path_to_reaction %>% 
  select(!`KEGG pathway`) %>% 
  group_by(Pathways) %>% 
  mutate(number_reactions = n()) %>% 
  distinct()

## Compounds per pathway

compound_path <- compounds %>% 
  filter(!is.na(`Pathways of compound`)) %>% 
  select(Compound, `Pathways of compound`) %>% 
  separate_rows(`Pathways of compound`, sep = " // ") %>%
  distinct() %>% 
  group_by(`Pathways of compound`) %>% 
  mutate(number_compounds = n())

```


## Matching arid

```{r}

## Compound

arid_compound_cov <- arid_comp_meta %>% 
  select(Compound, `Pathways of compound`, kegg_cid) %>% 
  distinct() %>% 
  group_by(`Pathways of compound`) %>% 
  mutate(number_compounds_arid = n())


## Ready 

compound_coverage <- compound_path %>% 
  inner_join(arid_compound_cov, by = c('Compound', 'Pathways of compound')) %>% 
  mutate(cpd_pct = number_compounds_arid / number_compounds * 100)

## Pathway

arid_path_metaG_ft <- arid_path_metaG %>% 
  rename(number_reactions_arid = n_reaction) %>% 
  distinct()


path_coverage <- path_rx_count %>% 
  inner_join(arid_path_metaG_ft, by = c('Pathways', 'Reactions of pathway' = 'Reaction')) %>%
  select(!`Reactions of pathway`) %>% 
  distinct() %>% 
  mutate(pathway_pct = number_reactions_arid / number_reactions * 100) %>% 
  left_join(pathway_name, by = 'Pathways') %>% 
  filter(is.na(`Engineered?`))

```


## Joining tables

```{r}

path_ready <- path_coverage %>% 
  select(Pathways, pathway_pct, `Common-Name`, Species)

compound_ready <- compound_coverage %>% 
  select(Compound, `Pathways of compound`, cpd_pct) %>% 
  distinct()

metacyc_final <- path_ready %>% 
  full_join(compound_ready, by = c('Pathways' = 'Pathways of compound')) %>% 
  mutate(across(c(pathway_pct, cpd_pct),
                ~ ifelse(is.na(.x), 0, .x))) 

path_tax_ft <- metacyc_final %>% 
  separate_rows(Species, sep = " // ") %>% 
  filter(str_detect(Species, 'TAX')) %>% 
  mutate(Species = str_remove_all(Species, 'TAX-'))

```

## Keeping only bacteria and archaea paths

```{r}


# Run only first time
# prepareDatabase(
#   sqlFile = "ncbi_taxonomy.sqlite",
#   getAccessions = FALSE
# )

taxids <- unique(path_tax_ft$Species)

lineage_dia <- getTaxonomy(
  taxids,
  sqlFile = "C:/Users/vivia/Escritorio/Arid_metaT-LCMS_2025/Arid_2_code/ncbi_taxonomy.sqlite")

lineage_df <- as.data.frame(lineage_dia) %>%
  tibble::rownames_to_column("staxids") %>% 
  mutate(staxids = as.character(staxids)) %>% 
  mutate(staxids = str_trim(staxids))

## Final table


final_path <- path_tax_ft %>% 
  left_join(lineage_df, by = c('Species' = 'staxids')) %>% 
  filter(domain != 'Eukaryota')


```



## Module classification

```{r}


path_median <- median(final_path$pathway_pct[final_path$pathway_pct > 0], 
                    na.rm = TRUE)

cpd_median <- median(final_path$cpd_pct[final_path$cpd_pct > 0], 
                    na.rm = TRUE)

path_classified <- final_path %>%
  ungroup() %>% 
  
  # Remove rows with missing coverage
  filter(!is.na(pathway_pct), !is.na(cpd_pct)) %>%
  
  # Classify modules
  mutate(
    path_class = ifelse(pathway_pct >= path_median, "High_gene", "Low_gene"),
    met_class  = ifelse(cpd_pct >= cpd_median, "High_met", "Low_met"),
    concordance = case_when(
      path_class == "High_gene" & met_class == "High_met" ~ "High-High (Concordant)",
      path_class == "High_gene" & met_class == "Low_met"  ~ "HighGene-LowMet",
      path_class == "Low_gene"  & met_class == "High_met" ~ "LowGene-HighMet",
      TRUE ~ "Low-Low"
    )
  )


```


## plot

```{r}

vip_list2 <- c('SERSYN-PWY', 'TRESYN-PWY', 'SO4ASSIM-PWY', 'PENTOSE-P-PWY', 
              'AMMOXID-PWY', 'TCA', 'DENITRIFICATION-PWY', 'BETSYN-PWY', 
              'PWY-3722', 'ARGSYN-PWY')

scatter_meta <- path_classified %>%
  select(Pathways, `Common-Name`, pathway_pct, cpd_pct, concordance) %>%
  distinct() %>%
  mutate(vip = ifelse(Pathways %in% vip_list2, `Common-Name`, NA)) %>%
  ggplot(aes(x = pathway_pct,
             y = cpd_pct,
             color = concordance)) +
  geom_point(size = 2, alpha = 0.8) +
  ggrepel::geom_label_repel(
    aes(label = vip),
    size = 2,
    show.legend = FALSE,
    max.overlaps = Inf,
    nudge_x = 1,
    nudge_y = 10,
    segment.color = "grey50",
    segment.size = 0.3
  ) +
  scale_color_manual(values = ggpubr::get_palette(palette = 'Set2', 4))+
  labs(
    title = "MetaCyc pathway vs Compound Coverage",
    x = "Pathway Percentage (%)",
    y = "Compound Percentage (%)"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 8),
    axis.title = element_text(face = "bold", size = 6),
    legend.position = "bottom",
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    axis.title.y = element_text(size = 7),
    panel.grid = element_blank(),
    legend.text = element_text(size = 5),
    legend.title = element_text(size = 6),
    legend.box.spacing = unit(.05, 'lines'),
    legend.key.size = unit(.5, 'lines'),
    plot.margin = margin(0, 0, 0, 0)
  )

scatter_meta


```



##Save

```{r}

joined_plots <- scatter_modules + scatter_meta +
  plot_layout(nrow = 2) &
  plot_annotation(title = "Metatranscriptomics - Metabolomics")&
  theme(plot.title = element_text(face = 'bold', hjust = 0.5, size = 8))

joined_plots

ggsave("output_figures/Supp_Figure_2.png", joined_plots, dpi = 300,
       units = 'mm', 
       height = 180, 
       width = 250)



```

