---
title: "Linear mixed effect model - metabolites - RP"
author: "Viviana Freire Zapata"
date: "9/30/2025"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: '2'
    collapsed: false
    highlight: tango
    css: style.css
editor_options: 
  chunk_output_type: console
---

# Data Pre-processing



# Importing libraries

```{r libraries, message=FALSE, warning=FALSE}

library(easystats)
library(matrixStats)
library(vegan)
library(GGally)
library(ggpubr)
library(glmnet)
library(ggbeeswarm)
library(lme4)
library(lmerTest)
library(readxl)
library(ggforce)
library(ggtext)
library(partR2)
library(tidyverse)
source('scripts/LC-MS-MS/5.Metabolite_model/functions_cdis_norm_stats.R')


```

# Loading data 

```{r}
## Raw matrix

RP_df <- read_csv("scripts/LC-MS-MS/1.Pre-processing/output/RP_matrix_clean_2025.csv")


## Loading annotation

annot <- read_csv("scripts/LC-MS-MS/3.Cleanning_Annotation/output/RP_final_annotation_checked.csv")


```

# ENVS 

```{r}

evs <- read_csv("input/env_data_ready_2025.csv") %>% 
  filter(!sample == c('SAMP1-S1-22', 'SAMP1-S2-22', 'SAMP1-S3-22', 'SAMP1-S4-22'))%>% 
  mutate(sample = str_replace_all(sample, "-", "_")) %>% 
  rename(SampleID = sample)

```

# Microbial diversity

```{r}

avs_shannon <- read_csv("input/ASV_shannon.csv") %>% 
  mutate(sample = str_replace_all(sample, "-", "_")) %>% 
  rename(SampleID = sample) %>% 
  select(!dataset)

```

# Metadata

```{r}

metadata <- read_csv("input/metadata_omics.csv") %>% 
  select(!c(SampleID_metaT, SampleID_HILIC))

```


## Prepare matrix

```{r}

RP_ready <- RP_df %>% 
  select(FeatureID, contains("RP")) %>% 
  filter(FeatureID %in% annot$FeatureID) %>% 
  column_to_rownames(var = "FeatureID")

```

# Preprare matrix - envs and microbial diversity

```{r}

envs_microbe <- evs %>% 
  mutate(C_N = total_C/total_N) %>% 
  left_join(avs_shannon, by = 'SampleID') %>% 
  left_join(metadata, by = c('SampleID' = 'treatment_code')) %>% 
  dplyr::select(!c(SampleID, sampling, stage, month, site)) %>%
  rename(SampleID = SampleID_RP) %>% 
  column_to_rownames(var = 'SampleID') %>% 
  select(total_N, ph_mean, moisture, NDVI, C_N)

```


## Standarization of envs matrix

```{r}

env_std <- decostand(envs_microbe, method = 'standardize') %>% 
  rownames_to_column(var = 'SampleID')

```


# Feature filtering

```{r}
### Calculating IQR to keep only traits that are changing more
## Filtering metabolites and genes that doesn't change much
## Idea decrease the number of features to have a more balance matrices 

iqr_test_matrix <- rowIQRs(as.matrix(RP_ready))

iqr_res <- as.data.frame(as.matrix(iqr_test_matrix))

# Filtering according to metaboAnalyts

# 1) Less than 250 variables: 5% will be filtered; 
# 2) Between 250 - 500 variables: 10% will be filtered; 
# 3) Between 500 - 1000 variables: 25% will be filtered; and 
# 4) Over 1000 variables: 40% will be filtered.

## 10% of the features will be filtered

# Sorting from largest to smallest (higher ranking = low iqr)
rk <- rank(-iqr_test_matrix, ties.method = 'random')

# Determining the number of remaining features
remain.num <- nrow(RP_ready)*.9

remain <- rk <= remain.num

features_keep <- rownames(RP_ready)[remain] 
  
## Filter matrix

matrix_ft <- RP_ready[features_keep,]

## Note: 919 features kept for downstream analysis
```


# Metabolite Normalization

## Median normalization

```{r}

norm_matrix <- median.norm(matrix_ft, transform_data = TRUE)

```

# Joining with metadata 

```{r}

temp_matrix <- as.data.frame(t(norm_matrix)) %>% 
  rownames_to_column(var = "SampleID") %>% 
  left_join(env_std, by = "SampleID") %>%
  left_join(metadata, c('SampleID' = 'SampleID_RP')) %>% 
  dplyr::select(!c(treatment_code, sampling, month, stage)) %>% 
  pivot_longer(!c(SampleID, total_N, moisture, ph_mean, C_N, NDVI, site),
               names_to = 'feature', values_to = 'intensity')

```


## Testing linear mixed-effect models

## Setting pH and MOISTURE as fixed effect and SITE as Random effect


```{r}

safe_lmer <- safely(lmerTest::lmer, otherwise = 'Model not worked')

lmer <- temp_matrix %>% 
  group_by(feature) %>%
  tidyr::nest() %>% # makes tibble of data for each feature, saves as 'lmer' column
  mutate( # change the data in 'lmer' column
    lmer_res = purrr::map(data, function(x) {
      lmer <- safe_lmer(intensity ~ total_N + moisture + ph_mean + C_N + NDVI + (1|site), 
                        data = x) # do the lmer test
      
      if(is.null(lmer$error)){
        return(lmer$result)
      } else {
        return(lmer$error)
      }
    } )
    
  )


```


## Calculating Contrast

```{r}

# Order of parameters using fixef(lmer$lmer_res[[1]])
# (Intercept)     total_N    moisture     ph_mean         C_N        NDVI 
# (Intercept) represents the fixed effect
# Thus,
# total_N  ---  L = c(0, 1, 0, 0, 0, 0)
# moisture ---  L = c(0, 0, 1, 0, 0, 0)
# pH_mean  ---  L = c(0, 0, 0, 1, 0, 0)
# C_N      ---  L = c(0, 0, 0, 0, 1, 0) 
# NDVI     ---  L = c(0, 0, 0, 0, 0, 1)

```


## total N

```{r}
df_contrasts_TN <- lmer %>%
  group_by(feature) %>%
  group_modify(~ {
    contest(.x$lmer_res[[1]],
            L = c(0, 1, 0, 0, 0, 0),
            joint = FALSE)
  }) %>%
  magrittr::set_colnames(c('feature', 'estimate', 'stderror', 
                           'df', 'tvalue', 'lower', 'upper', 'pvalue')) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(pvalue, method = 'fdr')) 

sig_feature_TN <- df_contrasts_TN %>% 
  filter(padj < 0.05) %>% 
  rename(FeatureID = feature)



```


### moisture

```{r}


df_contrasts_moisture <- lmer %>%
  group_by(feature) %>%
  group_modify(~ {
    contest(.x$lmer_res[[1]],
            L = c(0, 0, 1, 0, 0, 0),
            joint = FALSE)
  }) %>%
  magrittr::set_colnames(c('feature', 'estimate', 'stderror', 
                           'df', 'tvalue', 'lower', 'upper', 'pvalue')) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(pvalue, method = 'fdr')) 

sig_feature_moisture <- df_contrasts_moisture %>% 
  filter(padj < 0.05) %>% 
  rename(FeatureID = feature)


```


## pH

```{r}

df_contrasts_pH <- lmer %>%
  group_by(feature) %>%
  group_modify(~ {
    contest(.x$lmer_res[[1]],
            L = c(0, 0, 0, 1, 0, 0),
            joint = FALSE)
  }) %>%
  magrittr::set_colnames(c('feature', 'estimate', 'stderror', 
                           'df', 'tvalue', 'lower', 'upper', 'pvalue')) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(pvalue, method = 'fdr')) 

sig_feature_pH <- df_contrasts_pH %>% 
  filter(padj < 0.05) %>% 
  rename(FeatureID = feature)




```

## CN

```{r}

df_contrasts_cn <- lmer %>%
  group_by(feature) %>%
  group_modify(~ {
    contest(.x$lmer_res[[1]],
            L = c(0, 0, 0, 0, 1, 0),
            joint = FALSE)
  }) %>%
  magrittr::set_colnames(c('feature', 'estimate', 'stderror', 
                           'df', 'tvalue', 'lower', 'upper', 'pvalue')) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(pvalue, method = 'fdr')) 

sig_feature_cn <- df_contrasts_cn %>% 
  filter(padj < 0.05) %>% 
  rename(FeatureID = feature)



```


## NDVI

```{r}

df_contrasts_ndvi <- lmer %>%
  group_by(feature) %>%
  group_modify(~ {
    contest(.x$lmer_res[[1]],
            L = c(0, 0, 0, 0, 0, 1),
            joint = FALSE)
  }) %>%
  magrittr::set_colnames(c('feature', 'estimate', 'stderror', 
                           'df', 'tvalue', 'lower', 'upper', 'pvalue')) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(pvalue, method = 'fdr')) 

sig_feature_ndvi <- df_contrasts_ndvi %>% 
  filter(padj < 0.05) %>% 
  rename(FeatureID = feature)



```



#################### Testing foward selected model #############################


```{r}

safe_lmer <- safely(lmerTest::lmer, otherwise = 'Model not worked')

lmer_2 <- temp_matrix %>% 
  group_by(feature) %>%
  tidyr::nest() %>% # makes tibble of data for each feature, saves as 'lmer' column
  mutate( # change the data in 'lmer' column
    lmer_res = purrr::map(data, function(x) {
      lmer <- safe_lmer(intensity ~ moisture + ph_mean + C_N + (1|site), 
                        data = x) # do the lmer test
      
      if(is.null(lmer$error)){
        return(lmer$result)
      } else {
        return(lmer$error)
      }
    } )
    
  )



```


## Check model performance - lmer

```{r}

performance_list <- map2(lmer_2$feature, lmer_2$lmer_res, function(featureid, model){
  
  df_performance <- model_performance(model) %>% 
    mutate(featureID = featureid)
  
  if(ncol(df_performance) == 8){
    df_performance <- df_performance %>% 
      mutate(ICC = NA)
  }
  
  return(df_performance)
  
})

performance_df <- reduce(performance_list, rbind)

write_csv(performance_df, "scripts/LC-MS-MS/5.Metabolite_model/output/RP-Performance_model_results_forward.csv")


```


## Calculating Contrast

```{r}

# Order of parameters using fixef(lmer$lmer_res[[1]])
# (Intercept)    moisture     ph_mean         C_N  
# (Intercept) represents the fixed effect
# Thus,
# moisture  ---  L = c(0, 1, 0, 0)
# pH_mean ---    L = c(0, 0, 1, 0)
# C_N  ---       L = c(0, 0, 0, 1)


```


### moisture

```{r}


df_contrasts_moisture <- lmer_2 %>%
  group_by(feature) %>%
  group_modify(~ {
    contest(.x$lmer_res[[1]],
            L = c(0, 1, 0, 0),
            joint = FALSE)
  }) %>%
  magrittr::set_colnames(c('feature', 'estimate', 'stderror', 
                           'df', 'tvalue', 'lower', 'upper', 'pvalue')) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(pvalue, method = 'fdr')) 

sig_feature_moisture <- df_contrasts_moisture %>% 
  filter(padj < 0.05) %>% 
  rename(FeatureID = feature)


```


## pH

```{r}

df_contrasts_pH <- lmer_2 %>%
  group_by(feature) %>%
  group_modify(~ {
    contest(.x$lmer_res[[1]],
            L = c(0, 0, 1, 0),
            joint = FALSE)
  }) %>%
  magrittr::set_colnames(c('feature', 'estimate', 'stderror', 
                           'df', 'tvalue', 'lower', 'upper', 'pvalue')) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(pvalue, method = 'fdr')) 

sig_feature_pH <- df_contrasts_pH %>% 
  filter(padj < 0.05) %>% 
  rename(FeatureID = feature)



```

## CN

```{r}

df_contrasts_cn <- lmer_2 %>%
  group_by(feature) %>%
  group_modify(~ {
    contest(.x$lmer_res[[1]],
            L = c(0, 0, 0, 1),
            joint = FALSE)
  }) %>%
  magrittr::set_colnames(c('feature', 'estimate', 'stderror', 
                           'df', 'tvalue', 'lower', 'upper', 'pvalue')) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(pvalue, method = 'fdr')) 

sig_feature_cn <- df_contrasts_cn %>% 
  filter(padj < 0.05) %>% 
  rename(FeatureID = feature)



```


## Saving significant features

```{r}

moisture_ready <- sig_feature_moisture %>% 
  left_join(annot, by = 'FeatureID') %>% 
  mutate(env_sig = 'moisture')


ph_ready <- sig_feature_pH %>% 
  left_join(annot, by = 'FeatureID') %>% 
  mutate(env_sig = 'pH')


CN_ready <- sig_feature_cn %>% 
  left_join(annot, by = 'FeatureID') %>% 
  mutate(env_sig = 'CN')


full_table <- rbind(moisture_ready, ph_ready)


write_csv(full_table, "scripts/LC-MS-MS/5.Metabolite_model/output/RP_sig_features_model_fwd_selected_RP.csv")

```

