---
title: "Figure 2 - Cazyme metaT"
author: 'Viviana Freire Zapata'
output: html_notebook
editor_options: 
  chunk_output_type: console
---


```{r}
library(patchwork)
library(ggh4x)
library(readxl)
library(ggpubr)
library(vegan)
library(ggtext)
library(ggforce)
library(reshape2)
library(rstatix)
library(ggsankey)
library(tidyverse)

```

# Loading data

## Metadata

```{r}
metadata <- read_csv('input/metadata_omics.csv') 

metadata_ready <- metadata %>% 
  mutate(Month = factor(month, levels = c("May", "July_1", "July_2", "July_3",
                                          "August", "October", "May_22")))
```

## Taxonomy

```{r}
mag_taxonomy <- read_csv('input/mag_taxonomy_denovo.csv') %>%
  mutate(Class = ifelse(is.na(Class), 
                        glue::glue('Unassigned Class \nof {Phylum}'),
                        Class))

```


## CAZyme results - updated dbcan v3

```{r}

dbcan_sub <- read_tsv("input/CAZyme_dbcan3/dbcan_output/dbcan-sub.hmm.out") %>% 
  select(Geneid = `Gene ID`, dbcan_sub = `dbCAN subfam`, 
         Substrate)

```


## geTMM

```{r}
geTMM <- read_csv("output/geTMM_normalized_matrix_2025.csv")

geTMM_long <- geTMM %>% 
  pivot_longer(!Geneid, names_to = "sampleid", values_to = "geTMM")

```

## DE expression

## Loading DE genes
```{r}

de_res <- read_csv("output/DE_result_month_slice_10_2025.csv") %>% 
  filter(padj <= 0.01) %>% 
  filter(abs(log2FoldChange) > 2) ### REMEMBER THIS FILTERS FOR SELECTING DE GENES

```


## Loading HCLUST results

```{r}

hclust <- read_csv('output/Hclust_MAG_expression_2025.csv')


plot_groups <- hclust %>% 
  rename(cluster = order) %>% 
  mutate(cluster = paste0('Cluster ', cluster),
         response_type = case_when(str_detect(cluster, 'Cluster 1') ~ 'Dry',
                                   str_detect(cluster, 'Cluster 2') ~ 'Wet',
                                   .default = 'Pulse')) %>% 
  select(genome, cluster, response_type) %>% 
  distinct()

```


## Expressed genes

```{r}
expressed_cazy <- dbcan_sub %>% 
  inner_join(geTMM_long, by = 'Geneid') %>% 
  mutate(cazy_fam = case_when(str_detect(dbcan_sub, 'GH') ~ 'Glycoside Hydrolases',
                              str_detect(dbcan_sub, 'GT') ~ 'Glycosyl Transferases',
                              str_detect(dbcan_sub, 'PL') ~ 'Polysaccharide Lyases',
                              str_detect(dbcan_sub, 'CE') ~ 'Carbohydrate Esterases',
                              str_detect(dbcan_sub, 'AA') ~ 'Auxiliary Activities',
                              str_detect(dbcan_sub, 'CBM') ~ 'Carbohydrate Binding Module'),
         cazy_sub = str_remove(dbcan_sub, '_e.*'))

## Number of expressed subfamilies families
ex_sub_family <- expressed_cazy %>% 
  filter(geTMM > 0) %>% 
  select(Geneid, cazy_fam, cazy_sub) %>% 
  distinct() %>% 
  group_by(cazy_fam, cazy_sub) %>% 
  dplyr::count(name = 'num_subfamilies')


ex_family <- expressed_cazy %>% 
  filter(geTMM > 0) %>% 
  select(Geneid, cazy_fam) %>% 
  distinct() %>% 
  group_by(cazy_fam) %>% 
  dplyr::count(name = 'num_subfamilies')

```


## Supplementary table 5

```{r}

de_cazy_genes <- de_res %>% 
  filter(FeatureID %in% expressed_cazy$Geneid) %>% 
  inner_join(expressed_cazy, by = c('FeatureID' = 'Geneid')) %>%
  select(!c(sampleid, geTMM)) %>% 
  distinct()

write_csv(de_cazy_genes, "output_tables/Supplementary_data_2.csv")


```


## Expression by substrate

### Number of Cazymes per substrate

```{r}
cazy_per_sub <- expressed_cazy %>% 
  filter(geTMM > 0) %>% 
  select(Geneid, Substrate, cazy_fam) %>% 
  distinct() %>%
  mutate(Substrate = ifelse(Substrate == '-', 'No substrate', Substrate)) %>% 
  separate_longer_delim(Substrate, delim = ', ') %>% 
  distinct() %>% 
  group_by(Substrate, cazy_fam) %>% 
  dplyr::count()

cazy_per_sub_plot <- cazy_per_sub %>% 
  filter(n >= 30) %>% 
  group_by(Substrate) %>% 
  mutate(total = sum(n)) %>% 
  ggplot(aes(x = n,
             y = fct_reorder(Substrate, total),
             fill = cazy_fam))+
  geom_col(color = 'black')+
  labs(x = 'Number of genes expressed',
       y = 'Substrate',
       fill = 'CAZy family',
       title = 'Metatranscriptomics')+
  guides(fill = guide_legend(ncol = 1))+
  scale_fill_brewer(palette = 'Set2') +
  theme_bw()+
  theme(panel.grid = element_blank(),
        legend.position = c(0.7, 0.5),
        legend.text = element_text(size = 5),
        legend.title = element_text(size = 6),
        legend.box.spacing = unit(.05, 'lines'),
        legend.key.size = unit(.5, 'lines'),
        legend.background = element_blank(),
        legend.title.position = 'top',
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7),
        axis.text.y = element_text(size = 6),
        plot.title = element_text(face = 'bold', hjust = 0.5, size = 8),
        plot.margin = margin(0, 0, 0, 0))

cazy_per_sub_plot


```

## Counting the number of Cazymes families per MAGs - DE expressed

```{r}
cazy_per_class <- expressed_cazy %>% 
  filter(Geneid %in% de_res$FeatureID) %>% 
  filter(geTMM > 1) %>% 
  select(Geneid, Substrate) %>% 
  distinct() %>%
  mutate(bin = str_remove(Geneid, '_k.*')) %>% 
  inner_join(mag_taxonomy, by = 'bin') %>% 
  mutate(Substrate = ifelse(Substrate == '-', 'No substrate', Substrate)) %>% 
  separate_longer_delim(Substrate, delim = ', ') %>% 
  distinct() %>% 
  group_by(Substrate, Class, Phylum) %>% 
  dplyr::count()

cazy_class_bubble <- cazy_per_class %>%
  filter(Substrate != 'No substrate') %>% 
  rename('DE CAZyme' = 'n') %>% 
  filter(`DE CAZyme` > 1) %>% 
  ggplot() +
  geom_point(aes(x = Substrate,
                 y = Class,
                 size = `DE CAZyme`,
                 fill = Phylum),
             shape = 21,
             alpha = .5,
             color = 'black') + 
  # facet_grid(cols = vars(Phylum),
  #            scales = 'free',
  #            space = 'free') +
  guides(fill = guide_legend(ncol = 1),
         size = guide_legend(ncol = 2))+
  #labs(title = 'Metatranscriptomics')+
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, size = 6, hjust = 1),
        axis.text.y.left = element_text(size = 6),
        axis.title.y = element_text(size = 6),
        axis.title.x = element_blank(),
        panel.grid = element_blank(),
        strip.background = element_rect(fill = 'white'),
        strip.text.x.top = element_text(size = 7, angle = 90),
        legend.position = 'right',
        legend.text = element_text(size = 5),
        legend.title = element_text(size = 6),
        legend.box.spacing = unit(.05, 'lines'),
        legend.key.size = unit(.5, 'lines'),
        legend.box = "vertical",
        legend.spacing.y = unit(0.05, "lines"),
        legend.margin = margin(t = 0, b = 0),
        legend.title.position = 'top',
        plot.title = element_text(face = 'bold', hjust = 0.5, size = 8),
        plot.margin = margin(0, 0, 0, 0))

cazy_class_bubble


```


## Plotting expressed cazymes and substrates - Heatmap


## Heatmap of expression

```{r}
metaT_cazy_heat <- expressed_cazy %>% 
  filter(Geneid %in% de_res$FeatureID) %>% 
  select(Geneid, sampleid, dbcan_sub, geTMM) %>% 
  distinct() %>% 
  mutate(bin = str_remove(Geneid, '_k.*')) %>% 
  inner_join(mag_taxonomy, by = 'bin') %>% 
  left_join(plot_groups, by = c('bin' = 'genome')) %>% 
  group_by(Phylum, Class, sampleid, dbcan_sub, response_type) %>% 
  summarise(total_expression = sum(geTMM)) %>%
  mutate(sampleid = str_replace_all(sampleid, '-', '_')) %>% 
  inner_join(metadata_ready, by = c('sampleid' = 'treatment_code')) %>% 
  group_by(Phylum, Class, dbcan_sub, Month, response_type) %>% 
  summarise(mean_expr = mean(total_expression))

```


```{r}

substrate_subfamily <- expressed_cazy %>% 
  filter(Geneid %in% de_res$FeatureID) %>% 
  select(dbcan_sub, Substrate) %>% 
  mutate(Substrate = ifelse(Substrate == '-', 'No substrate', Substrate)) %>% 
  separate_longer_delim(Substrate, delim = ', ') %>% 
  distinct()

## Filter Class that express only 1 class

filter_class <- c('Blastocatellia', 'Chloroflexia', 'Dehalococcoidia', 
                  'Gemmatimonadetes', 'Methylomirabilia', 'Binatia')

cazy_heat <- metaT_cazy_heat %>%
  inner_join(substrate_subfamily, by = 'dbcan_sub') %>%
  group_by(dbcan_sub, Substrate) %>% 
  mutate(to_order = mean(mean_expr)) %>% 
  arrange(to_order) %>% 
  mutate(order = n():1) %>% 
  filter(!Substrate == 'No substrate',
         !(Class %in% filter_class)) %>% 
  ggplot(aes(x = Month,
             y = fct_reorder(dbcan_sub, order),
             fill = log(mean_expr + 1)))+
  geom_tile()+
  scale_fill_viridis_c(option = 'A', direction = -1)+
  facet_grid2(rows = vars(Substrate),
              cols = vars(Class),
              scales = 'free',
              space = 'free',
              strip = strip_nested(clip = 'off'),
              switch = 'y')+
  labs(y = 'CAZyme subfamily',
       fill = 'Mean expression per month')+
  theme_bw()+
  theme(axis.title.x = element_blank(),
        legend.position = 'bottom',
        axis.ticks.y = element_blank(),
        strip.text.y.left = element_text(angle = 0, hjust = 1, size = 5),
        strip.text.x.top = element_text(hjust = 1, angle = 90, size = 5),
        strip.background = element_blank(),
        axis.text.x = element_text(angle = 90, size = 6),
        axis.text.y = element_text(size = 5),
        panel.grid = element_blank(),
        strip.placement = 'outer',
        legend.text = element_text(size = 6),
        legend.key.size = unit(.5, 'lines'),
        legend.box.spacing = unit(.01, 'lines'),
        legend.title = element_text(size = 6))

cazy_heat

ggsave("output_figures/Sup_figure_4.png", cazy_heat, dpi = 300,
       units = 'mm', 
       height = 225, 
       width = 180)

```

# Per substrate

```{r}
metaT_cazy_heat_subs <- expressed_cazy %>% 
  filter(Geneid %in% de_res$FeatureID) %>%
  select(Geneid, sampleid, Substrate, geTMM) %>% 
  mutate(Substrate = ifelse(Substrate == '-', 'No substrate', Substrate)) %>% 
  separate_longer_delim(Substrate, delim = ', ') %>% 
  distinct() %>% 
  mutate(bin = str_remove(Geneid, '_k.*')) %>% 
  inner_join(mag_taxonomy, by = 'bin') %>% 
  left_join(plot_groups, by = c('bin' = 'genome')) %>% 
  group_by(Phylum, Class, sampleid, Substrate, response_type) %>% 
  summarise(total_expression = sum(geTMM)) %>%
  mutate(sampleid = str_replace_all(sampleid, '-', '_')) %>% 
  inner_join(metadata_ready, by = c('sampleid' = 'treatment_code')) %>%
  group_by(Phylum, Class, Substrate, Month, response_type) %>% 
  summarise(mean_expr = mean(total_expression)) 


cazy_heat_subs <- metaT_cazy_heat_subs %>% 
  filter(Substrate != 'No substrate') %>% 
  ggplot(aes(x = Month,
             y = Substrate,
             fill = log(mean_expr + 1)))+
  geom_tile()+
  scale_fill_viridis_c(option = 'A', direction = -1) +
  facet_grid2(cols = vars(response_type, Class),
              scales = 'free',
              space = 'free',
              strip = strip_nested(clip = 'off')) +
  labs(y = 'Substrate',
       fill = 'Mean expression per month')+
  theme_bw()+
  theme(axis.title.x = element_blank(),
        legend.position = 'bottom',
        # strip.text.y.left = element_text(angle = 0, size = 6),
        # strip.text.y.right = element_text(angle = 0, size = 6),
        strip.text.x.top = element_text(angle = 90, size = 6),
        strip.background.x = element_blank(),
        axis.text.x = element_text(angle = 90, size = 6),
        axis.text.y = element_text(size = 6),
        axis.title.y = element_text(size = 7),
        panel.grid = element_blank(),
        strip.placement = 'outer',
        legend.text = element_text(size = 5),
        legend.title = element_text(size = 6),
        legend.box.spacing = unit(.05, 'lines'),
        legend.key.size = unit(.5, 'lines'))

cazy_heat_subs


```


## Filtering cazyme most expressed


```{r}
## Filter only highly expressed modules

top_modules <- metaT_cazy_heat_subs %>% 
  filter(!Substrate == 'No substrate') %>% 
  group_by(Substrate) %>% 
  summarise(total_exp = sum(mean_expr)) %>% 
  slice_max(order_by = total_exp, n = 10)

```


## Sankey

```{r}

taxonomy_order <- mag_taxonomy %>% 
  select(Domain, Phylum, Class) %>%
  distinct() %>% 
  arrange(Domain, Phylum, Class) %>% 
  pivot_longer(everything(), names_to = 'name', values_to = 'node') %>% 
  filter(!str_detect(node, '^ ')) %>% 
  distinct() %>% 
  mutate(order = n():1)

month_order <- tibble(node = c('May', 
                               'July_1',
                               'July_2',
                               'July_3',
                               'August',
                               'October',
                               'May_22')) %>% 
  mutate(name = 'Month',
         order = 1:7)

substrate_order <- tibble(node = c("beta-mannan",
                                   "cellulose",
                                   "chitin",
                                   "exo-polysaccharide",
                                   "glucooligosaccharide",
                                   "glycogen",
                                   "peptidoglycan",
                                   "starch",
                                   "sucrose",
                                   "xylan")) %>% 
    mutate(name = 'Substrate',
         order = 10:1)

## Create Sankey ready df

df_long <- metaT_cazy_heat_subs %>% 
  filter(Substrate %in% top_modules$Substrate) %>%
  #filter(Substrate != 'No substrate') %>% 
  mutate(month_response = glue::glue('{Month}-{response_type}')) %>% 
  split(.$month_response) %>% 
  imap(function(df, m){
    df %>% 
      ungroup() %>% 
      make_long(Phylum, Class, Substrate, value = mean_expr)%>% 
      left_join(rbind(taxonomy_order, month_order, substrate_order), by = c('node', 
                                                                            'x' = 'name')) %>% 
      mutate(x = factor(x, levels = c('Phylum', 'Class', 'Substrate')),
             order = ifelse(is.na(order), 1, order),
             month_response = m)
  }) %>% purrr::reduce(rbind) %>% 
  separate_wider_delim(month_response, names = c('Month', 'response_type'), delim = '-') %>% 
  mutate(Month = factor(Month, levels = c('May', 'July_1', 'July_2', 'July_3',
                                          'August', 'October', 'May_22'))) %>% 
  filter(Month %in% c('May', 'July_1', 'October'),
         response_type != 'Pulse')


sankey_plot <- df_long %>% 
  mutate(response_type = glue::glue('{response_type}\nadapted')) %>% 
  ggplot(aes(x = x,
             next_x = next_x,
             node = fct_reorder(node, dplyr::desc(order)),
             next_node = next_node,
             fill = factor(node),
             label = node,
             value = value)) +
  geom_sankey(show.legend = FALSE,
              flow.alpha = 0.5,
              node.color = 'black') +
  # geom_sankey_text(size = 2,
  #                  show.legend = FALSE,
  #                  position = position_nudge(x = .1),
  #                  hjust = 0) +
  geom_sankey_label(size = 1.8, color = 1, fill = "white") +
  labs(title = 'Carbohydrate-active enzymes - Metatranscriptomics')+
  facet_grid(cols = vars(Month),
             rows = vars(response_type),
             switch = 'y',
             space = 'fixed',
             scales = 'free_y') +
  theme_sankey() +
  theme(axis.title = element_blank(),
        plot.title = element_text(face = 'bold', hjust = 0.5, size = 8),
        plot.subtitle = element_text(hjust = 0.5),
        strip.placement = 'outer',
        strip.text = element_text(angle = 0, size = 6, face = 'bold'),
        strip.text.y.left = element_text(angle = 90, size = 6, face = 'bold'),
        panel.spacing.y = unit(0, 'lines'),
        plot.margin = margin(0, 0, 0, 0)) +
  coord_cartesian(clip = 'off') +
  scale_fill_manual(values = ggpubr::get_palette('Paired', 27))


sankey_plot

```


## Saving plot

```{r}

## Final figure
des <- "
AB
CC
"


final_fig <- free(plot_count) + free(plot_count_2)+ free(sankey_plot) +
  plot_layout(heights = c(2, 4),
              design = des) +
  plot_annotation(tag_levels = list(c('A', 'B', 'C'))) &
  theme(legend.key.spacing = unit(.1, 'lines'))

final_fig

final_fig <- final_fig +
  plot_annotation(title = "Metatranscriptomics")&
  theme(plot.title = element_text(face = 'bold', hjust = 0.5))

final_fig

```



## Save final plot


```{r}


ggsave("output_figures/Figure_2.png", final_fig, dpi = 300,
       units = 'mm', 
       height = 220, 
       width = 180)


```


## Supplementary

```{r}

sup_fig <- free(cazy_per_sub_plot)+ free(cazy_class_bubble) +
  plot_annotation(tag_levels = list(c('A', 'B'))) &
  theme(axis.title.y = element_blank(),
        legend.key.spacing = unit(.1, 'lines'))

sup_fig



ggsave("output_figures/Supp_Fig_3.png", sup_fig, dpi = 300,
       units = 'mm', 
       width = 180)

```

