---
title: "Omics integration figure"
author: "Viviana Freire-Zapata"
date: "10/24/2025"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: '2'
    collapsed: false
    highlight: tango
    css: style.css
editor_options: 
  chunk_output_type: console
---

# Data Pre-processing

## 1. Setup

### Importing libraries

```{r libraries, message=FALSE, warning=FALSE}

library(matrixStats)
library(dtwclust)
library(ggtext)
library(ggh4x)
library(ggvenn)
library(patchwork)
library(mixOmics)
library(tidyverse)

source('scripts/LC-MS-MS/5.Metabolite_model/functions_cdis_norm_stats.R')

```


# Loading data

## DIABLO results

```{r}

metabolites <- read_csv("output/DIABLO_metabolomics_features_11-10-2025.csv")

metabolites_c1 <- metabolites %>% 
  filter(component == 'comp1')


microbe <- read_csv("output/DIABLO_metaG_features_11-10-2025.csv")

microbe_c1 <- microbe %>% 
  filter(component == 'comp1')


```

# LME results

```{r}

RP_lme <- read_csv("scripts/LC-MS-MS/5.Metabolite_model/output/RP_sig_features_model_fwd_selected_RP.csv")

```

# Comparing LME vs DIABLO

```{r}

metabo_lme <- RP_lme %>% 
  dplyr::select(FeatureID) %>% 
  distinct()

metabo_diablo <- metabolites_c1 %>% 
  dplyr::select(FeatureID) %>% 
  distinct()

data_list <- c(DIABLO = metabo_diablo,
               LMEM = metabo_lme)


```

## Venn diagram

```{r}

venn_lme_diablo <- ggvenn(data_list,
                          fill_color = c("#4C72B0", "#55A868"),  
                          stroke_size = 0.6,
                          set_name_size = 2.5,
                          text_size = 3,
                          show_percentage = FALSE,
                          auto_scale = TRUE) +
  
  theme_void() +
  coord_fixed(clip = 'off', ratio = 0.5) +
  theme(
    legend.position = "none",
    plot.margin = margin(0, 0, 0, 0, "lines"))

venn_lme_diablo


```


## MetaG - metaT

# DIABLO loadings

```{r}
# function for repeated elements
rep_elements <- function(type, n, ...){
  
  res_final <- list()
  
  dots <- list(...)
  
  for(i in 1:n){
    
    if(type == 'blank'){
      res_final[[i]] <- element_blank()
    } else if(type == 'text'){
      res_final[[i]] <- element_text(color = dots$color,
                                     size = dots$size,
                                     face = dots$face,
                                     angle = dots$angle)
    } else if(type == 'rect'){
      res_final[[i]] <- element_rect(fill = dots$fill,
                                     color = dots$color)
    }
    
  }
  
  return(res_final)
}
```


```{r}

## Component 1

comp_1_mag <- microbe %>% 
  filter(component == 'comp1') %>% 
  ggplot(aes(x = importance,
             y = fct_reorder(KO, dplyr::desc(abs(importance))),
             fill = GroupContrib)) +
  geom_col(color = 'black', linewidth = 0.01) +
  facet_grid2(rows = vars(Class, MAG),
              switch = 'y',
              space = 'free',
              scales = 'free',
              strip = strip_nested(
                text_y = c(rep_elements('text', 15, angle = 0, size = 5),
                           rep_elements('text', 64, angle = 90, size = 0.001,
                                        color = 'transparent')),
                background_y = element_blank(),
                clip = 'off',
                size = 'variable'
              )
  ) +
  scale_fill_manual(values = c("#E69F00", "#7BBBE5"))+
  labs(x = "Loading Importance",
       y = "MAG_KO",
       fill = "Group Contribution") +
  theme_void() +
  theme(
    axis.text.x = element_text(size = 5),
    axis.title.x = element_text(size = 6),
    legend.position = 'none',
    strip.placement = 'outer',
    legend.title = element_text(size = 6), 
    legend.text = element_text(size = 5), 
    legend.key.size = unit(.3, 'lines'), 
    legend.spacing.x = unit(0.1, "cm"),
    panel.spacing = unit(0.1, "lines"), 
  )

comp_1_mag
```

## expression 

```{r}

kos_diablo_1 <- microbe %>% 
  filter(component == 'comp1') %>% 
  dplyr::select(MAG, KO, importance)

```

## Metadata

```{r}
metadata <- read_csv('input/metadata_omics.csv') 

metadata_ready <- metadata %>% 
  mutate(Month = factor(month, levels = c("May", "July_1", "July_2", "July_3",
                                          "August", "October", "May_22")))
```

## Taxonomy

```{r}

mag_taxonomy <- read_csv('input/mag_taxonomy_denovo.csv') %>% 
  mutate(Class = ifelse(is.na(Class), 
                        glue::glue('Unassigned Class \nof {Phylum}'),
                        Class))

```


## MAGs annotations

```{r}

annot_cydbs <- read_csv("input/annot_with_cydbs_2.23.25.csv") %>% 
  rename(Geneid = `...1`)

```


## geTMM

```{r}
geTMM <- read_csv("output/geTMM_normalized_matrix_2025.csv")

geTMM_long <- geTMM %>% 
  pivot_longer(!Geneid, names_to = "sampleid", values_to = "geTMM")

```

## Joining expression to annotations

```{r}

annot_vip <- geTMM_long %>% 
  left_join(annot_cydbs, by = 'Geneid') %>%
  group_by(sampleid, fasta, ko_id) %>%
  filter(fasta %in% kos_diablo_1$MAG,
         ko_id %in% kos_diablo_1$KO) %>% 
  summarise(getmm_sum = sum(geTMM)) %>% 
  mutate(sampleid = str_replace_all(sampleid, '-', '_')) %>% 
  left_join(metadata_ready, by = c('sampleid' = 'treatment_code')) %>% 
  group_by(fasta, ko_id, Month) %>% 
  summarise(getmm_mean = mean(getmm_sum)) %>% 
  filter(getmm_mean > 0) %>% 
  inner_join(kos_diablo_1, by = c('fasta' = 'MAG', 'ko_id' = 'KO'))

```

## Heatmap KOs DIABLO
```{r}

htmp_diablo <- annot_vip %>% 
  group_by(fasta, ko_id) %>% 
  mutate(zscore = (getmm_mean - mean(getmm_mean))/ sd(getmm_mean)) %>% 
  ungroup() %>% 
  left_join(mag_taxonomy, by = c('fasta' = 'bin')) %>% 
  ggplot() + 
  geom_tile(aes(x = Month, 
                y = fct_reorder(ko_id, dplyr::desc(abs(importance))), 
                fill = zscore)) + 
  # scale_fill_viridis_c(option = 'A', direction = -1) + 
  scale_fill_distiller(palette = 'RdBu', direction = -1) + 
  facet_grid2(rows = vars(Class, fasta),
              scales = 'free', 
              space = 'free',
              strip = strip_nested(
                text_y = element_blank(),
                background_y = element_blank()
              )) + 
  theme(panel.grid = element_blank(), 
        axis.title = element_blank(), # closer to panels 
        panel.spacing = unit(0.1, "lines"), 
        axis.text.y = element_text(size = 4), 
        axis.text.x = element_text(size = 5, 
                                   angle = 45, hjust = 1), 
        legend.title = element_text(size = 6), 
        legend.text = element_text(size = 5), 
        legend.position = 'bottom', 
        legend.key.size = unit(.3, 'lines'), 
        legend.spacing.x = unit(0.1, "cm"), 
        plot.margin = margin(0.1, 0.1, 0.1, 0.1, 'lines'))

htmp_diablo

htmp_loadings <- comp_1_mag + htmp_diablo + 
  plot_layout(guides = 'collect') &
  theme(legend.position = 'bottom')

htmp_loadings

```


## DIABLO PCA

```{r}
diablo_moisture <- readRDS("output/diablo_moisture_final_11_10_25.rds")


set.seed(123) # For reproducibility


perf.diablo_final = perf(diablo_moisture, validation = 'Mfold',
                         folds = 5, nrepeat = 100, dist = "mahalanobis.dist")

plot(perf.diablo_final)

# Performance with Majority vote
perf.diablo_final$MajorityVote.error.rate

# Performance with Weighted vote
perf.diablo_final$WeightedVote.error.rate

```


```{r}
plotIndiv(diablo_moisture,
          block = 'metaT',
          comp = c(1,2))

diablo_ordi <- read_csv("output/Diablo_scores_11_10_25.csv")

axis1 <- "Variate 1: metaT (69%); metaB (14%)"
axis2 <- "Variate 2: metaT (11%); metaB (14%)"

plot_ordi <- diablo_ordi %>%
  left_join(metadata_ready, by = c("names" = "treatment_code")) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(aes(color = group, shape = Month), size = 2, alpha = 0.9) +
  labs(
    x = axis1,
    y = axis2,
    color = "Group",
    shape = "Month") +
  scale_color_manual(values = c("#E69F00", "#7BBBE5"))+
  theme_bw() +
  theme(
    panel.grid.major = element_line(color = "grey85", size = 0.3),
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 6, face = "bold"),
    axis.text = element_text(size = 5, color = "black"),
    legend.position = "right",
    legend.background = element_blank(),
    legend.key = element_blank(),
    legend.text = element_text(size = 5),
    legend.title = element_text(size = 6, face = "bold"),
    legend.key.size = unit(0.15, 'lines'),
    legend.box = 'vertical',
    legend.box.spacing = unit(0.1, 'lines')) +
  scale_shape_manual(values = c(16, 17, 15, 18, 8, 3)) 

plot_ordi

```

```{r}
# loadings_df_comp1 <- plotLoadings(diablo_moisture, contrib = 'max', comp = 1) %>% 
#   reduce(rbind) %>% 
#   rownames_to_column(var = 'FeatureID') %>% 
#   mutate(component = 'comp1') %>% 
#   filter(!str_detect(FeatureID, 'FT'))
```


## Metabolomics - DTW

```{r}

metabolites_both <- metabolites %>% 
  mutate(lme_diablo = if_else(FeatureID %in% RP_lme$FeatureID,
                              'both', 'no')) %>% 
  filter(lme_diablo == 'both') %>% 
  dplyr::select(FeatureID) %>% 
  distinct()


```


# LC-MS - RP

# Loading data 

```{r}
## Raw matrix

RP_df <- read_csv("scripts/LC-MS-MS/1.Pre-processing/output/RP_matrix_clean_2025.csv")


## Loading annotation

annot <- read_csv("scripts/LC-MS-MS/3.Cleanning_Annotation/output/RP_final_annotation_checked.csv")

```


## Prepare matrix

```{r}

rp_ready <- RP_df %>% 
  dplyr::select(FeatureID, contains("RP")) %>% 
  filter(FeatureID %in% annot$FeatureID) %>% 
  column_to_rownames(var = "FeatureID")

```

# Feature filtering

```{r}
### Calculating IQR to keep only traits that are changing more
## Filtering metabolites and genes that doesn't change much
## Idea decrease the number of features to have a more balance matrices 

iqr_test_matrix <- rowIQRs(as.matrix(rp_ready))

iqr_res <- as.data.frame(as.matrix(iqr_test_matrix))

# Filtering according to metaboAnalyts

# 1) Less than 250 variables: 5% will be filtered; 
# 2) Between 250 - 500 variables: 10% will be filtered; 
# 3) Between 500 - 1000 variables: 25% will be filtered; and 
# 4) Over 1000 variables: 40% will be filtered.

## 10% of the features will be filtered

# Sorting from largest to smallest (higher ranking = low iqr)
rk <- rank(-iqr_test_matrix, ties.method = 'random')

# Determining the number of remaining features
remain.num <- nrow(rp_ready)*.9

remain <- rk <= remain.num

features_keep <- rownames(rp_ready)[remain] 

## Filter matrix

matrix_ft <- rp_ready[features_keep,]

## Note: 919 features kept for downstream analysis
```


# Metabolite Normalization

## Median normalization

```{r}

norm_matrix <- median.norm(matrix_ft, transform_data = TRUE)


```


## Filtering full matrix 

```{r}

norm_matrix_sig <- norm_matrix %>% 
  rownames_to_column(var = 'FeatureID') %>% 
  filter(FeatureID %in% metabolites_both$FeatureID) %>% 
  pivot_longer(!FeatureID, names_to = 'SampleID', values_to = 'norm_int')

```

## Adding class info


```{r}

name_class_rp <- read_csv("scripts/LC-MS-MS/4.Get_inchis/output/RP_classyfire_from_name_ready.csv") %>% 
  filter(FeatureID %in% norm_matrix_sig$FeatureID) %>% 
  dplyr::select(FeatureID, Name, superclass, class, subclass, level_5 = `level 5`) %>% 
  mutate(LC_mode = 'RP')

canopus_class_rp <- read_csv("scripts/LC-MS-MS/4.Get_inchis/output/RP_classyfire_canopus_ready.csv") %>% 
  filter(FeatureID %in% norm_matrix_sig$FeatureID) %>% 
  dplyr::select(FeatureID, Name,
                superclass = `ClassyFire#superclass`,
                class = `ClassyFire#class`, 
                subclass = `ClassyFire#subclass`, 
                level_5 = `ClassyFire#level 5`) %>% 
  mutate(LC_mode = 'RP')

```

## Combine class info in a single df

```{r}

class_info <- rbind(name_class_rp, canopus_class_rp)

```


# Dynamic time warping - RP MODE


## Prepare matrix for DTW analysis 
## metabolites in rows, Month in columns - Time in order

## Working only with significant features

```{r}

data_dtw_RP <- norm_matrix_sig %>% 
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_RP')) %>% 
  group_by(FeatureID, month) %>% 
  summarise(mean_intensity = mean(norm_int)) %>% 
  pivot_wider(names_from = month, values_from = mean_intensity) %>% 
  column_to_rownames(var = "FeatureID")

data_ready_ord_RP <- data_dtw_RP[,c('May', 'June', 'July_1',
                                    'July_2', 'July_3', 'August', 'September',
                                    'October')]

```


## Dynamic time warping

```{r}

dwt_test_RP <- tsclust(data_ready_ord_RP, 
                       type = 'h', 
                       k = 4,
                       preproc = zscore,
                       distance = 'sbd',
                       centroid = shape_extraction,
                       control = hierarchical_control(method = 'average'))


dwt_test_RP@clusinfo

plot(dwt_test_RP, type = 'sc')

## Testing clustering

cvi(dwt_test_RP, type = c("Sil", "Dunn", "DB", "CH"))


```

### Classes of metabolites from Dwt

## Adding class info

## Final plot 

```{r}

clusts_rp <- tibble(FeatureID = names(dwt_test_RP@cluster),
                    cluster = paste0('Cluster ', dwt_test_RP@cluster))


```

## RP

```{r}
feature_dwt_clusts_rp <- as_tibble(dwt_test_RP@datalist) %>% 
  rowid_to_column(var = 'time') %>% 
  mutate(time = case_when(time == 1 ~ 'May',
                          time == 2 ~ 'June',
                          time == 3 ~ 'July_1',
                          time == 4 ~ 'July_2',
                          time == 5 ~ 'July_3',
                          time == 6 ~ 'August',
                          time == 7 ~ 'September',
                          time == 8 ~ 'October')) %>% 
  mutate(time = factor(time, levels = c('May', 'June', 'July_1', 'July_2', 'July_3',
                                        'August', 'September', 'October'))) %>% 
  pivot_longer(!time, names_to = 'FeatureID', values_to = 'value') %>% 
  left_join(clusts_rp, by = 'FeatureID') %>% 
  mutate(cluster = factor(cluster, levels = paste0('Cluster ', 1:4)),
         LC_mode = 'RP')


rename_vec <- set_names(1:length(unique(dwt_test_RP@cluster)),
                        paste0('Cluster ', unique(dwt_test_RP@cluster)))


```

## Centroids RP
```{r}
dwt_centroids_RP <- as.data.frame(dwt_test_RP@centroids) %>% 
  rename(rename_vec) %>% 
  rowid_to_column(var = 'time') %>% 
  pivot_longer(!time, names_to = 'cluster', values_to = 'value') %>% 
  mutate(cluster = factor(cluster, levels = paste0('Cluster ', 1:4)),
         LC_mode = 'RP') %>% 
  filter(cluster != 'Cluster 4')

```

## RP DTW plot

```{r}

combined_dtw <- feature_dwt_clusts_rp %>% 
  left_join(class_info, by = c('FeatureID', 'LC_mode')) %>% 
  filter(!is.na(superclass),
         cluster != 'Cluster 4')

class_colors <- combined_dtw %>% 
  dplyr::select(superclass) %>% 
  distinct() %>% 
  pull()

my_colors_class <- c(
  "Organoheterocyclic compounds"          = "#E69F00",  # orange
  "Organic acids and derivatives"         = "#56B4E9",  # sky blue
  "Organosulfur compounds"                = "#009E73",  # green
  "Benzenoids"                            = "#F0E442",  # yellow
  "Organic oxygen compounds"              = "#0072B2",  # dark blue
  "Organic nitrogen compounds"            = "#D55E00",  # red-orange
  "Phenylpropanoids and polyketides"     = "#CC79A7",  # pink/magenta
  "Lipids and lipid-like molecules"      = "#999999",  # gray
  "Alkaloids and derivatives"             = "#F0A3FF",  # light purple
  "Nucleosides, nucleotides, and analogues" = "#00BFC4" # cyan
)



feature_dwt_plot <- combined_dtw %>% 
  ggplot(aes(x = time, y = value)) +
  geom_line(aes(group = FeatureID, color = superclass),
            alpha = 0.4, linewidth = 0.5) +
  geom_line(data = . %>%
              group_by(cluster, time) %>%
              summarise(cluster_mean = mean(value), .groups = "drop"),
            aes(y = cluster_mean, group = cluster),
            color = "black", linewidth = 1.2) +
  geom_line(data = dwt_centroids_RP,
            aes(x = time, y = value, group = 1),
            linetype = "dashed", color = "gray25",
            linewidth = 1, alpha = 0.8) +
  geom_vline(xintercept = c(1.5, 5.5), linetype = "dashed", color = "gray40") +
  facet_wrap2(~cluster,
              scales = "free_y",
              ncol = 1,
              strip = strip_nested()) +
  labs(y = "Z-score (normalized abundance)",
       x = NULL,
       color = "Superclass") +
  scale_color_manual(values = my_colors_class)+
  theme_bw(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 6),
    axis.text.y = element_text(size = 6),
    axis.title.y = element_text(size = 8, face = "bold"),
    strip.text.x = element_text(face = "bold", size = 7),
    strip.background = element_rect(fill = "white", color = "black", linewidth = 0.3),
    legend.position = "none",
    legend.title = element_text(size = 8, face = "bold"),
    legend.text = element_text(size = 7),
    plot.title = element_text(size = 10, face = "bold", hjust = 0.5),
    plot.margin = margin(0.3, 0.3, 0.3, 0.3, "cm")) 

feature_dwt_plot

```

## Metabolite bubble plot
```{r}

plot_test <- combined_dtw %>% 
  filter(superclass != "Alkaloids and derivatives") %>% 
  dplyr::select(!c(time, value)) %>% distinct() %>% 
  group_by(cluster, LC_mode, superclass, class, subclass) %>% 
  dplyr::count() %>%
  ggplot() +
  geom_point(aes(x = factor(cluster),
                 y = subclass,
                 fill = superclass,
                 size = n),
             shape = 21, 
             color = "gray20", 
             alpha = 0.9) +
  facet_grid2(
    rows = vars(superclass),
    cols = vars(LC_mode),
    scales = "free",
    space = "free")+
  labs(y = "Subclass",
       size = "Feature count") +
  scale_fill_manual(values = my_colors_class) +
  guides(fill = 'none') +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid.major.y = element_line(color = "gray90", size = 0.2),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 6),
    strip.text.x = element_text(face = "bold", size = 5),
    strip.text.y = element_text(face = "bold", size = 5, angle = 0),
    strip.background = element_rect(fill = "white", color = "gray80"),
    legend.position = "bottom",
    legend.title = element_text(size = 6),
    legend.text = element_text(size = 5),
    panel.spacing = unit(0.2, "lines"),
    legend.key.size = unit(0.15, 'lines'),
    legend.box = 'vertical')

plot_test

```


## Final plot

```{r}

lay <- "
AABB
CCDE
CCDE
"

final <- free(plot_ordi) + free(venn_lme_diablo) + htmp_loadings + 
  feature_dwt_plot + plot_test +
  plot_layout(design = lay, widths = c(.1, 3.5, 1.5, 1.5), heights = c(1, 2, 2) ) +
  plot_annotation(tag_levels = list(c('A', 'B', 'C', '', 'D', 'E')))+
  theme(plot.margin = unit(c(.1, .1, .1, .1), 'lines'))

# final

ggsave("output_figures/Figure_5.png", final, dpi = 300,
       units = 'mm',
       height = 220, 
       width = 180)

ggsave("output_figures/Figure_5.svg", final, dpi = 300,
       units = 'mm', 
       height = 220, 
       width = 180)
```


## Pathways KOs

```{r}

ko_c1 <- microbe_c1 %>% 
  separate_longer_delim(name, delim = ";")

dry_path <- ko_c1 %>% 
  filter(GroupContrib == 'dry') %>% 
  dplyr::select(MAG, KO, pathway, Domain, Phylum, Class, Order, Family) %>% 
  group_by(KO, pathway) %>% 
  dplyr::count()


wet_path <- ko_c1 %>% 
  filter(GroupContrib == 'wet') %>% 
  dplyr::select(MAG, KO, pathway, Domain, Phylum, Class, Order, Family) %>% 
  group_by(KO,pathway) %>% 
  dplyr::count()


```

