---
title: "Figure 1"
author: "Viviana Freire Zapata"
date: "9/30/2025"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: '2'
    collapsed: false
    highlight: tango
    css: style.css
editor_options: 
  chunk_output_type: console
---

# Data Pre-processing

## 1. Setup

### Importing libraries

```{r libraries, message=FALSE, warning=FALSE}

library(easystats)
library(vegan)
library(GGally)
library(ggpubr)
library(glmnet)
library(ggbeeswarm)
library(lme4)
library(lmerTest)
library(readxl)
library(ggforce)
library(ggtext)
library(patchwork)
library(matrixStats)
library(tidyverse)
source('figure_scripts/functions_cdis_exploration.R')
source('figure_scripts/functions_cdis_norm_stats.R')
source('figure_scripts/functions_extra.R')
```

### Loading data 

```{r}
## Raw matrix

metaG <- read_tsv("input/coverage_coverm_aln.tsv")

metaT <- read_csv("output/geTMM_normalized_matrix_2025.csv")


```

## Loading metadata

```{r}

## Metadata

metadata <- read_csv("input/metadata_omics.csv")

metadata_T <- read_csv("input/metadata_omics.csv") %>% 
  filter(!is.na(SampleID_metaT)) %>% 
  mutate(treatment_code = str_replace_all(treatment_code, '_', '-') )

```

## Creating colors 

```{r}

month_list <- metadata %>% 
  mutate(month = factor(month, levels = c('May', 'June', 'July_1', 'July_2',
                                          'July_3', 'August', 'September', 
                                          'October', 'May_22'))) %>% 
  select(month) %>% 
  distinct() %>% 
  pull()

my_colors <- set_names(get_palette('Spectral', 12)[c(1:2, 9:12, 4, 5, 3)],
                          nm = levels(month_list))

stage_list <- metadata %>% 
  mutate(stage = factor(stage, levels = c('Pre_monsoon', 'During_monsoon', 
                                          'Post_monsoon'))) %>%
  select(stage) %>% 
  distinct() %>% 
  pull()

stage_colors <- set_names(c('#FEDD7F', '#9bc1bc', '#ed6a5a'),
                          nm = levels(stage_list))


```


### NMDS - TRANSCRIPTS

## Note: genes/metabolites as columns and samples as rows ##

Calculating distance matrix

```{r}
metaT <- metaT %>% 
  column_to_rownames(var = "Geneid")

## Filtering 

## Removing genes with zero counts 

metaT_ft <- metaT[rowSums(metaT >= 5) >= 3,]

```


```{r}
t_norm_matrix <- t(metaT_ft)

# distance matrix by Bray because relative abundance mode was selected
dm <- vegdist(t_norm_matrix[metadata_T$treatment_code,], method = 'bray')

```


Performing NMDS analysis

#NMDS stress
# > 0.2 is deemed suspect
# approaching 0.3 is arbitrary
# = < 0.1 fair
# = < 0.5 good


```{r}
set.seed(123)
nmds <- metaMDS(dm,
                k = 2,
                maxit = 999,
                trymax = 500,
                wascores = TRUE)
stressplot(nmds)

```


Extracting NMDS scores and plotting

```{r}

nmds_scores <- as.data.frame(scores(nmds, display = 'sites')) %>%
  rownames_to_column(var = 'SampleID') %>% 
  left_join(metadata_T, by = c('SampleID' = 'treatment_code'))

nmds_scores$month <- factor(nmds_scores$month, levels = c("May", "July_1", "July_2", "July_3", "August", "October", "May_22"))

nmds_scores$stage <- factor(nmds_scores$stage, levels = c("Pre_monsoon", "During_monsoon", "Post_monsoon"))

```


```{r}
## Testing PERMANOVA

# distance matrix by Bray because relative abundance mode was selected

dm_all <- vegdist(t_norm_matrix[metadata_T$treatment_code,], method = "bray")

set.seed(456)

permanova <- adonis2(dm_all ~ month,
                     data=metadata_T, 
                     permutations=999,
                     method="bray",
                     by = 'terms')

permanova

```

## get Centroid

```{r}

centroid <- nmds_scores %>% 
  group_by(stage) %>% 
  summarise(axis1 = mean(NMDS1),
            axis2 = mean(NMDS2)) %>% 
  ungroup()

```

## extract convex hull

```{r}

stage.hull <- nmds_scores %>% 
  group_by(stage) %>%
  slice(chull(NMDS1, NMDS2))

```

```{r}

nmds_plot_metaT <- nmds_scores %>% 
  ggplot(aes(x = NMDS1,
             y = NMDS2))+
  geom_point(aes(color = month),
             size = 2) +
  geom_point(data = centroid, 
             aes(x = axis1, y = axis2, 
                 color = stage), 
             size = 3, shape = 17)+
  geom_polygon(data = stage.hull, 
               aes(x = NMDS1, y = NMDS2, 
                   fill = stage, 
                   group = stage), 
               alpha = 0.30) +
  scale_color_manual(values = my_colors)+
  scale_fill_manual(values = stage_colors)+
  annotate("text", x = 0.55, y = -0.3,
           label = paste(c('Stress: ', round(nmds$stress, 4), "\n",
                           'Month pvalue: ', 
                           round(permanova$`Pr(>F)`[1], 4),"\n",
                           'Month R²: ', round(permanova$R2[1], 2)),
                         collapse = ''),
           size = 2.5)+
  labs(fill = "Monsoon stage", shape = "Monsoon stage",
       title = "Metatranscriptome")+
  theme_bw() +
  theme(plot.title = element_text(face = 'bold', hjust = 0.5),
        plot.subtitle = element_markdown(),
        legend.position = 'none',
        legend.box = 'vertical',
        legend.text = element_markdown(),
        panel.grid = element_blank()) 

nmds_plot_metaT


```


### NMDS - metaG

## Note: genes/metabolites as columns and samples as rows ##

Calculating distance matrix

```{r}
metaG <- metaG %>% 
  column_to_rownames(var = "Genome") 

metaG_ready <- metaG %>% 
  rename_with(.fn = function(x) str_remove(x, '_L000_R1_001.anqdpht.fastq.gz Trimmed Mean')) %>% 
  rownames_to_column(var = "Genome") %>% 
  pivot_longer(!Genome, names_to = "SampleID", values_to = "abundance") %>% 
  mutate(SampleID = str_remove(SampleID, "_S.*")) %>% 
  group_by(Genome, SampleID) %>% 
  summarise(abundance_mean = mean(abundance)) %>% 
  ungroup() %>% 
  pivot_wider(names_from = SampleID, values_from = abundance_mean) %>% 
  column_to_rownames(var = "Genome")


metadata_metaG <- metadata %>% 
  mutate(SampleID_g = str_replace_all(SampleID_metaT, "RNA", "DNA")) %>% 
  filter(!is.na(SampleID_metaT)) %>% 
  rename(SampleID = SampleID_g)

```


```{r}
t_norm_matrix <- t(metaG_ready)

# distance matrix by Bray because relative abundance mode was selected

dm <- vegdist(t_norm_matrix[metadata_metaG$SampleID,], method = 'bray')

```


Performing NMDS analysis

#NMDS stress
# > 0.2 is deemed suspect
# approaching 0.3 is arbitrary
# = < 0.1 fair
# = < 0.5 good


```{r}
set.seed(123)
nmds <- metaMDS(dm,
                k = 2,
                maxit = 999,
                trymax = 500,
                wascores = TRUE)
stressplot(nmds)
```


Extracting NMDS scores and plotting

```{r}

nmds_scores <- as.data.frame(scores(nmds, display = 'sites')) %>%
  rownames_to_column(var = 'SampleID') %>% 
  left_join(metadata_metaG, by = 'SampleID')

nmds_scores$month <- factor(nmds_scores$month, levels = c("May", "July_1", "July_2", "July_3", "August", "October", "May_22"))

nmds_scores$stage <- factor(nmds_scores$stage, levels = c("Pre_monsoon", "During_monsoon", "Post_monsoon"))

```


```{r}
## Testing PERMANOVA

# distance matrix by Bray because relative abundance mode was selected

dm_all <- vegdist(t_norm_matrix[metadata_metaG$SampleID,], method = "bray")

set.seed(456)

permanova <- adonis2(dm_all ~ month,
                     data=metadata_metaG, 
                     permutations=999,
                     method="bray",
                     by = 'terms')

permanova

```


## get Centroid

```{r}

centroid <- nmds_scores %>% 
  group_by(stage) %>% 
  summarise(axis1 = mean(NMDS1),
            axis2 = mean(NMDS2)) %>% 
  ungroup()

```

## extract convex hull

```{r}

stage.hull <- nmds_scores %>% 
  group_by(stage) %>%
  slice(chull(NMDS1, NMDS2))

```

```{r}

nmds_plot_metaG <- nmds_scores %>% 
  ggplot(aes(x = NMDS1,
             y = NMDS2))+
  geom_point(aes(color = month),
             size = 2) +
  geom_point(data = centroid, 
             aes(x = axis1, y = axis2, 
                 color = stage), 
             size = 3, shape = 17)+
  geom_polygon(data = stage.hull, 
               aes(x = NMDS1, y = NMDS2, 
                   fill = stage, 
                   group = stage), 
               alpha = 0.30) +
  scale_color_manual(values = my_colors)+
  scale_fill_manual(values = stage_colors)+
  annotate("text", x = -1, y = -0.46,
           label = paste(c('Stress: ', round(nmds$stress, 4), "\n",
                           'Month pvalue: ', 
                           round(permanova$`Pr(>F)`[1], 4),"\n",
                           'Month R²: ', round(permanova$R2[1], 2)),
                         collapse = ''),
           size = 2.5)+
  labs(fill = "Monsoon stage", shape = "Monsoon stage",
       title = "Metagenome")+
  theme_bw() +
  theme(plot.title = element_text(face = 'bold', hjust = 0.5),
        plot.subtitle = element_markdown(),
        legend.position = 'none',
        legend.box = 'vertical',
        legend.text = element_markdown(),
        panel.grid = element_blank()) 

nmds_plot_metaG


```


# LC-MS - RP

# Loading data 

```{r}
## Raw matrix

RP_df <- read_csv("scripts/LC-MS-MS/1.Pre-processing/output/RP_matrix_clean_2025.csv")


## Loading annotation

annot <- read_csv("scripts/LC-MS-MS/3.Cleanning_Annotation/output/RP_final_annotation_11_20_25.csv")

```


## Prepare matrix

```{r}

rp_ready <- RP_df %>% 
  select(FeatureID, contains("RP")) %>% 
  filter(FeatureID %in% annot$FeatureID) %>% 
  column_to_rownames(var = "FeatureID") 

```


# Feature filtering

```{r}
### Calculating IQR to keep only traits that are changing more
## Filtering metabolites and genes that doesn't change much
## Idea decrease the number of features to have a more balance matrices 

iqr_test_matrix <- rowIQRs(as.matrix(rp_ready))

iqr_res <- as.data.frame(as.matrix(iqr_test_matrix))

# Filtering according to metaboAnalyts

# 1) Less than 250 variables: 5% will be filtered; 
# 2) Between 250 - 500 variables: 10% will be filtered; 
# 3) Between 500 - 1000 variables: 25% will be filtered; and 
# 4) Over 1000 variables: 40% will be filtered.

## 10% of the features will be filtered

# Sorting from largest to smallest (higher ranking = low iqr)
rk <- rank(-iqr_test_matrix, ties.method = 'random')

# Determining the number of remaining features
remain.num <- nrow(rp_ready)*.9

remain <- rk <= remain.num

features_keep <- rownames(rp_ready)[remain] 

## Filter matrix

matrix_ft <- rp_ready[features_keep,]

## Note: 919 features kept for downstream analysis
```


# Metabolite Normalization

## Median normalization

```{r}

norm_matrix <- median.norm(matrix_ft, transform_data = TRUE)


```


### NMDS - LC-MS

Calculating distance matrix

```{r}
t_norm_matrix <- t(norm_matrix)

metadata_rp <- metadata %>% 
  filter(!is.na(SampleID_RP))


dm <- vegdist(t_norm_matrix[metadata_rp$SampleID_RP,], method = 'bray')

```


Performing NMDS analysis

#NMDS stress
# > 0.2 is deemed suspect
# approaching 0.3 is arbitrary
# = < 0.1 fair
# = < 0.5 good


```{r}
set.seed(123)
nmds <- metaMDS(dm,
                k = 2,
                maxit = 999,
                trymax = 500,
                wascores = TRUE)
stressplot(nmds)
```


Extracting NMDS scores and plotting

```{r}

nmds_scores <- as.data.frame(scores(nmds, display = 'sites')) %>%
  rownames_to_column(var = 'SampleID') %>% 
  left_join(metadata_rp, by = c('SampleID' = 'SampleID_RP'))

nmds_scores$month <- factor(nmds_scores$month, levels = c("May", "June", "July_1", "July_2", "July_3", "August", "September", "October"))

nmds_scores$stage <- factor(nmds_scores$stage, levels = c("Pre_monsoon", "During_monsoon", "Post_monsoon"))


```


```{r}
## Testing PERMANOVA

# distance matrix by Bray because relative abundance mode was selected

dm_all <- vegdist(t_norm_matrix[metadata_rp$SampleID_RP,], method = "bray")

set.seed(456)

permanova <- adonis2(dm_all ~ month,
                     data=metadata_rp, 
                     permutations=999,
                     strata = metadata_rp$site,
                     method="bray",
                     by = 'terms')

permanova

```

## get Centroid

```{r}

centroid <- nmds_scores %>% 
  group_by(stage) %>% 
  summarise(axis1 = mean(NMDS1),
            axis2 = mean(NMDS2)) %>% 
  ungroup()

```

## extract convex hull

```{r}

stage.hull <- nmds_scores %>% 
  group_by(stage) %>%
  slice(chull(NMDS1, NMDS2))

```

```{r}

nmds_plot_rp <- nmds_scores %>% 
  ggplot(aes(x = NMDS1,
             y = NMDS2))+
  # geom_mark_ellipse(aes(fill = stage),
  #                   alpha = 0.05,
  #                   show.legend = FALSE) +
  #scale_fill_manual(values = c('firebrick4', 'steelblue', 'orange')) +
  geom_point(aes(color = month),
             size = 2) +
  geom_point(data = centroid, 
             aes(x = axis1, y = axis2, 
                 color = stage), 
             size = 3, shape = 17)+
  geom_polygon(data = stage.hull, 
               aes(x = NMDS1, y = NMDS2, 
                   fill = stage, 
                   group = stage), 
               alpha = 0.30) +
  scale_color_manual(values = my_colors)+
  scale_fill_manual(values = stage_colors)+
  annotate("text", x = -0.01, y = -0.01,
           label = paste(c('Stress: ', round(nmds$stress, 4), "\n",
                           'Month pvalue: ', 
                           round(permanova$`Pr(>F)`[1], 4),"\n",
                           'Month R²: ', round(permanova$R2[1], 2)),
                         collapse = ''),
           size = 2.5)+
  labs(fill = "Monsoon stage", shape = "Monsoon stage",
       title = "Metabolomics - RP mode")+
  theme_bw() +
  theme(plot.title = element_text(face = 'bold', hjust = 0.5),
        plot.subtitle = element_markdown(),
        panel.grid = element_blank(),
        legend.position = 'none',
        legend.box = 'vertical',
        legend.text = element_markdown()) 

nmds_plot_rp


```


# Creating metadata for plot

```{r}

### Creating Month lengend

metadata_plot <- tibble(Month = c('May', 'June', 'July_1', 'July_2',
                                  'July_3', 'August', 'September', 
                                  'October', 'May_22'),
                        Stage = c('Pre monsoon', 'Pre monsoon', 'During monsoon', 
                                  'During monsoon', 'During monsoon', 'During monsoon',
                                  'Post monsoon','Post monsoon', 'Pre monsoon'))


plot_label <- metadata_plot %>% 
  mutate(Month = factor(Month, levels = c('May', 'June', 'July_1', 'July_2',
                                          'July_3', 'August', 'September', 
                                          'October', 'May_22'))) %>% 
  ggplot()+
  geom_point(aes(x = Month,
                 y = 1,
                 color = Month))+
  scale_color_manual(values = my_colors)+
  theme_bw()+
  theme(legend.position = 'bottom')

plot_label

leg_month <- get_legend(plot_label)


### Creating monsoon stage

stage_list <- metadata_plot %>% 
  dplyr::select(Stage) %>% 
  distinct() %>% 
  mutate(Stage = factor(Stage, levels = c('Pre monsoon', 'During monsoon',
                                          'Post monsoon')))


stage_colors2 <- set_names(c('#FEDD7F', '#9bc1bc', '#ed6a5a'),
                           nm = levels(stage_list))

plot_stage <- metadata_plot %>% 
  mutate(Stage = factor(Stage, levels = c('Pre monsoon', 'During monsoon',
                                          'Post monsoon'))) %>% 
  mutate(Month = factor(Month, levels = c('May', 'June', 'July_1', 'July_2',
                                          'July_3', 'August', 'September', 
                                          'October', 'May_22'))) %>% 
  ggplot()+
  geom_point(aes(x = Stage,
                 y = 1,
                 color = Stage),
             shape = 15)+
  scale_color_manual(values = stage_colors2) +
  ggnewscale::new_scale_color() +
  geom_point(aes(x = Month,
                 y = 1,
                 color = Month),
             shape = 16) +
  scale_color_manual(values = my_colors) +
  guides(color = guide_legend(nrow = 2)) +
  theme_bw()+
  theme(legend.position = 'bottom',
        legend.box = 'vertical',
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 6),
        legend.key.size = unit(.1, 'lines'),
        legend.key.spacing = unit(0.3, 'lines'),
        legend.spacing.y = unit(-0.1, 'lines'))

plot_stage

leg_stage <- get_legend(plot_stage)
leg_stage

```


## Expression clusters


## Loading HCLUST results

```{r}

hclust <- read_csv('output/Hclust_MAG_expression_2025.csv')

```


```{r}

plot_groups <- hclust %>% 
  rename(cluster = order) %>% 
  mutate(month = factor(month, levels = c('May', 'July_1', 'July_2', 'July_3', 'August',
                                          'October','May_22')),
         cluster = paste0('Cluster ', cluster),
         response_type = case_when(str_detect(cluster, 'Cluster 1') ~ 'Dry',
                                   str_detect(cluster, 'Cluster 2') ~ 'Wet',
                                   .default = 'Pulse'))

cluster_colors <- c('Dry' = '#F28482', 
                    'Wet' = '#608CD8',
                    'Pulse' = '#84A59D')


dry_mags <- plot_groups %>% 
  filter(response_type == 'Dry') %>% 
  select(genome) %>% 
  distinct()

wet_mags <- plot_groups %>% 
  filter(response_type == 'Wet') %>% 
  select(genome) %>% 
  distinct()

pulse_mags <- plot_groups %>% 
  filter(response_type == 'Pulse') %>% 
  select(genome) %>% 
  distinct()


dat_text <- data.frame(
  label = c("65 MAGs - DRY", "182 MAGs - WET", "32 MAGs - PULSE"),
  cluster   = c('Cluster 1', 'Cluster 2', 'Cluster 3'))


cluster_plot <- plot_groups %>% 
  ggplot(aes(x = month,
             y = mean_zs))+
  geom_line(aes(group = genome,
                color = response_type),
            alpha = 0.5)+
  geom_line(data = . %>% 
              group_by(cluster, month) %>% summarise(cluster_mean = mean(mean_zs)), 
            aes(y = cluster_mean, 
                group = cluster), 
            color = "gray15", size = 1) +
  geom_vline(xintercept = c(1.5, 5.5), linetype = 'dashed')+
  geom_text(data = dat_text,
            mapping = aes(x = -Inf, y = -Inf, label = label),
            hjust   = -0.1,
            vjust   = -1,
            size = 2) + 
  scale_color_manual(values = cluster_colors)+
  facet_grid(. ~ cluster, 
             scales = "free")+
  labs(y = 'geTMM mean per month\n per MAG (zscore)',
       color = 'Preference')+
  theme_bw()+
  theme(legend.position = 'none',
        strip.background.x = element_blank(),
        panel.grid = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 90))

cluster_plot



```


## Taxonomy of clusters

```{r}

taxonomy_clusters <- plot_groups %>% 
  select(!c(month, geTMM_month, mean_zs)) %>%
  distinct() %>% 
  mutate(Class = ifelse(is.na(Class), 
                        glue::glue('Unassigned Class of {Phylum}'), 
                        Class)) %>% 
  group_by(cluster, Phylum, Class, response_type) %>% 
  count() %>% 
  ungroup() %>%
  mutate(Class_lump = fct_lump_n(Class, w = n, n = 15)) %>%
  group_by(cluster, Class_lump) %>%
  summarise(n = sum(n)) %>%
  group_by(cluster) %>%
  mutate(perc = n / sum(n) *100)



tax_plot <- taxonomy_clusters %>% 
  group_by(Class_lump) %>%
  ggplot(aes(x = cluster,
             y = n,
             fill = Class_lump))+
  geom_col(position = 'stack',
           #color = 'black',
           alpha = 0.8,
           width = 0.5)+
  labs(y = 'Number of MAGs',
       fill = 'Class')+
  scale_fill_manual(values = ggpubr::get_palette(palette = 'Paired', 19)) +
  theme_bw() +
  theme(axis.text.y = element_text(size = 7),
        axis.text.x = element_text(size = 7),
        axis.title.x = element_blank(),
        legend.position = 'none',
        panel.grid = element_blank(),
        legend.text = element_text(size = 6),
        legend.key.size = unit(.5, 'lines'),
        legend.box.spacing = unit(.01, 'lines'),
        legend.title.position = 'top',
        legend.title = element_text(size = 7))

tax_plot



## Extracting legend


leg_clus <- taxonomy_clusters %>% 
  group_by(Class_lump) %>%
  ggplot(aes(x = cluster,
             y = n,
             fill = Class_lump))+
  geom_col(position = 'stack',
           #color = 'black',
           alpha = 0.8)+
  scale_fill_manual(values = ggpubr::get_palette(palette = 'Paired', 19)) +
  labs(fill = 'Class')+
  theme_bw() +
  theme(axis.text.y = element_text(size = 7),
        axis.text.x = element_text(size = 7),
        axis.title.x = element_blank(),
        legend.position = 'right',
        panel.grid = element_blank(),
        legend.text = element_text(size = 6),
        legend.key.size = unit(.5, 'lines'),
        legend.box.spacing = unit(.01, 'lines'),
        legend.title.position = 'top',
        legend.title = element_text(size = 7))


tax_legend <- get_legend(leg_clus)

tax_legend

```

### Saving supplementary table 1

```{r}

tax_temp <- plot_groups %>% 
  select(!c(month, geTMM_month, mean_zs)) %>%
  distinct() %>% 
  mutate(Class = ifelse(is.na(Class), 
                        glue::glue('Unassigned Class of {Phylum}'), 
                        Class))

write_csv(tax_temp, "output_tables/Supplementary_data_1.csv")


```


## RDA - LC-MS RP

# ENVS 

```{r}

evs <- read_csv("input/env_data_ready_2025.csv") %>% 
  filter(!sample == c('SAMP1-S1-22', 'SAMP1-S2-22', 'SAMP1-S3-22', 'SAMP1-S4-22'))%>% 
  mutate(sample = str_replace_all(sample, "-", "_")) %>% 
  rename(SampleID = sample)

```

# Microbial diversity

```{r}

avs_shannon <- read_csv("input/ASV_shannon.csv") %>% 
  mutate(sample = str_replace_all(sample, "-", "_")) %>% 
  rename(SampleID = sample) %>% 
  select(!dataset)

```


# Preprare matrix - envs and microbial diversity

```{r}

envs_microbe <- evs %>% 
  mutate(C_N = total_C/total_N) %>% 
  left_join(avs_shannon, by = 'SampleID') %>% 
  left_join(metadata, by = c('SampleID' = 'treatment_code')) %>% 
  dplyr::select(!c(SampleID, sampling, stage, month, site)) %>%
  rename(SampleID = SampleID_RP) %>% 
  column_to_rownames(var = 'SampleID') %>% 
  select(total_N, ph_mean, moisture, C_N, `NO3-N (ppm)`, `NH4-N (ppm)`, Shannon)

```


## Standarization of envs matrix

```{r}

env_std <- decostand(envs_microbe, method = 'standardize')

```

## Checking for collinearity 

```{r}
ggpairs(env_std)

```


## Running RDA

```{r}

## Transpose normalized matrix

rp_df2 <- as.data.frame(t(norm_matrix))

# Building model with all variables

rda_model <- rda(rp_df2 ~ ., data = env_std)

plot(performance::check_collinearity(rda_model))

# Quick plot to see how it is working

ordiplot(rda_model, scaling = 1, choices = c(1,2))

```


```{r}

summary(rda_model)

anova.cca(rda_model, step = 1000)

anova.cca(rda_model, step = 1000, by = "term")

```



```{r}

# Selecting variables ----

fwd.sel <- ordiR2step(rda(rp_df2 ~ 1, data = env_std), # simple model,
                      scope = rda_model, # full model
                      direction = 'forward',
                      R2scope = TRUE,
                      pstep = 1000,
                      trace = TRUE,
                      na.action = na.exclude) 

fwd.sel$call

```


## Model based on foward selection

```{r}

envs_fwd_sel <- envs_microbe %>% 
  select(moisture, ph_mean, C_N)

env_std3 <- decostand(envs_fwd_sel, method = 'standardize')


# Building model with all variables

rda_model_fwd <- rda(rp_df2 ~ ., data = env_std3)


```

## Check R2 of the new model
```{r}


RsquareAdj(rda_model_fwd)
## 0.2411

```


## Testing the singnificace of the model


```{r}

anova.cca(rda_model_fwd, step = 1000)

anova.cca(rda_model_fwd, step = 1000, by = 'term')


total_variance <- sum(rda_model_fwd$CCA$eig)/(sum(rda_model_fwd$CCA$eig)+ sum(rda_model_fwd$CA$eig)) *100

```


## RDA plot 

```{r}

#### Scaling 2

# Custom triplot code!

## extract % explained by the first 2 axes
perc <- round(100*(summary(rda_model_fwd)$cont$importance[2, 1:2]), 2)

## extract scores - these are coordinates in the RDA space
sc_sites <- as.data.frame(scores(rda_model_fwd, display="sites", choices=c(1,2), 
                                 scaling=2)) %>%
  rownames_to_column(var = 'SampleID') %>% 
  left_join(metadata_rp, by = c('SampleID' = 'SampleID_RP')) %>% 
  mutate(color = my_colors[month]) 


sc_species <- as.data.frame(scores(rda_model_fwd, display="species", choices=c(1,2), scaling=2))

sc_bp <- as.data.frame(scores(rda_model_fwd, display="bp", choices=c(1, 2), scaling=2)) %>% 
  rownames_to_column(var = 'envs')



RDA_plot_rp <- sc_sites %>% 
  ggplot()+
  geom_point(aes(x = RDA1,
                 y = RDA2,
                 color = month),
             size = 2) +
  scale_color_manual(values = my_colors)+
  geom_point(data = sc_species,
             aes(x = RDA1,
                 y = RDA2),
             fill = 'gray80',
             size = 1,
             shape = 22) +
  geom_segment(data = sc_bp, 
               aes(x = 0, y = 0, 
                   xend = 3*RDA1, yend = 3*RDA2), 
               color = "black", 
               linewidth = 0.5,
               arrow = arrow(length = unit(.5, 'lines'))) +
  geom_text(data = sc_bp,
            aes(x = 3*RDA1+.3, y = 3*RDA2 +.3,
                label = envs),
            color = "black",
            fontface = "bold",
            size = 2) +
  labs(title = "Metabolomics - RDA - Scaling 2",
       x = paste0("RDA1 (", perc[1], "%)" ) ,
       y = paste0("RDA2 (", perc[2], "%)"))+
  theme_bw() +
  theme(plot.title = element_text(face = 'bold', hjust = 0.5),
        plot.subtitle = element_markdown(),
        panel.grid = element_blank(),
        legend.position = 'none',
        legend.box = 'vertical',
        legend.text = element_markdown()) 

RDA_plot_rp

```



## Saving plot

```{r}

## Final figure
des <- "
ABB
CDE
FGG
HHH
"


final_fig <- nmds_plot_metaG + nmds_plot_metaT + free(cluster_plot) +
  tax_plot + tax_legend + nmds_plot_rp + RDA_plot_rp + leg_stage +
  plot_layout(heights = c(1, 1, 1, .1),
              widths = c(2, 1, 1),
              design = des) +
  plot_annotation(tag_levels = list(c('A', 'B', 'C', 'D', '', 'E', 'F'))) &
  theme(text = element_text(size = 7))

final_fig

```



## Save final plot


```{r}

# ggsave("output_figures/Figure_1.svg", final_fig, dpi = 300,
#        units = 'mm', 
#        height = 220, 
#        width = 180)

ggsave("output_figures//Figure_1.png", final_fig, dpi = 300,
       units = 'mm', 
       height = 220, 
       width = 180)


```

