---
title: "Figure 3"
output: html_notebook
editor_options: 
  chunk_output_type: console
---


```{r}
library(patchwork)
library(ggh4x)
library(readxl)
library(treedataverse)
library(ggpubr)
library(vegan)
library(ggtext)
library(ggforce)
library(reshape2)
library(rstatix)
library(ggsankey)
library(ggrepel)
library(tidyverse)

```

# Loading data

## Metadata

```{r}
metadata <- read_csv('input/metadata_omics.csv') 

metadata_ready <- metadata %>% 
  mutate(Month = factor(month, levels = c("May", "July_1", "July_2", "July_3",
                                          "August", "October", "May_22")))
```

## Taxonomy

```{r}

mag_taxonomy <- read_csv('input/mag_taxonomy_denovo.csv') %>% 
  mutate(Class = ifelse(is.na(Class), 
                        glue::glue('Unassigned Class \nof {Phylum}'),
                        Class))

```


## MAGs annotations

```{r}

# mag_completeness <- read_tsv("../../../input/annotations_merged.tsv") %>%
#   rename(Geneid = `...1`) %>% 
#   select(fasta, contains('bin')) %>% 
#   distinct()

annot_cydbs <- read_csv("input/annot_with_cydbs_2.23.25.csv") %>% 
  rename(Geneid = `...1`)

```


## geTMM

```{r}
geTMM <- read_csv("output/geTMM_normalized_matrix_2025.csv")

geTMM_long <- geTMM %>% 
  pivot_longer(!Geneid, names_to = "sampleid", values_to = "geTMM")

```

## Loading HCLUST results

```{r}

hclust <- read_csv('output/Hclust_MAG_expression_2025.csv')


plot_groups <- hclust %>% 
  rename(cluster = order) %>% 
  mutate(cluster = paste0('Cluster ', cluster),
         response_type = case_when(str_detect(cluster, 'Cluster 1') ~ 'Dry',
                                   str_detect(cluster, 'Cluster 2') ~ 'Wet',
                                   .default = 'Pulse')) %>% 
  select(genome, cluster, response_type) %>% 
  distinct()

```


## Joining expression to annotations

```{r}

annot_vip <- geTMM_long %>% 
  left_join(annot_cydbs, by = 'Geneid') %>%
  group_by(sampleid, fasta, ko_id) %>% 
  summarise(getmm_sum = sum(geTMM)) %>% 
  mutate(sampleid = str_replace_all(sampleid, '-', '_')) %>% 
  left_join(metadata_ready, by = c('sampleid' = 'treatment_code')) %>% 
  group_by(fasta, ko_id, Month) %>% 
  summarise(getmm_mean = mean(getmm_sum)) %>% 
  filter(getmm_mean > 0) %>% 
  left_join(mag_taxonomy, by = c('fasta' = 'bin')) %>% 
  left_join(plot_groups, by = c('fasta' = 'genome'))

```


## Expression calculated reaction level

```{r}

rx_xprss <- read.csv("input/Definitions_2025/output/Reaction_expression_0.5_per_sample_final.csv") %>% 
  rename('ID' = 'module_id')

def <- read_csv("input/Definitions_2025/reaction_level_definitions_may_2025.csv")


```


## Joining expression with annotation of modules

```{r}

delete <- c('Anammox', 'Commamox')

kegg_ready <- rx_xprss %>% 
  left_join(def, by = 'ID') %>% 
  filter(!Category_1 %in% delete)


## Step 1 = filter those modules with expression zero in all samples

rx_ft_kegg <- kegg_ready %>% 
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_metaT')) %>% 
  group_by(ID, bin, Month) %>% 
  summarise(expression = mean(expression)) %>%
  left_join(def, by = 'ID') %>%
  filter(expression > 0) %>% 
  left_join(mag_taxonomy, by = 'bin') %>% 
  left_join(plot_groups, by = c('bin' = 'genome')) %>% 
  ungroup()

```

### Checking expressed pathway completeness - 

```{r}
percent_rx_ft_kegg <- kegg_ready %>%
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_metaT')) %>%
  group_by(ID, bin, Month) %>%
  summarise(expression = mean(expression)) %>%
  left_join(def, by = 'ID') %>%
  group_by(Category_2, Category_1, bin, Month) %>%
  mutate(n_steps = n()) %>%
  filter(expression > 0) %>%
  mutate(steps_present = n(),
         percent_steps_expressed = steps_present / n_steps) %>%
  select(ID, bin, Month, Category_1, Category_2, percent_steps_expressed) %>%
  left_join(mag_taxonomy, by = 'bin')

```


```{r}

number_mags_exp <- percent_rx_ft_kegg %>%
  ungroup() %>%
  filter(percent_steps_expressed >= 0.5) %>%
  select(Category_1, bin) %>%
  distinct() %>%
  group_by(Category_1) %>%
  count()


```


## Keeping only MAGs that have at least 50% of the pathway steps

```{r}

mags_ft <- percent_rx_ft_kegg %>%
  ungroup() %>%
  filter(percent_steps_expressed >= 0.5) %>%
  select(ID, bin) %>%
  distinct()

```


## Line plots - Key pathways


```{r}

category_1_keep <- c('Glycolysis (Embden-Meyerhof pathway)', 
                     'Glycolysis, core module involving three-carbon compounds',
                     'Pentose P pathway, oxidative phase',
                     'Pentose P pathway, non-oxidative phase',
                     'TCA',
                     'Pyruvate oxidation',
                     'PRPP biosynthesis',
                     'Nitrification',
                     'Denitrification',
                     'Dissimilatory nitrate reduction',
                     'Assimilatory nitrate reduction',
                     'Assimilatory sulfate reduction',
                     'Dissimilatory sulfate reduction',
                     'Glyoxylate cycle',
                     'Ethylmalonyl pathway')

```


```{r}


rx_line <- kegg_ready %>%
  inner_join(mags_ft, by = c('ID', 'bin')) %>% ## Filtering MAGs pathway > 50
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_metaT')) %>%
  group_by(ID, SampleID, Month) %>%
  summarise(expression = sum(expression)) %>%
  left_join(def, by = 'ID') %>%
  group_by(Category_2, Category_1, Month) %>%
  summarise(expression = mean(expression)) %>%
  filter(Category_1 %in% category_1_keep) %>% 
  mutate(Category_1 = ifelse(Category_1 == 'Glycolysis (Embden-Meyerhof pathway)',
                             'Glycolysis', Category_1),
         Category_1 = ifelse(Category_1 == 'Glycolysis, core module involving three-carbon compounds', 'Glycolysis 3 carbon', Category_1),
         Category_1 = ifelse(Category_1 == 'Pentose P pathway, oxidative phase',
                             'Pentose P, oxidative phase', Category_1),
         Category_1 = ifelse(Category_1 == 'Pentose P pathway, non-oxidative phase',
                             'Pentose P, non-oxidative phase', Category_1)) %>% 
  filter(expression > 0) %>%
  split(.$Category_2)

```


## Line plot - Community Level

```{r}

stages <- tibble(
  stage = c('Pre monsoon', 'During monsoon', 'Post monsoon', 'Pre monsoon'),
  start = c(-Inf, 1.5, 5.5, 6.5),
  end = c(1.5, 5.5, 6.5, Inf)
)


```

## Figure for Carbon

```{r}

rx_line_C <- kegg_ready %>%
  inner_join(mags_ft, by = c('ID', 'bin')) %>% ## Filtering MAGs pathway > 50
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_metaT')) %>%
  group_by(ID, SampleID, Month) %>%
  summarise(expression = sum(expression)) %>%
  left_join(def, by = 'ID') %>%
  group_by(Category_2, Category_1, Month) %>%
  summarise(expression = mean(expression)) %>%
  filter(Category_1 %in% category_1_keep) %>% 
  mutate(Category_1 = ifelse(Category_1 == 'Glycolysis (Embden-Meyerhof pathway)',
                             'Glycolysis', Category_1),
         Category_1 = ifelse(Category_1 == 'Glycolysis, core module involving three-carbon compounds', 'Glycolysis 3 C', Category_1),
         Category_1 = ifelse(Category_1 == 'Pentose P pathway, oxidative phase',
                             'PPP_ox', Category_1),
         Category_1 = ifelse(Category_1 == 'Pentose P pathway, non-oxidative phase',
                             'PPP_non-ox', Category_1)) %>% 
  filter(expression > 0) %>%
  filter(Category_2 == 'Central carbon metabolism',
         Category_1 != 'Glycolysis 3 C')

## Set colors

modules_colors_C <- c(
  'Glycolysis' = "#E64B35",               
  #'Glycolysis 3 C' = "#D08770",      
  'PPP_non-ox' = "#6A3D9A",  
  'PPP_ox' = "#00A087",     
  'Pyruvate oxidation' = "#3C5488",       
  'TCA' = "#7E6148"                        
)




## Plot

line_c_plot <- rx_line_C %>%
  ggplot() +
  geom_rect(data = stages,
            aes(xmin = start,
                xmax = end,
                ymin = -Inf,
                ymax = Inf,
                fill = stage),
            alpha = .15,
            show.legend = FALSE) +
  geom_line(aes(x = Month, y = expression,
                color = Category_1,
                group = Category_1),
            linewidth = 0.5) +
  geom_point(aes(x = Month, y = expression,
                 color = Category_1,
                 group = Category_1),
             size = 1.5) +
  geom_label_repel(aes(x = Month, 
                       y = expression,
                       label = Category_1,
                       color = Category_1),
                   data = rx_line_C %>% filter(Month == 'May_22'),
                   nudge_x = 1, 
                   nudge_y = 10,
                   size = 2,
                   show.legend = FALSE,
                   segment.color = "grey60",
                   box.padding = 0.3)+
  scale_fill_manual(values = c('Pre monsoon' = '#F2F2F2', 
                               'During monsoon' = '#7BBBE5', 
                               'Post monsoon' = '#E8F0FA')) +
  scale_color_manual(values = modules_colors_C)+
  scale_y_continuous(limits = c(0, NA)) +
  labs(title = 'Carbon metabolism',
       y = 'geTMM')+
  guides(color = guide_legend(ncol =2)) +
  theme_bw(base_size = 9, base_family = "sans") +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", linewidth = 0.4),
    plot.title = element_text(size = 7, face = "bold", hjust = 0.5),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 6, face = "bold"),
    axis.text = element_text(color = "black", size = 6),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.ticks = element_line(color = "black", linewidth = 0.3),
    legend.position = "none",
    plot.margin = margin(5, 5, 5, 5)
  )

line_c_plot

```


## Figures - Nitrogen

```{r}

rx_line_N <- kegg_ready %>%
  inner_join(mags_ft, by = c('ID', 'bin')) %>% ## Filtering MAGs pathway > 50
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_metaT')) %>%
  group_by(ID, SampleID, Month) %>%
  summarise(expression = sum(expression)) %>%
  left_join(def, by = 'ID') %>%
  group_by(Category_2, Category_1, Month) %>%
  summarise(expression = mean(expression)) %>%
  filter(Category_1 %in% category_1_keep) %>% 
  mutate(Category_1 = ifelse(Category_1 == 'Assimilatory nitrate reduction',
                             'ANR', Category_1),
         Category_1 = ifelse(Category_1 == 'Dissimilatory nitrate reduction', 
                             'DNR', Category_1)) %>% 
  filter(expression > 0) %>%
  filter(Category_2 == 'Nitrogen metabolism')


### Plot Nitrogen


modules_colors_N <- c('ANR' = 'darkblue',
                      'DNR' = "#6A3D9A",
                      'Denitrification' = "#7E6148"  ,
                      'Nitrification' = '#E64B35')

## Plot

line_N_plot <- rx_line_N %>%
  ggplot() +
  geom_rect(data = stages,
            aes(xmin = start,
                xmax = end,
                ymin = -Inf,
                ymax = Inf,
                fill = stage),
            alpha = .15,
            show.legend = FALSE) +
  geom_line(aes(x = Month, y = expression,
                color = Category_1,
                group = Category_1),
            linewidth = 0.5) +
  geom_point(aes(x = Month, y = expression,
                 color = Category_1,
                 group = Category_1),
             size = 1.5) +
  geom_label_repel(aes(x = Month, 
                       y = expression,
                       label = Category_1,
                       color = Category_1),
                   data = rx_line_N %>% filter(Month == 'May_22'),
                   nudge_x = 1, 
                   nudge_y = 10,
                   size = 2,
                   show.legend = FALSE,
                   segment.color = "grey60",
                   box.padding = 0.3)+
  scale_fill_manual(values = c('Pre monsoon' = '#F2F2F2', 
                               'During monsoon' = '#7BBBE5', 
                               'Post monsoon' = '#E8F0FA')) +
  scale_color_manual(values = modules_colors_N)+
  scale_y_continuous(limits = c(0, NA)) +
  labs(title = 'Nitrogen metabolism',
       y = 'geTMM')+
  guides(color = guide_legend(ncol =2)) +
  theme_bw(base_size = 9, base_family = "sans") +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", linewidth = 0.4),
    plot.title = element_text(size = 7, face = "bold", hjust = 0.5),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 6, face = "bold"),
    axis.text = element_text(color = "black", size = 6),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.ticks = element_line(color = "black", linewidth = 0.3),
    legend.position = "none",
    plot.margin = margin(5, 5, 5, 5)
  )

line_N_plot

```

## Figures - Sulfur

```{r}

rx_line_S <- kegg_ready %>%
  inner_join(mags_ft, by = c('ID', 'bin')) %>% ## Filtering MAGs pathway > 50
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_metaT')) %>%
  group_by(ID, SampleID, Month) %>%
  summarise(expression = sum(expression)) %>%
  left_join(def, by = 'ID') %>%
  group_by(Category_2, Category_1, Month) %>%
  summarise(expression = mean(expression)) %>%
  filter(Category_1 %in% category_1_keep) %>% 
  mutate(Category_1 = ifelse(Category_1 == 'Assimilatory sulfate reduction',
                             'ASR', Category_1),
         Category_1 = ifelse(Category_1 == 'Dissimilatory sulfate reduction', 
                             'DSR', Category_1)) %>% 
  filter(expression > 0) %>%
  filter(Category_2 == 'Sulfur metabolism')


### Plot Sulfur


modules_colors_S <- c('ASR' = 'brown',
                      'DSR' = 'purple4')

## Plot

line_S_plot <- rx_line_S %>% 
  ggplot() +
  geom_rect(data = stages,
            aes(xmin = start,
                xmax = end,
                ymin = -Inf,
                ymax = Inf,
                fill = stage),
            alpha = .15,
            show.legend = FALSE) +
  geom_line(aes(x = Month, y = expression,
                color = Category_1,
                group = Category_1),
            linewidth = 0.5) +
  geom_point(aes(x = Month, y = expression,
                 color = Category_1,
                 group = Category_1),
             size = 1.5) +
  geom_label_repel(aes(x = Month, 
                       y = expression,
                       label = Category_1,
                       color = Category_1),
                   data = rx_line_S %>% filter(Month == 'May_22'),
                   nudge_x = 1, 
                   nudge_y = 10,
                   size = 2,
                   show.legend = FALSE,
                   segment.color = "grey60",
                   box.padding = 0.3)+
  scale_fill_manual(values = c('Pre monsoon' = '#F2F2F2', 
                               'During monsoon' = '#7BBBE5', 
                               'Post monsoon' = '#E8F0FA')) +
  scale_color_manual(values = modules_colors_S)+
  scale_y_continuous(limits = c(0, NA)) +
  labs(title = 'Sulfur metabolism',
       y = 'geTMM')+
  guides(color = guide_legend(ncol =2)) +
  theme_bw(base_size = 9, base_family = "sans") +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", linewidth = 0.4),
    plot.title = element_text(size = 7, face = "bold", hjust = 0.5),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 6, face = "bold"),
    axis.text = element_text(color = "black", size = 6),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.ticks = element_line(color = "black", linewidth = 0.3),
    legend.position = "none",
    plot.margin = margin(5, 5, 5, 5)
  )


line_S_plot


```


## Line plot -- Class level -- Supplementary


```{r}


rx_line_class <- kegg_ready %>%
  inner_join(mags_ft, by = c('ID', 'bin')) %>% ## Filtering MAGs pathway > 50
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_metaT')) %>%
  left_join(mag_taxonomy, by = 'bin') %>%
  group_by(ID, SampleID, Month, Class) %>%
  summarise(expression = sum(expression)) %>%
  left_join(def, by = 'ID') %>%
  group_by(Category_2, Category_1, Month, Class) %>%
  summarise(expression = mean(expression)) %>% 
  filter(Category_1 %in% category_1_keep) %>% 
  mutate(Category_1 = ifelse(Category_1 == 'Glycolysis (Embden-Meyerhof pathway)',
                             'Glycolysis', Category_1),
         Category_1 = ifelse(Category_1 == 'Glycolysis, core module involving three-carbon compounds', 'Glycolysis 3 carbon', Category_1),
         Category_1 = ifelse(Category_1 == 'Pentose P pathway, oxidative phase',
                             'Pentose P, oxidative phase', Category_1),
         Category_1 = ifelse(Category_1 == 'Pentose P pathway, non-oxidative phase',
                             'Pentose P, non-oxidative phase', Category_1)) %>% 
  filter(expression > 0) 


```

## Filter top 20% expressed Class per module

```{r}

top_expressed_class <- rx_line_class %>%
  ungroup() %>% 
  group_by(Category_1, Class) %>%
  summarise(mean_exp = mean(expression),
            total_exp = sum(expression)) %>% 
  group_by(Category_1) %>% 
  slice_max(order_by = mean_exp, prop = 0.2) %>% 
  pull(Class) %>% 
  unique()

## Prepare for map

rx_line_class_2 <- rx_line_class %>% 
  filter(Class %in% top_expressed_class) %>% 
  split(.$Category_1)

```


## Line plot - Class Level - Supplementary

```{r}


## Set color CLASS

class_mags <-reduce(rx_line_class_2, rbind) %>% 
  ungroup() %>% 
  select(Class) %>% 
  distinct() %>% 
  pull()


class_color <- set_names(ggpubr::get_palette('npg', 8),
                         nm = sort(class_mags))


line_map_class <- imap(rx_line_class_2, function(cat, name) {
  
  line_nito_plot <- cat %>%
    ggplot() +
    geom_rect(data = stages,
              aes(xmin = start,
                  xmax = end,
                  ymin = -Inf,
                  ymax = Inf,
                  fill = stage),
              alpha = .3,
              show.legend = FALSE) +
    geom_line(aes(x = Month, y = expression,
                  color = Class,
                  group = Class),
              linewidth = 0.7) +
    geom_point(aes(x = Month, y = expression,
                   color = Class,
                   group = Class),
               size = 1) +
    scale_fill_manual(values = c('Pre monsoon' = '#F2F2F2', 
                                 'During monsoon' = '#7BBBE5', 
                                 'Post monsoon' = '#E8F0FA')) +
    scale_color_manual(values = class_color)+
    scale_y_continuous(limits = c(0, NA)) +
    labs(y = 'geTMM',
         color = 'Class',
         title = name)+
    #guides(color = guide_legend(ncol =2)) +
    theme_bw() +
    theme(
      panel.grid = element_blank(),                        # remove grid lines
      plot.title = element_text(size = 6, face = "bold", hjust = 0.5),  # centered, small title
      axis.title.x = element_blank(),                     # no x-axis title
      axis.title.y = element_text(size = 7, face = "bold"),
      axis.text = element_text(size = 6),                # smaller axis text
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.ticks.length = unit(0.15, "cm"),              # shorter ticks
      legend.title = element_text(size = 6, face = "bold"),
      legend.text = element_text(size = 6),
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.box = "horizontal",
      legend.key.size = unit(0.35, "lines"),             # smaller legend keys
      legend.title.position = "top",
      plot.margin = margin(2, 2, 2, 2)                  # smaller margins around plot
    )
  
  return(line_nito_plot)
  
  
})


for_legend <- data.frame(Class = class_mags) %>% 
  ggplot(aes(x = 1, y = Class, color = Class)) +
  geom_point(size = 2) +  # smaller points to save space
  scale_color_manual(values = class_color) +
  theme_bw(base_size = 7) + 
  theme(
    panel.grid = element_blank(),                  # remove grid
    panel.border = element_blank(),                # optional: remove border
    plot.title = element_text(size = 7, face = "bold", hjust = 0.5),
    axis.title = element_blank(),                  # remove axis titles
    axis.text = element_blank(),                   # remove axis text
    axis.ticks = element_blank(),                  # remove ticks
    legend.title = element_text(size = 7, face = "bold"),
    legend.text = element_text(size = 6),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "horizontal",
    legend.key.size = unit(0.35, "lines"),
    legend.spacing.x = unit(0.2, "cm"),
    legend.title.position = "top",
    plot.margin = margin(1, 1, 1, 1)             # minimal margins
  ) +
  guides(
    color = guide_legend(nrow = 1, byrow = TRUE)
  )


legend <- ggpubr::get_legend(for_legend)
```


```{r}

keep_plot <- c(6, 7, 8, 11, 10, 9, 1, 3, 4, 2)

line_map_filtered <- line_map_class[keep_plot]

des <- "
ABCDE
FGHIJ
KKKKK
"

plot_filt <- reduce(line_map_filtered, `+`) +
  legend +
  plot_layout(design = des,
              heights = c(1, 1, .1)) &
  theme(legend.position = 'none')


plot_filt
```


## Save final plot


```{r}

ggsave("output_figures/Supp_Figure_3.svg", plot_filt, dpi = 300,
       units = 'mm', 
       height = 160, 
       width = 180)

ggsave("output_figures/Supp_Figure_3.png", plot_filt, dpi = 300,
       units = 'mm', 
       height = 160, 
       width = 180)



```


## Other pathways

```{r}

keep_plot_2 <- c(12, 13)

line_map_filtered <- line_map_class[keep_plot_2]

des <- "
AABB
KKKK
"

plot_filt <- reduce(line_map_filtered, `+`) + legend +
  plot_layout(design = des,
              heights = c(1,.1)) &
  theme(legend.position = 'none')


plot_filt

```



```{r}

ggsave("output_figures/Supp_Figure_4.svg", plot_filt, dpi = 300,
       units = 'mm', 
       height = 100, 
       width = 180)

ggsave("output_figures/Supp_Figure_4.png", plot_filt, dpi = 300,
       units = 'mm', 
       height = 100, 
       width = 180)


```


## Expression at CLass level -- key genes

## Expression key genes - Heatmap

```{r}

key_genes <- read_csv("input/Key_genes_3_18_25.csv")


```


```{r}

annot_class_ko <- geTMM_long %>% 
  left_join(annot_cydbs, by = 'Geneid') %>%
  filter(!is.na(ko_id)) %>% 
  left_join(plot_groups, by = c('fasta' = 'genome')) %>% 
  left_join(mag_taxonomy, by = c('fasta' = 'bin')) %>% 
  group_by(sampleid, Class, response_type, ko_id) %>% 
  summarise(getmm_sum = sum(geTMM)) %>%
  mutate(sampleid = str_replace_all(sampleid, '-', '_')) %>% 
  left_join(metadata_ready, by = c('sampleid' = 'treatment_code')) %>%
  group_by(Class, ko_id, response_type, Month) %>% 
  summarise(getmm_mean = mean(getmm_sum)) %>% 
  filter(getmm_mean > 0) 


```


```{r}

annot_class_vip <- annot_class_ko %>% 
  inner_join(key_genes, by = 'ko_id')

## Plot only top expressed

top_kos <- annot_class_vip %>%
  group_by(Class, Month, response_type) %>%
  slice_max(order_by = getmm_mean, prop = 0.2) %>%
  select(ko_id) %>%
  pull() %>%
  unique()

supp_ko <- c("Gluconeogenesis", "EMC", "Glyoxylate bypass", "Glycine cleavage",
             "Arginine biosynthesis", "methionine sulfoxide reductases", "Other")

ko_vip_heatmap <- annot_class_vip %>% 
  filter(ko_id %in% top_kos,
         !Category_2 %in% supp_ko) %>%
  mutate(Category_2 = case_when(str_detect(Category_2, 'Carbon fixation') ~'C fixation',
                                str_detect(Category_2, 'Pentose phosphate pathway') ~'PPP',
                                str_detect(Category_2, 'Nitrogen metabolism') ~'N metabolism',
                                str_detect(Category_2, 'Sulfur metabolism') ~'S metabolism',
                                str_detect(Category_2, 'Phosphorous metabolism') ~'P metabolism',
                                .default = Category_2),
         Category_2 = str_replace(Category_2, ' ', '\n')) %>% 
  mutate(Category_2 = factor(Category_2, 
                             levels = c('C\nfixation',
                                        'TCA',
                                        'PPP',
                                        'Glycolisis',
                                        'Piruvate\noxidation',
                                        'N\nmetabolism',
                                        'S\nmetabolism',
                                        'P\nmetabolism'))) 


top_class <- ko_vip_heatmap %>% 
  filter(response_type != 'Pulse') %>% 
  group_by(ko_id, response_type, Class) %>% 
  summarise(mean_exp = mean(getmm_mean)) %>% 
  slice_max(order_by = mean_exp, prop = 0.1) %>% 
  pull(Class) %>% 
  unique()



class_vip_ko_vip <- ko_vip_heatmap %>% 
  filter(Class %in% top_class) %>% 
  group_by(gene_abb, Category_2, Class, response_type) %>%
  mutate(zscore = (getmm_mean - mean(getmm_mean)) / sd(getmm_mean))


```

## Heatmap key genes

```{r}

htmp_plot <-  class_vip_ko_vip %>% 
  filter(response_type != 'Pulse',
         !is.na(Category_2)) %>% 
  ggplot() +
  geom_tile(aes(x = Month,
                y = gene_abb,
                fill = zscore)) + 
  # scale_fill_viridis_c(option = 'A', direction = -1) +
  scale_fill_distiller(palette = 'RdBu', direction = -1) +
  facet_grid2(cols = vars(response_type, Class),
              rows = vars(Category_2),
              switch = 'y',
              scales = 'free',
              space = 'free',
              strip = strip_nested(
                text_x = elem_list_text(angle = c(0, 0, rep(0, 100))),
                text_y = elem_list_text(angle = c(rep(90, 8), rep(0, 100)),
                                        face = c(rep('bold', 8), 
                                                 rep('plain', 100))),
                size = 'variable',
                clip = 'off'
              )) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.title = element_blank(),
        strip.background = element_blank(),
        strip.text = element_text(size = 4, margin = margin(b = 0, t = 0)),
        strip.text.y.left = element_text(angle = 0, margin = margin(r = 0, l = 0)),
        strip.placement = "outside",      # closer to panels
        panel.spacing = unit(0.1, "lines"),
        axis.text.y = element_text(size = 5),
        axis.text.x = element_text(size = 5, angle = 45, hjust = 1),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 5),
        legend.position = 'bottom',
        legend.key.size = unit(.5, 'lines'),
        legend.spacing.x = unit(0.1, "cm"),
        plot.margin = margin(0.1, 0.1, 0.1, 0.1, 'lines'))

htmp_plot

```

## Stitching figure

```{r}
layout <- "
ABC
DDD
"

final_fig <- line_c_plot + line_N_plot + line_S_plot + free(htmp_plot) +
  plot_layout(design = layout, heights = c(1, 4)) +
  plot_annotation(tag_levels = 'A')


final_fig

ggsave('output_figures/Figure_3.png', 
       dpi = 300,
       units = 'mm', 
       height = 200, 
       width = 180)
ggsave('output_figures/Figure_3.svg', 
       dpi = 300,
       units = 'mm', 
       height = 200, 
       width = 180)
```

## Other pathways - Supplementary heatmap

```{r}

ko_vip_heatmap_2 <- annot_class_vip %>% 
  filter(ko_id %in% top_kos,
         Category_2 %in% supp_ko)


top_class <- ko_vip_heatmap_2 %>% 
  filter(response_type != 'Pulse') %>% 
  group_by(ko_id, response_type, Class) %>% 
  summarise(mean_exp = mean(getmm_mean)) %>% 
  slice_max(order_by = mean_exp, prop = 0.2) %>% 
  pull(Class) %>% 
  unique()



class_vip_ko_vip <- ko_vip_heatmap_2 %>% 
  filter(Class %in% top_class) %>% 
  group_by(gene_abb, Category_2, Class, response_type) %>%
  mutate(zscore = (getmm_mean - mean(getmm_mean)) / sd(getmm_mean))

```

```{r}


supp_htmp_plot <-  class_vip_ko_vip %>% 
  filter(response_type != 'Pulse',
         !is.na(Category_2)) %>% 
  ggplot() +
  geom_tile(aes(x = Month,
                y = gene_abb,
                fill = zscore)) + 
  # scale_fill_viridis_c(option = 'A', direction = -1) +
  scale_fill_distiller(palette = 'RdBu', direction = -1) +
  facet_grid2(cols = vars(response_type, Class),
              rows = vars(Category_2),
              switch = 'y',
              scales = 'free',
              space = 'free',
              strip = strip_nested(
                text_x = elem_list_text(angle = c(0, 0, rep(0, 100))),
                text_y = elem_list_text(angle = c(rep(90, 8), rep(0, 100)),
                                        face = c(rep('bold', 8), 
                                                 rep('plain', 100))),
                size = 'variable',
                clip = 'off'
              )) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.title = element_blank(),
        strip.background = element_blank(),
        strip.text = element_text(size = 4, margin = margin(b = 0, t = 0)),
        strip.text.y.left = element_text(angle = 0, margin = margin(r = 0, l = 0)),
        strip.placement = "outside",      # closer to panels
        panel.spacing = unit(0.1, "lines"),
        axis.text.y = element_text(size = 5),
        axis.text.x = element_text(size = 5, angle = 45, hjust = 1),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 5),
        legend.position = 'bottom',
        legend.key.size = unit(.5, 'lines'),
        legend.spacing.x = unit(0.1, "cm"),
        plot.margin = margin(0.1, 0.1, 0.1, 0.1, 'lines'))

supp_htmp_plot


ggsave('output_figures/Supp_Figure_5.png', 
       dpi = 300,
       units = 'mm', 
       height = 200, 
       width = 180)
ggsave('output_figures/Supp_Figure_5.svg', 
       dpi = 300,
       units = 'mm', 
       height = 200, 
       width = 180)

```


## Expression at MAG level -- key genes

## Expression key genes - Heatmap


```{r}

annot_MAG_ko <- geTMM_long %>% 
  left_join(annot_cydbs, by = 'Geneid') %>%
  filter(!is.na(ko_id)) %>% 
  left_join(plot_groups, by = c('fasta' = 'genome')) %>% 
  group_by(sampleid, fasta, response_type, ko_id) %>% 
  summarise(getmm_sum = sum(geTMM)) %>%
  mutate(sampleid = str_replace_all(sampleid, '-', '_')) %>% 
  left_join(metadata_ready, by = c('sampleid' = 'treatment_code')) %>%
  group_by(fasta, ko_id, response_type, Month) %>% 
  summarise(getmm_mean = mean(getmm_sum)) %>% 
  left_join(mag_taxonomy, by = c('fasta' = 'bin')) %>% 
  filter(getmm_mean > 0) 


```


```{r}

annot_MAG_vip <- annot_MAG_ko %>% 
  inner_join(key_genes, by = 'ko_id')

write_csv(annot_MAG_vip, 'output_tables/Supplementary_table_7.csv')

## Plot only top expressed

top_kos <- annot_MAG_vip %>%
  group_by(fasta, Month, response_type) %>%
  slice_max(order_by = getmm_mean, prop = 0.1) %>% 
  select(ko_id) %>% 
  pull() %>% 
  unique()

supp_ko <- c("Gluconeogenesis", "EMC", "Glyoxylate bypass", "Glycine cleavage",
             "Arginine biosynthesis", "methionine sulfoxide reductases", "Other")

ko_vip_heatmap <- annot_MAG_vip %>% 
  filter(ko_id %in% top_kos,
         !Category_2 %in% supp_ko) %>%
  mutate(Category_2 = case_when(str_detect(Category_2, 'Carbon fixation') ~'C fixation',
                                str_detect(Category_2, 'Pentose phosphate pathway') ~'PPP',
                                str_detect(Category_2, 'Nitrogen metabolism') ~'N metabolism',
                                str_detect(Category_2, 'Sulfur metabolism') ~'S metabolism',
                                str_detect(Category_2, 'Phosphorous metabolism') ~'P metabolism',
                                .default = Category_2),
         Category_2 = str_replace(Category_2, ' ', '\n')) %>% 
  mutate(Category_2 = factor(Category_2, 
                             levels = c('C\nfixation',
                                        'TCA',
                                        'PPP',
                                        'Glycolisis',
                                        'Piruvate\noxidation',
                                        'N\nmetabolism',
                                        'S\nmetabolism',
                                        'P\nmetabolism'))) 

mag_vip_ko <- ko_vip_heatmap %>% 
  group_by(gene_abb, Category_2, fasta, response_type) %>%
  mutate(zscore = (getmm_mean - mean(getmm_mean)) / sd(getmm_mean)) %>% 
  split(.$Category_2)


```

## Heatmap key genes

```{r}

mag_heatmap <- map(mag_vip_ko, function(cat2, name){
  
  htmp_plot <-  cat2 %>% 
    filter(Month %in% c('May', 'July_1', 'October')) %>% 
    ggplot() +
    geom_tile(aes(x = Month,
                  y = fasta,
                  fill = zscore)) + 
    # scale_fill_viridis_c(option = 'A', direction = -1) +
    scale_fill_distiller(palette = 'RdBu', direction = -1) +
    facet_grid2(cols = vars(gene_abb),
                rows = vars(Class, Order),
                switch = 'y',
                scales = 'free',
                space = 'free') +
    theme_bw() +
    theme(panel.grid = element_blank(),
          axis.title = element_blank(),
          strip.background = element_blank(),
          strip.text = element_text(size = 5, margin = margin(b = 0, t = 0)),
          strip.text.y.left = element_text(angle = 0, margin = margin(r = 0, l = 0)),
          strip.placement = "outside",      
          panel.spacing = unit(0.1, "lines"),
          axis.text.y = element_blank(),
          axis.text.x = element_text(size = 5, angle = 45, hjust = 1),
          legend.title = element_text(size = 7),
          legend.text = element_text(size = 5),
          legend.position = 'bottom',
          legend.key.size = unit(.5, 'lines'),
          legend.spacing.x = unit(0.1, "cm"),
          plot.margin = margin(0.1, 0.1, 0.1, 0.1, 'lines'))
  
  return(htmp_plot)
  
})

mag_heatmap$`Piruvate
oxidation`

```

