---
title: "Figure 3"
output: html_notebook
editor_options: 
  chunk_output_type: console
---


```{r}
library(patchwork)
library(ggh4x)
library(ggpubr)
library(ggtext)
library(ggforce)
library(ggrepel)
library(taxonomizr)
library(tidyverse)

```

# Loading data

## Metadata

```{r}
metadata <- read_csv('input/metadata_omics.csv') 

metadata_ready <- metadata %>% 
  mutate(Month = factor(month, levels = c("May", "July_1", "July_2", "July_3",
                                          "August", "October", "May_22")))
```

## Taxonomy

```{r}

mag_taxonomy <- read_csv('input/mag_taxonomy_denovo.csv') %>% 
  mutate(Class = ifelse(is.na(Class), 
                        glue::glue('Unassigned Class \nof {Phylum}'),
                        Class))

```


## MAGs annotations

```{r}

annot_cydbs <- read_csv("input/annot_with_cydbs_2.23.25.csv") %>% 
  rename(Geneid = `...1`)

```


## geTMM

```{r}
geTMM <- read_csv("output/geTMM_normalized_matrix_2025.csv")

geTMM_long <- geTMM %>% 
  pivot_longer(!Geneid, names_to = "sampleid", values_to = "geTMM")

```

## Loading HCLUST results

```{r}

hclust <- read_csv('output/Hclust_MAG_expression_2025.csv')


plot_groups <- hclust %>% 
  rename(cluster = order) %>% 
  mutate(cluster = paste0('Cluster ', cluster),
         response_type = case_when(str_detect(cluster, 'Cluster 1') ~ 'Dry',
                                   str_detect(cluster, 'Cluster 2') ~ 'Wet',
                                   .default = 'Pulse')) %>% 
  select(genome, cluster, response_type) %>% 
  distinct()

```


## Joining expression to annotations

```{r}

annot_vip <- geTMM_long %>% 
  left_join(annot_cydbs, by = 'Geneid') %>%
  group_by(sampleid, fasta, ko_id) %>% 
  summarise(getmm_sum = sum(geTMM)) %>% 
  mutate(sampleid = str_replace_all(sampleid, '-', '_')) %>% 
  left_join(metadata_ready, by = c('sampleid' = 'treatment_code')) %>% 
  group_by(fasta, ko_id, Month) %>% 
  summarise(getmm_mean = mean(getmm_sum)) %>% 
  filter(getmm_mean > 0) %>% 
  left_join(mag_taxonomy, by = c('fasta' = 'bin')) %>% 
  left_join(plot_groups, by = c('fasta' = 'genome'))

```


## Expression calculated reaction level

```{r}

rx_xprss <- read.csv("input/Definitions_2025/output/Reaction_expression_0.5_per_sample_final.csv") %>% 
  rename('ID' = 'module_id')

def <- read_csv("input/Definitions_2025/reaction_level_definitions_may_2025.csv")


```


## Joining expression with annotation of modules

```{r}

delete <- c('Anammox', 'Commamox')

kegg_ready <- rx_xprss %>% 
  left_join(def, by = 'ID') %>% 
  filter(!Category_1 %in% delete)


## Step 1 = filter those modules with expression zero in all samples

rx_ft_kegg <- kegg_ready %>% 
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_metaT')) %>% 
  group_by(ID, bin, Month) %>% 
  summarise(expression = mean(expression)) %>%
  left_join(def, by = 'ID') %>%
  filter(expression > 0) %>% 
  left_join(mag_taxonomy, by = 'bin') %>% 
  left_join(plot_groups, by = c('bin' = 'genome')) %>% 
  ungroup()

```

### Checking expressed pathway completeness - 

```{r}

percent_rx_ft_kegg <- kegg_ready %>%
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_metaT')) %>%
  group_by(ID, bin, Month) %>%
  summarise(expression = mean(expression)) %>%
  left_join(def, by = 'ID') %>%
  group_by(Category_2, Category_1, bin, Month) %>%
  mutate(n_steps = n()) %>%
  filter(expression > 0) %>%
  mutate(steps_present = n(),
         percent_steps_expressed = steps_present / n_steps) %>%
  select(ID, bin, Month, Category_1, Category_2, percent_steps_expressed) %>%
  left_join(mag_taxonomy, by = 'bin')

```


```{r}

number_mags_exp <- percent_rx_ft_kegg %>%
  ungroup() %>%
  filter(percent_steps_expressed >= 0.5) %>%
  dplyr::select(Category_1, bin) %>%
  distinct() %>%
  group_by(Category_1) %>%
  dplyr::count()


```


## Keeping only MAGs that have at least 50% of the pathway steps

```{r}

mags_ft <- percent_rx_ft_kegg %>%
  ungroup() %>%
  filter(percent_steps_expressed >= 0.5) %>%
  select(ID, bin) %>%
  distinct()

```


## MetaCyc

## List of only Prokariote paths

```{r}

pathway_metacyc <- read_tsv("input/MetaCyc_db_2026/All-pathways-of-MetaCyc_path_name.txt") %>% 
  select(Pathways, `Common-Name`, `Engineered?`, Species) %>% 
  filter(is.na(`Engineered?`)) %>% 
  separate_rows(Species, sep = " // ") %>% 
  filter(str_detect(Species, 'TAX')) %>% 
  mutate(Species = str_remove_all(Species, 'TAX-'))


## ADD taxa

taxids <- unique(pathway_metacyc$Species)

lineage_dia <- getTaxonomy(
  taxids,
  sqlFile = "C:/Users/vivia/Escritorio/Arid_metaT-LCMS_2025/Arid_2_code/ncbi_taxonomy.sqlite")

lineage_df <- as.data.frame(lineage_dia) %>%
  tibble::rownames_to_column("staxids") %>% 
  mutate(staxids = as.character(staxids)) %>% 
  mutate(staxids = str_trim(staxids))

## Final table


final_path_proka <- pathway_metacyc  %>% 
  left_join(lineage_df, by = c('Species' = 'staxids')) %>% 
  filter(domain != 'Eukaryota') %>% 
  select(Pathways) %>% 
  distinct() %>% 
  pull()

```


```{r}

## Loading MetaCyc pathways with at least >50% expression

arid_path_meta <- read_csv("scripts/LC_metat_network/output/MetaCyc_arid_pathway.csv") %>% 
  filter(is.na(`Engineered?`)) %>% 
  filter(Pathways %in% final_path_proka)


number_mags_exp_metacy <- arid_path_meta %>% 
  ungroup() %>%
  select(Pathways, bin) %>%
  distinct() %>%
  group_by(Pathways) %>%
  dplyr::count() %>% 
  left_join(pathway_metacyc, by = 'Pathways') %>% 
  select(Pathways, n, `Common-Name`) %>% 
  distinct()



```

## Pathway count per MAG

```{r}

category_1_keep <- c('Glycolysis (Embden-Meyerhof pathway)', 
                     'Glycolysis, core module involving three-carbon compounds',
                     'Pentose P pathway, oxidative phase',
                     'Pentose P pathway, non-oxidative phase',
                     'TCA',
                     'Pyruvate oxidation',
                     'PRPP biosynthesis',
                     'Nitrification',
                     'Denitrification',
                     'Dissimilatory nitrate reduction',
                     'Assimilatory nitrate reduction',
                     'Assimilatory sulfate reduction',
                     'Dissimilatory sulfate reduction',
                     'Glyoxylate cycle',
                     'Ethylmalonyl pathway',
                     'Reductive glycine pathway',
                     'Calvin cycle',
                     'Reductive citrate cycle',
                     "Glycine cleavage system")

metacyc_keep <- c('PWY-5484', 'ANAGLYCOLYSIS-PWY', 'GLYCOLYSIS',
                  'PWY-8178', 'NONOXIPENT-PWY', 'OXIDATIVEPENT-PWY',
                  'TCA', 'PYRUVDEHYD-PWY', 'PWY0-662', 'AMMOXID-PWY',
                  'PWY490-3', 'PWY0-1581', 'SO4ASSIM-PWY', 'PWY1ZNC-1',
                  'GLYOXDEG-PWY', 'GLYOXYLATE-BYPASS', 'PWY-5741', 'PWY-8303',
                  'CALVIN-PWY', 'P42-PWY', 'P23-PWY', 'PWY-8303', 'GLYCLEAV-PWY',
                  'PWY-6964',
                  'PWY-1881',
                  'PWY-7750',
                  'PWY-2301')



kegg_count_vip <- number_mags_exp %>% 
  filter(Category_1 %in% category_1_keep) %>% 
   mutate(pathway_category = case_when(
    
    Category_1 %in% c(
      "Glycolysis (Embden-Meyerhof pathway)",
      "Glycolysis, core module involving three-carbon compounds",
      "Pentose P pathway, oxidative phase",
      "Pentose P pathway, non-oxidative phase",
      "TCA",
      "Pyruvate oxidation",
      "Glyoxylate cycle",
      "Ethylmalonyl pathway"
    ) ~ "C metabolism",
    
    Category_1 %in% c(
      "Calvin cycle",
      "Reductive citrate cycle",
      "Reductive glycine pathway"
    ) ~ "C fixation",
    
    Category_1 %in% c(
      "Nitrification",
      "Denitrification",
      "Dissimilatory nitrate reduction",
      "Assimilatory nitrate reduction"
    ) ~ "N metabolism",
    
    Category_1 %in% c(
      "Assimilatory sulfate reduction",
      "Dissimilatory sulfate reduction"
    ) ~ "S metabolism",
    
    Category_1 == "PRPP biosynthesis" ~ "Biosynthesis",
    
    Category_1 == "Glycine cleavage system" ~ "C1 metabolism",
    
    TRUE ~ "Other"))

metacyc_count_vip <- number_mags_exp_metacy %>% 
  filter(Pathways %in% metacyc_keep) %>% 
  mutate(name_fix = str_remove_all(`Common-Name`, "\\<sub\\>"),
         name_fix = str_remove_all(name_fix, "\\<\\/sub\\>"),
         name_fix = str_remove_all(name_fix, "\\<\\i\\>"),
         name_fix = str_remove_all(name_fix, "\\<\\/i\\>")) %>% 
   mutate(pathway_category = case_when(

    str_detect(name_fix,
               "glycolysis|pentose phosphate|glyoxylate|glycolate|ethylmalonyl|TCA cycle|pyruvate") ~ 
      "C metabolism",

    str_detect(name_fix,
               "Calvin|reductive TCA|reductive glycine") ~ 
      "C fixation",

    str_detect(name_fix,
               "ammonia oxidation|nitrate reduction|ammonia assimilation") ~ 
      "N metabolism",

    str_detect(name_fix,
               "sulfate reduction") ~ 
      "S metabolism",

    str_detect(name_fix,
               "formate oxidation|carbon monoxide oxidation|glycine cleavage") ~ 
      "C1 metabolism",

    str_detect(name_fix,
               "PRPP biosynthesis|inositol biosynthesis") ~ 
      "Biosynthesis",

    TRUE ~ "Other"
  ))

## KEGG

plot_count <- kegg_count_vip %>% 
  ggplot(aes(x = n,
             y = fct_reorder(Category_1, n)))+
  labs(y = 'KEGG Module',
       x = 'Expressed MAGs')+
  facet_grid2(rows = vars(pathway_category),
              space = 'free',
              scales = 'free')+
  geom_col(fill = '#264653')+
  theme_classic()+
  theme(
        strip.background = element_blank(),
        strip.text.y = element_text(angle = 0, face = "bold", size = 6),
        axis.text.y = element_text(size = 5),
        axis.text.x = element_text(size = 5),
        axis.title.x = element_text(size = 6),
        axis.ticks.y = element_blank(),
        plot.margin = margin(2, 2, 2, 2))

plot_count

## MetaCyc

plot_count_2 <- metacyc_count_vip %>% 
  ggplot(aes(x = n,
             y = fct_reorder(name_fix, n))) +
  geom_col(fill = "#2a9d8f", width = 0.7) +
  facet_grid2(
    rows = vars(pathway_category),
    space = "free",
    scales = "free"
  ) +
  labs(
    x = "Expressed MAGs",
    y = 'MetaCyc Pathway'
  ) +
  theme_classic(base_size = 8) +
  theme(
    strip.background = element_blank(),
    strip.text.y = element_text(angle = 0, face = "bold", size = 6),
    axis.text.y = element_text(size = 5),
    axis.text.x = element_text(size = 5),
    axis.title.x = element_text(size = 6),
    axis.ticks.y = element_blank(),
    plot.margin = margin(2, 2, 2, 2))

plot_count_2



```

## MetaCyc pathways expression -- Supplementary

```{r}

metacyc_class <- read_csv("scripts/LC_metat_network/output/MetaCyc_arid_pathway_CLASS.csv")

meta_class_ft <- metacyc_class %>% 
  filter(Pathways %in% metacyc_keep) %>% 
  group_by(Pathways, `Common-Name`, Class) %>%
  mutate(zscore = (path_expr_month - mean(path_expr_month)) / sd(path_expr_month)) %>% 
  mutate(name_fix = str_remove_all(`Common-Name`, "\\<sub\\>"),
         name_fix = str_remove_all(name_fix, "\\<\\/sub\\>"),
         name_fix = str_remove_all(name_fix, "\\<\\i\\>"),
         name_fix = str_remove_all(name_fix, "\\<\\/i\\>")) %>% 
   mutate(pathway_category = case_when(

    str_detect(name_fix,
               "glycolysis|pentose phosphate|glyoxylate|glycolate|ethylmalonyl|TCA cycle|pyruvate") ~ 
      "Carbon metabolism",

    str_detect(name_fix,
               "Calvin|reductive TCA|reductive glycine") ~ 
      "Carbon fixation",

    str_detect(name_fix,
               "ammonia oxidation|nitrate reduction|ammonia assimilation") ~ 
      "Nitrogen metabolism",

    str_detect(name_fix,
               "sulfate reduction") ~ 
      "Sulfur metabolism",

    str_detect(name_fix,
               "formate oxidation|carbon monoxide oxidation|glycine cleavage") ~ 
      "C1 metabolism",

    str_detect(name_fix,
               "PRPP biosynthesis|inositol biosynthesis") ~ 
      "Biosynthesis",

    TRUE ~ "Other"
  ))


## Filter top 20% CLASS

top_expressed_class <- meta_class_ft %>%
  ungroup() %>% 
  group_by(name_fix, Class) %>%
  summarise(mean_exp = mean(path_expr_month),
            total_exp = sum(path_expr_month)) %>% 
  group_by(name_fix) %>% 
  slice_max(order_by = mean_exp, prop = 0.2) %>% 
  pull(Class) %>% 
  unique()

metacyc_heat <- meta_class_ft %>% 
  filter(Class %in% top_expressed_class) %>% 
  mutate(month = factor(month, levels = c("May", "July_1", "July_2", "July_3",
                                          "August", "October", "May_22"))) %>% 
  ggplot() +
  geom_tile(aes(x = month,
                y = name_fix,
                fill = zscore)) + 
  # scale_fill_viridis_c(option = 'A', direction = -1) +
  scale_fill_distiller(palette = 'RdBu', direction = -1) +
  facet_grid2(cols = vars(Class),
              rows = vars(pathway_category),
              switch = 'y',
              scales = 'free',
              space = 'free',
              strip = strip_nested(
                text_x = elem_list_text(angle = c(0, 0, rep(0, 100))),
                text_y = elem_list_text(angle = c(rep(90, 8), rep(0, 100)),
                                        face = c(rep('bold', 8), 
                                                 rep('plain', 100))),
                size = 'variable',
                clip = 'off'
              )) +
  labs(title = 'Metatranscriptomics - MetaCyc')+
  theme_bw() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = 'bold', hjust = 0.5),
        axis.title = element_blank(),
        strip.background = element_blank(),
        strip.text = element_text(size = 4, margin = margin(b = 0, t = 0)),
        strip.text.y.left = element_text(angle = 0, margin = margin(r = 0, l = 0)),
        strip.placement = "outside",      # closer to panels
        panel.spacing = unit(0.1, "lines"),
        axis.text.y = element_text(size = 5),
        axis.text.x = element_text(size = 5, angle = 45, hjust = 1),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 5),
        legend.position = 'bottom',
        legend.key.size = unit(.5, 'lines'),
        legend.spacing.x = unit(0.1, "cm"),
        plot.margin = margin(0.1, 0.1, 0.1, 0.1, 'lines'))

metacyc_heat


ggsave("output_figures/Supp_Figure_7.png", metacyc_heat, dpi = 300,
       units = 'mm', 
       height = 160, 
       width = 180)


```


