---
title: "Figure 3"
output: html_notebook
editor_options: 
  chunk_output_type: console
---


```{r}
library(patchwork)
library(ggh4x)
library(readxl)
library(ggpubr)
library(vegan)
library(ggtext)
library(ggforce)
library(reshape2)
library(rstatix)
library(ggsankey)
library(ggrepel)
library(tidyverse)

```

# Loading data

## Metadata

```{r}
metadata <- read_csv('input/metadata_omics.csv') 

metadata_ready <- metadata %>% 
  mutate(Month = factor(month, levels = c("May", "July_1", "July_2", "July_3",
                                          "August", "October", "May_22")))
```

## Taxonomy

```{r}

mag_taxonomy <- read_csv('input/mag_taxonomy_denovo.csv') %>% 
  mutate(Class = ifelse(is.na(Class), 
                        glue::glue('Unassigned Class \nof {Phylum}'),
                        Class))

```


## MAGs annotations

```{r}

annot_cydbs <- read_csv("input/annot_with_cydbs_2.23.25.csv") %>% 
  rename(Geneid = `...1`)

```


## geTMM

```{r}
geTMM <- read_csv("output/geTMM_normalized_matrix_2025.csv")

geTMM_long <- geTMM %>% 
  pivot_longer(!Geneid, names_to = "sampleid", values_to = "geTMM")

```

## Loading HCLUST results

```{r}

hclust <- read_csv('output/Hclust_MAG_expression_2025.csv')


plot_groups <- hclust %>% 
  rename(cluster = order) %>% 
  mutate(cluster = paste0('Cluster ', cluster),
         response_type = case_when(str_detect(cluster, 'Cluster 1') ~ 'Dry',
                                   str_detect(cluster, 'Cluster 2') ~ 'Wet',
                                   .default = 'Pulse')) %>% 
  select(genome, cluster, response_type) %>% 
  distinct()

```


## Joining expression to annotations

```{r}

annot_vip <- geTMM_long %>% 
  left_join(annot_cydbs, by = 'Geneid') %>%
  group_by(sampleid, fasta, ko_id) %>% 
  summarise(getmm_sum = sum(geTMM)) %>% 
  mutate(sampleid = str_replace_all(sampleid, '-', '_')) %>% 
  left_join(metadata_ready, by = c('sampleid' = 'treatment_code')) %>% 
  group_by(fasta, ko_id, Month) %>% 
  summarise(getmm_mean = mean(getmm_sum)) %>% 
  filter(getmm_mean > 0) %>% 
  left_join(mag_taxonomy, by = c('fasta' = 'bin')) %>% 
  left_join(plot_groups, by = c('fasta' = 'genome'))

```


## Expression calculated reaction level

```{r}

rx_xprss <- read.csv("input/Definitions_2025/output/Reaction_expression_0.5_per_sample_final.csv") %>% 
  rename('ID' = 'module_id')

def <- read_csv("input/Definitions_2025/reaction_level_definitions_may_2025.csv")


```


## Joining expression with annotation of modules

```{r}

delete <- c('Anammox', 'Commamox')

kegg_ready <- rx_xprss %>% 
  left_join(def, by = 'ID') %>% 
  filter(!Category_1 %in% delete)


## Step 1 = filter those modules with expression zero in all samples

rx_ft_kegg <- kegg_ready %>% 
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_metaT')) %>% 
  group_by(ID, bin, Month) %>% 
  summarise(expression = mean(expression)) %>%
  left_join(def, by = 'ID') %>%
  filter(expression > 0) %>% 
  left_join(mag_taxonomy, by = 'bin') %>% 
  left_join(plot_groups, by = c('bin' = 'genome')) %>% 
  ungroup()

```

### Checking expressed pathway completeness - 

```{r}
percent_rx_ft_kegg <- kegg_ready %>%
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_metaT')) %>%
  group_by(ID, bin, Month) %>%
  summarise(expression = mean(expression)) %>%
  left_join(def, by = 'ID') %>%
  group_by(Category_2, Category_1, bin, Month) %>%
  mutate(n_steps = n()) %>%
  filter(expression > 0) %>%
  mutate(steps_present = n(),
         percent_steps_expressed = steps_present / n_steps) %>%
  select(ID, bin, Month, Category_1, Category_2, percent_steps_expressed) %>%
  left_join(mag_taxonomy, by = 'bin')

```


```{r}

number_mags_exp <- percent_rx_ft_kegg %>%
  ungroup() %>%
  filter(percent_steps_expressed >= 0.5) %>%
  select(Category_1, bin) %>%
  distinct() %>%
  group_by(Category_1) %>%
  count()


```


## Keeping only MAGs that have at least 50% of the pathway steps

```{r}

mags_ft <- percent_rx_ft_kegg %>%
  ungroup() %>%
  filter(percent_steps_expressed >= 0.5) %>%
  select(ID, bin) %>%
  distinct()

```


## Line plots - Key pathways


```{r}

category_1_keep <- c('Glycolysis (Embden-Meyerhof pathway)', 
                     'Glycolysis, core module involving three-carbon compounds',
                     'Pentose P pathway, oxidative phase',
                     'Pentose P pathway, non-oxidative phase',
                     'TCA',
                     'Pyruvate oxidation',
                     'PRPP biosynthesis',
                     'Nitrification',
                     'Denitrification',
                     'Dissimilatory nitrate reduction',
                     'Assimilatory nitrate reduction',
                     'Assimilatory sulfate reduction',
                     'Dissimilatory sulfate reduction',
                     'Glyoxylate cycle',
                     'Ethylmalonyl pathway',
                     'Reductive glycine pathway',
                     'Calvin cycle',
                     'Reductive citrate cycle')

```


```{r}


rx_line <- kegg_ready %>%
  inner_join(mags_ft, by = c('ID', 'bin')) %>% ## Filtering MAGs pathway > 50
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_metaT')) %>%
  group_by(ID, SampleID, Month) %>%
  summarise(expression = sum(expression)) %>%
  left_join(def, by = 'ID') %>%
  group_by(Category_2, Category_1, Month) %>%
  summarise(expression = mean(expression)) %>%
  filter(Category_1 %in% category_1_keep) %>% 
  mutate(Category_1 = ifelse(Category_1 == 'Glycolysis (Embden-Meyerhof pathway)',
                             'Glycolysis', Category_1),
         Category_1 = ifelse(Category_1 == 'Glycolysis, core module involving three-carbon compounds', 'Glycolysis 3 carbon', Category_1),
         Category_1 = ifelse(Category_1 == 'Pentose P pathway, oxidative phase',
                             'Pentose P, oxidative phase', Category_1),
         Category_1 = ifelse(Category_1 == 'Pentose P pathway, non-oxidative phase',
                             'Pentose P, non-oxidative phase', Category_1)) %>% 
  filter(expression > 0) %>%
  split(.$Category_2)

```


## Loading MetaCyc

```{r}

meta_comm <- read_csv("scripts/LC_metat_network/output/MetaCyc_arid_pathway_COMMUNITY.csv")


metacyc_keep <- c('PWY-5484', 'ANAGLYCOLYSIS-PWY', 'GLYCOLYSIS',
                  'PWY-8178', 'NONOXIPENT-PWY', 'OXIDATIVEPENT-PWY',
                  'TCA', 'PYRUVDEHYD-PWY', 'PWY0-662', 'AMMOXID-PWY',
                  'PWY490-3', 'PWY0-1581', 'SO4ASSIM-PWY', 'PWY1ZNC-1',
                  'GLYOXDEG-PWY', 'GLYOXYLATE-BYPASS', 'PWY-5741', 'PWY-8303',
                  'CALVIN-PWY', 'P42-PWY', 'P23-PWY', 'PWY-8303', 'GLYCLEAV-PWY',
                  'PWY-6964',
                  'PWY-1881',
                  'PWY-7750',
                  'PWY-2301')

## Filter 

meta_comm_ft <- meta_comm %>% 
  filter(Pathways %in% metacyc_keep) %>% 
  mutate(name_fix = str_remove_all(`Common-Name`, "\\<sub\\>"),
         name_fix = str_remove_all(name_fix, "\\<\\/sub\\>"),
         name_fix = str_remove_all(name_fix, "\\<\\i\\>"),
         name_fix = str_remove_all(name_fix, "\\<\\/i\\>")) %>% 
   mutate(pathway_category = case_when(

    str_detect(name_fix,
               "glycolysis|pentose phosphate|glyoxylate|glycolate|ethylmalonyl|TCA cycle|pyruvate") ~ 
      "Carbon metabolism",

    str_detect(name_fix,
               "Calvin|reductive TCA|reductive glycine") ~ 
      "Carbon fixation",

    str_detect(name_fix,
               "ammonia oxidation|nitrate reduction|ammonia assimilation") ~ 
      "Nitrogen metabolism",

    str_detect(name_fix,
               "sulfate reduction") ~ 
      "Sulfur metabolism",

    str_detect(name_fix,
               "formate oxidation|carbon monoxide oxidation|glycine cleavage") ~ 
      "C1 metabolism",

    str_detect(name_fix,
               "PRPP biosynthesis|inositol biosynthesis") ~ 
      "Biosynthesis",

    TRUE ~ "Other"
  ))

```


## Line plot - Community Level

```{r}

stages <- tibble(
  stage = c('Pre monsoon', 'During monsoon', 'Post monsoon', 'Pre monsoon'),
  start = c(-Inf, 1.5, 5.5, 6.5),
  end = c(1.5, 5.5, 6.5, Inf)
)


```

## Figure for Carbon

```{r}

rx_line_C <- kegg_ready %>%
  inner_join(mags_ft, by = c('ID', 'bin')) %>% ## Filtering MAGs pathway > 50
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_metaT')) %>%
  group_by(ID, SampleID, Month) %>%
  summarise(expression = sum(expression)) %>%
  left_join(def, by = 'ID') %>%
  group_by(Category_2, Category_1, Month) %>%
  summarise(expression = mean(expression)) %>%
  filter(Category_1 %in% category_1_keep) %>% 
  mutate(Category_1 = ifelse(Category_1 == 'Glycolysis (Embden-Meyerhof pathway)',
                             'Glycolysis', Category_1),
         Category_1 = ifelse(Category_1 == 'Glycolysis, core module involving three-carbon compounds', 'Glycolysis 3 C', Category_1),
         Category_1 = ifelse(Category_1 == 'Pentose P pathway, oxidative phase',
                             'PPP_ox', Category_1),
         Category_1 = ifelse(Category_1 == 'Pentose P pathway, non-oxidative phase',
                             'PPP_non-ox', Category_1)) %>% 
  filter(expression > 0) %>%
  filter(Category_2 == 'Central carbon metabolism',
         Category_1 != 'Glycolysis 3 C') %>% 
  mutate(db = 'KEGG')


## METACYC

meta_C <- meta_comm_ft %>% 
  filter(pathway_category == 'Carbon metabolism') %>% 
  filter(!name_fix %in% c('glyoxylate cycle', 'glycolate and glyoxylate degradation II',
                         'ethylmalonyl-CoA pathway', 'incomplete reductive TCA cycle',
                         'reductive TCA cycle I')) %>% 
  mutate(name_fix = ifelse(name_fix == 'TCA cycle I (prokaryotic)',
                             'TCA', name_fix),
         name_fix = ifelse(name_fix == 'pyruvate decarboxylation to acetyl CoA I', 'Pyruvate oxidation', name_fix),
                  name_fix = ifelse(name_fix == 'pentose phosphate pathway (non-oxidative branch) I',
                             'PPP_non-oxI', name_fix),
         name_fix = ifelse(name_fix == 'pentose phosphate pathway (non-oxidative branch) II',
                             'PPP_non-oxII', name_fix),
         name_fix = ifelse(name_fix == 'pentose phosphate pathway (oxidative branch) I',
                             'PPP_ox', name_fix)) %>%
  mutate(db = 'MetaCyc') %>% 
  select(pathway_category, name_fix, month, path_expr_month, db) %>% 
  rename(Category_2 = pathway_category,
         Category_1 = name_fix,
         Month = month,
         expression = path_expr_month) %>% 
  filter(Category_1 %in% c('glycolysis III (from glucose)',
                           'PPP_ox',
                           'PPP_non-oxI',
                           'Pyruvate oxidation',
                           'TCA')) %>% 
  mutate(Category_1 = case_when(Category_1 == 'glycolysis III (from glucose)' ~ 'Glycolysis',
                                Category_1 == 'PPP_non-oxI' ~ 'PPP_non-ox',
                                TRUE ~ Category_1))


## Joining both

join_C <- rbind(rx_line_C, meta_C) %>%
  group_by(db, Category_1) %>%
  mutate(zscore = (expression - mean(expression))/sd(expression))


## Set colors

modules_colors_C <- c(
  'Glycolysis' = "#E64B35",               
  'PPP_non-ox' = "#6A3D9A",  
  'PPP_ox' = "#00A087",     
  'Pyruvate oxidation' = "#3C5488",       
  'TCA' = "#7E6148"                        
)




## Plot

line_c_plot <- join_C %>%
  mutate(Month = factor(Month, levels = c("May", "July_1", "July_2", "July_3",
                                          "August", "October", "May_22")))  %>% 
  ggplot() +
  geom_rect(data = stages,
            aes(xmin = start,
                xmax = end,
                ymin = -Inf,
                ymax = Inf,
                fill = stage),
            alpha = .15,
            show.legend = FALSE) +
  geom_line(aes(x = Month, y = log10(expression),
                color = Category_1,
                group = interaction(Category_1, db),
                linetype = db)) +
  geom_point(aes(x = Month, y = log10(expression),
                 color = Category_1,
                 group = Category_1),
             size = 1.5) +
  geom_label_repel(aes(x = Month, 
                       y = log10(expression),
                       label = Category_1,
                       color = Category_1),
                   data = join_C %>% filter(Month == 'May_22',
                                            db == 'KEGG'),
                   nudge_x = 1, 
                   nudge_y = .1,
                   size = 2,
                   show.legend = FALSE,
                   segment.color = "grey60",
                   box.padding = 0.3)+
  scale_fill_manual(values = c('Pre monsoon' = '#F2F2F2', 
                               'During monsoon' = '#7BBBE5', 
                               'Post monsoon' = '#E8F0FA')) +
  scale_color_manual(values = modules_colors_C)+
  scale_y_continuous(limits = c(NA, NA)) +
  labs(title = 'Carbon metabolism',
       y = 'log10(geTMM)')+
  guides(color = guide_legend(ncol =2)) +
  theme_bw(base_size = 9, base_family = "sans") +
  # facet_wrap(~db,
  #            scales = 'free_y') +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", linewidth = 0.4),
    plot.title = element_text(size = 7, face = "bold", hjust = 0.5),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 6, face = "bold"),
    axis.text = element_text(color = "black", size = 6),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.ticks = element_line(color = "black", linewidth = 0.3),
    legend.position = "none",
    plot.margin = margin(5, 5, 5, 5)
  )

line_c_plot

```


## Figures - Nitrogen

```{r}

rx_line_N <- kegg_ready %>%
  inner_join(mags_ft, by = c('ID', 'bin')) %>% ## Filtering MAGs pathway > 50
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_metaT')) %>%
  group_by(ID, SampleID, Month) %>%
  summarise(expression = sum(expression)) %>%
  left_join(def, by = 'ID') %>%
  group_by(Category_2, Category_1, Month) %>%
  summarise(expression = mean(expression)) %>%
  filter(Category_1 %in% category_1_keep) %>% 
  mutate(Category_1 = ifelse(Category_1 == 'Assimilatory nitrate reduction',
                             'ANR', Category_1),
         Category_1 = ifelse(Category_1 == 'Dissimilatory nitrate reduction', 
                             'DNR', Category_1)) %>% 
  filter(expression > 0) %>%
  filter(Category_2 == 'Nitrogen metabolism') %>% 
  mutate(db = "KEGG")

## Metacyc


meta_N <- meta_comm_ft %>% 
  filter(pathway_category == 'Nitrogen metabolism',
         name_fix != "ammonia assimilation cycle II") %>%
  mutate(name_fix = ifelse(name_fix == 'ammonia oxidation I (nitrification)',
                             'Nitrification', name_fix),
         name_fix = ifelse(name_fix == 'nitrate reduction IX (dissimilatory)', 'DNR', name_fix),
                  name_fix = ifelse(name_fix == 'nitrate reduction VI (assimilatory)',
                             'ANR', name_fix)) %>%
  mutate(db = 'MetaCyc') %>% 
  select(pathway_category, name_fix, month, path_expr_month, db) %>% 
  rename(Category_2 = pathway_category,
         Category_1 = name_fix,
         Month = month,
         expression = path_expr_month)


join_N <- rbind(rx_line_N, meta_N) %>%
  group_by(db, Category_1) %>%
  mutate(zscore = (expression - mean(expression))/sd(expression))

### Plot Nitrogen


modules_colors_N <- c('ANR' = 'darkblue',
                      'DNR' = "#6A3D9A",
                      'Denitrification' = "#7E6148"  ,
                      'Nitrification' = '#E64B35')

## Plot

line_N_plot <- join_N %>%
  mutate(Month = factor(Month, levels = c("May", "July_1", "July_2", "July_3",
                                          "August", "October", "May_22")))  %>% 
  ggplot() +
  geom_rect(data = stages,
            aes(xmin = start,
                xmax = end,
                ymin = -Inf,
                ymax = Inf,
                fill = stage),
            alpha = .15,
            show.legend = FALSE) +
  geom_line(aes(x = Month, y = log10(expression),
                color = Category_1,
                group = interaction(Category_1, db),
                linetype = db)) +
  geom_point(aes(x = Month, y = log10(expression),
               color = Category_1,
               group = Category_1),
           size = 1.5) +
  geom_label_repel(aes(x = Month, 
                       y = log10(expression),
                       label = Category_1,
                       color = Category_1),
                   data = join_N %>% filter(Month == 'May_22',
                                            db == 'KEGG'),
                   nudge_x = 1, 
                   nudge_y = .1,
                   size = 2,
                   show.legend = FALSE,
                   segment.color = "grey60",
                   box.padding = 0.3)+
  scale_fill_manual(values = c('Pre monsoon' = '#F2F2F2', 
                               'During monsoon' = '#7BBBE5', 
                               'Post monsoon' = '#E8F0FA')) +
  scale_color_manual(values = modules_colors_N)+
  scale_y_continuous(limits = c(NA, NA)) +
  labs(title = 'Nitrogen metabolism',
       y = 'log10(geTMM)')+
  guides(color = guide_legend(ncol =2)) +
  theme_bw(base_size = 9, base_family = "sans") +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", linewidth = 0.4),
    plot.title = element_text(size = 7, face = "bold", hjust = 0.5),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 6, face = "bold"),
    axis.text = element_text(color = "black", size = 6),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.ticks = element_line(color = "black", linewidth = 0.3),
    legend.position = "none",
    plot.margin = margin(5, 5, 5, 5)
  )

line_N_plot

```

## Figures - Sulfur

```{r}

rx_line_S <- kegg_ready %>%
  inner_join(mags_ft, by = c('ID', 'bin')) %>% ## Filtering MAGs pathway > 50
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_metaT')) %>%
  group_by(ID, SampleID, Month) %>%
  summarise(expression = sum(expression)) %>%
  left_join(def, by = 'ID') %>%
  group_by(Category_2, Category_1, Month) %>%
  summarise(expression = mean(expression)) %>%
  filter(Category_1 %in% category_1_keep) %>% 
  mutate(Category_1 = ifelse(Category_1 == 'Assimilatory sulfate reduction',
                             'ASR', Category_1),
         Category_1 = ifelse(Category_1 == 'Dissimilatory sulfate reduction', 
                             'DSR', Category_1)) %>% 
  filter(expression > 0) %>%
  filter(Category_2 == 'Sulfur metabolism') %>% 
  mutate(db = 'KEGG')

## MetaCyc


meta_S <- meta_comm_ft %>% 
  filter(pathway_category == 'Sulfur metabolism') %>% 
  mutate(db = 'MetaCyc') %>% 
  select(pathway_category, name_fix, month, path_expr_month, db) %>% 
  rename(Category_2 = pathway_category,
         Category_1 = name_fix,
         Month = month,
         expression = path_expr_month) %>% 
  mutate(Category_1 = case_when(Category_1 == 'assimilatory sulfate reduction IV' ~ 'ASR IV',
                                Category_1 == 'assimilatory sulfate reduction I' ~ 'ASR I',
                                TRUE ~ Category_1))


join_S <- rbind(rx_line_S, meta_S) %>%
  group_by(db, Category_1) %>%
  mutate(zscore = (expression - mean(expression))/sd(expression))


### Plot Sulfur


modules_colors_S <- c('DSR' = '#219ebc',
                      'ASR' = '#023047',
                      'ASR IV' = '#ffb703',
                      'ASR I' = '#fb8500')

## Plot

line_S_plot <- join_S %>%
    mutate(Month = factor(Month, levels = c("May", "July_1", "July_2", "July_3",
                                          "August", "October", "May_22")))  %>% 
  ggplot() +
  geom_rect(data = stages,
            aes(xmin = start,
                xmax = end,
                ymin = -Inf,
                ymax = Inf,
                fill = stage),
            alpha = .15,
            show.legend = FALSE) +
  geom_line(aes(x = Month, y = log10(expression),
                color = Category_1,
                group = interaction(Category_1, db),
                linetype = db)) +
  geom_point(aes(x = Month, y = log10(expression),
               color = Category_1,
               group = Category_1),
           size = 1.5) +
  geom_label_repel(aes(x = Month, 
                       y = log10(expression),
                       label = Category_1,
                       color = Category_1),
                   data = join_S %>% filter(Month == 'May_22'),
                   nudge_x = 1, 
                   nudge_y = .1,
                   size = 2,
                   show.legend = FALSE,
                   segment.color = "grey60",
                   box.padding = 0.3)+
  scale_fill_manual(values = c('Pre monsoon' = '#F2F2F2', 
                               'During monsoon' = '#7BBBE5', 
                               'Post monsoon' = '#E8F0FA')) +
  scale_color_manual(values = modules_colors_S)+
  scale_y_continuous(limits = c(NA, NA)) +
  labs(title = 'Sulfur metabolism',
       y = 'log10(geTMM)')+
  guides(color = guide_legend(ncol =2)) +
  theme_bw(base_size = 9, base_family = "sans") +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", linewidth = 0.4),
    plot.title = element_text(size = 7, face = "bold", hjust = 0.5),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 6, face = "bold"),
    axis.text = element_text(color = "black", size = 6),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.ticks = element_line(color = "black", linewidth = 0.3),
    legend.position = "none",
    plot.margin = margin(5, 5, 5, 5)
  )


line_S_plot


```


## Line plot -- Class level -- Supplementary


```{r}


rx_line_class <- kegg_ready %>%
  inner_join(mags_ft, by = c('ID', 'bin')) %>% ## Filtering MAGs pathway > 50
  left_join(metadata_ready, by = c('SampleID' = 'SampleID_metaT')) %>%
  left_join(mag_taxonomy, by = 'bin') %>%
  group_by(ID, SampleID, Month, Class) %>%
  summarise(expression = sum(expression)) %>%
  left_join(def, by = 'ID') %>%
  group_by(Category_2, Category_1, Month, Class) %>%
  summarise(expression = mean(expression)) %>% 
  filter(Category_1 %in% category_1_keep) %>% 
  mutate(Category_1 = ifelse(Category_1 == 'Glycolysis (Embden-Meyerhof pathway)',
                             'Glycolysis', Category_1),
         Category_1 = ifelse(Category_1 == 'Glycolysis, core module involving three-carbon compounds', 'Glycolysis 3 carbon', Category_1),
         Category_1 = ifelse(Category_1 == 'Pentose P pathway, oxidative phase',
                             'Pentose P, oxidative phase', Category_1),
         Category_1 = ifelse(Category_1 == 'Pentose P pathway, non-oxidative phase',
                             'Pentose P, non-oxidative phase', Category_1)) %>% 
  filter(expression > 0) 


```

## Filter top 20% expressed Class per module

```{r}

top_expressed_class <- rx_line_class %>%
  ungroup() %>% 
  group_by(Category_1, Class) %>%
  summarise(mean_exp = mean(expression),
            total_exp = sum(expression)) %>% 
  group_by(Category_1) %>% 
  slice_max(order_by = mean_exp, prop = 0.2) %>% 
  pull(Class) %>% 
  unique()

## Prepare for map

rx_line_class_2 <- rx_line_class %>% 
  filter(Class %in% top_expressed_class) %>% 
  split(.$Category_1)

```


## Line plot - Class Level - Supplementary

```{r}


## Set color CLASS

class_mags <-reduce(rx_line_class_2, rbind) %>% 
  ungroup() %>% 
  select(Class) %>% 
  distinct() %>% 
  pull()


class_color <- set_names(ggpubr::get_palette('npg', 8),
                         nm = sort(class_mags))


line_map_class <- imap(rx_line_class_2, function(cat, name) {
  
  line_nito_plot <- cat %>%
    ggplot() +
    geom_rect(data = stages,
              aes(xmin = start,
                  xmax = end,
                  ymin = -Inf,
                  ymax = Inf,
                  fill = stage),
              alpha = .3,
              show.legend = FALSE) +
    geom_line(aes(x = Month, y = expression,
                  color = Class,
                  group = Class),
              linewidth = 0.7) +
    geom_point(aes(x = Month, y = expression,
                   color = Class,
                   group = Class),
               size = 1) +
    scale_fill_manual(values = c('Pre monsoon' = '#F2F2F2', 
                                 'During monsoon' = '#7BBBE5', 
                                 'Post monsoon' = '#E8F0FA')) +
    scale_color_manual(values = class_color)+
    scale_y_continuous(limits = c(0, NA)) +
    labs(y = 'geTMM',
         color = 'Class',
         title = name)+
    #guides(color = guide_legend(ncol =2)) +
    theme_bw() +
    theme(
      panel.grid = element_blank(),                        # remove grid lines
      plot.title = element_text(size = 6, face = "bold", hjust = 0.5),  # centered, small title
      axis.title.x = element_blank(),                     # no x-axis title
      axis.title.y = element_text(size = 7, face = "bold"),
      axis.text = element_text(size = 6),                # smaller axis text
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.ticks.length = unit(0.15, "cm"),              # shorter ticks
      legend.title = element_text(size = 6, face = "bold"),
      legend.text = element_text(size = 6),
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.box = "horizontal",
      legend.key.size = unit(0.35, "lines"),             # smaller legend keys
      legend.title.position = "top",
      plot.margin = margin(2, 2, 2, 2)                  # smaller margins around plot
    )
  
  return(line_nito_plot)
  
  
})


for_legend <- data.frame(Class = class_mags) %>% 
  ggplot(aes(x = 1, y = Class, color = Class)) +
  geom_point(size = 2) +  # smaller points to save space
  scale_color_manual(values = class_color) +
  theme_bw(base_size = 7) + 
  theme(
    panel.grid = element_blank(),                  # remove grid
    panel.border = element_blank(),                # optional: remove border
    plot.title = element_text(size = 7, face = "bold", hjust = 0.5),
    axis.title = element_blank(),                  # remove axis titles
    axis.text = element_blank(),                   # remove axis text
    axis.ticks = element_blank(),                  # remove ticks
    legend.title = element_text(size = 7, face = "bold"),
    legend.text = element_text(size = 6),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "horizontal",
    legend.key.size = unit(0.35, "lines"),
    legend.spacing.x = unit(0.2, "cm"),
    legend.title.position = "top",
    plot.margin = margin(1, 1, 1, 1)             # minimal margins
  ) +
  guides(
    color = guide_legend(nrow = 1, byrow = TRUE)
  )


legend <- ggpubr::get_legend(for_legend)
```


```{r}

keep_plot <- c(7, 8, 9, 11, 12, 10, 1, 4, 5, 2)



line_map_filtered <- line_map_class[keep_plot]

des <- "
ABCDE
FGHIJ
KKKKK
"

plot_filt <- reduce(line_map_filtered, `+`) +
  legend +
  plot_layout(design = des,
              heights = c(1, 1, .1)) &
  theme(legend.position = 'none')

plot_filt <- plot_filt +
  plot_annotation(title = "Metatranscriptomics")&
  theme(plot.title = element_text(face = 'bold', hjust = 0.5))

plot_filt

```


## Save final plot


```{r}


# ggsave("output_figures/Supp_Figure_5.svg", plot_filt, dpi = 300,
#        units = 'mm', 
#        height = 160, 
#        width = 180)

ggsave("output_figures/Supp_Figure_5.png", plot_filt, dpi = 300,
       units = 'mm', 
       height = 160, 
       width = 180)



```


## Other pathways

```{r}

keep_plot_2 <- c(13, 16)

line_map_filtered <- line_map_class[keep_plot_2]

des <- "
AABB
KKKK
"

plot_filt <- reduce(line_map_filtered, `+`) + legend +
  plot_layout(design = des,
              heights = c(1,.1)) &
  theme(legend.position = 'none')


plot_filt

plot_filt <- plot_filt +
  plot_annotation(title = "Metatranscriptomics")&
  theme(plot.title = element_text(face = 'bold', hjust = 0.5))

plot_filt

```



```{r}

# ggsave("output_figures/Supp_Figure_6.svg", plot_filt, dpi = 300,
#        units = 'mm', 
#        height = 100, 
#        width = 180)

ggsave("output_figures/Supp_Figure_6.png", plot_filt, dpi = 300,
       units = 'mm', 
       height = 100, 
       width = 180)


```


## Expression at CLass level -- key genes

## Expression key genes - Heatmap

```{r}

key_genes <- read_csv("input/Key_genes_3_18_25.csv")


```


```{r}

annot_class_ko <- geTMM_long %>% 
  left_join(annot_cydbs, by = 'Geneid') %>%
  filter(!is.na(ko_id)) %>% 
  left_join(plot_groups, by = c('fasta' = 'genome')) %>% 
  left_join(mag_taxonomy, by = c('fasta' = 'bin')) %>% 
  group_by(sampleid, Class, response_type, ko_id) %>% 
  summarise(getmm_sum = sum(geTMM)) %>%
  mutate(sampleid = str_replace_all(sampleid, '-', '_')) %>% 
  left_join(metadata_ready, by = c('sampleid' = 'treatment_code')) %>%
  group_by(Class, ko_id, response_type, Month) %>% 
  summarise(getmm_mean = mean(getmm_sum)) %>% 
  filter(getmm_mean > 0) 


```


```{r}

annot_class_vip <- annot_class_ko %>% 
  inner_join(key_genes, by = 'ko_id')

## Plot only top expressed

top_kos <- annot_class_vip %>%
  group_by(Class, Month, response_type) %>%
  slice_max(order_by = getmm_mean, prop = 0.2) %>%
  select(ko_id) %>%
  pull() %>%
  unique()

supp_ko <- c("Gluconeogenesis", "EMC", "Glyoxylate bypass", "Glycine cleavage",
             "Arginine biosynthesis", "methionine sulfoxide reductases", "Other")

ko_vip_heatmap <- annot_class_vip %>% 
  filter(ko_id %in% top_kos,
         !Category_2 %in% supp_ko) %>%
  mutate(Category_2 = case_when(str_detect(Category_2, 'Carbon fixation') ~'C fixation',
                                str_detect(Category_2, 'Pentose phosphate pathway') ~'PPP',
                                str_detect(Category_2, 'Nitrogen metabolism') ~'N metabolism',
                                str_detect(Category_2, 'Sulfur metabolism') ~'S metabolism',
                                str_detect(Category_2, 'Phosphorous metabolism') ~'P metabolism',
                                .default = Category_2),
         Category_2 = str_replace(Category_2, ' ', '\n')) %>% 
  mutate(Category_2 = factor(Category_2, 
                             levels = c('C\nfixation',
                                        'TCA',
                                        'PPP',
                                        'Glycolisis',
                                        'Piruvate\noxidation',
                                        'N\nmetabolism',
                                        'S\nmetabolism',
                                        'P\nmetabolism'))) 


top_class <- ko_vip_heatmap %>% 
  filter(response_type != 'Pulse') %>% 
  group_by(ko_id, response_type, Class) %>% 
  summarise(mean_exp = mean(getmm_mean)) %>% 
  slice_max(order_by = mean_exp, prop = 0.1) %>% 
  pull(Class) %>% 
  unique()



class_vip_ko_vip <- ko_vip_heatmap %>% 
  filter(Class %in% top_class) %>% 
  group_by(gene_abb, Category_2, Class, response_type) %>%
  mutate(zscore = (getmm_mean - mean(getmm_mean)) / sd(getmm_mean))


```

## Heatmap key genes

```{r}

htmp_plot <-  class_vip_ko_vip %>% 
  filter(response_type != 'Pulse',
         !is.na(Category_2)) %>% 
  ggplot() +
  geom_tile(aes(x = Month,
                y = gene_abb,
                fill = zscore)) + 
  # scale_fill_viridis_c(option = 'A', direction = -1) +
  scale_fill_distiller(palette = 'RdBu', direction = -1) +
  facet_grid2(cols = vars(response_type, Class),
              rows = vars(Category_2),
              switch = 'y',
              scales = 'free',
              space = 'free',
              strip = strip_nested(
                text_x = elem_list_text(angle = c(0, 0, rep(0, 100))),
                text_y = elem_list_text(angle = c(rep(90, 8), rep(0, 100)),
                                        face = c(rep('bold', 8), 
                                                 rep('plain', 100))),
                size = 'variable',
                clip = 'off'
              )) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.title = element_blank(),
        strip.background = element_blank(),
        strip.text = element_text(size = 4, margin = margin(b = 0, t = 0)),
        strip.text.y.left = element_text(angle = 0, margin = margin(r = 0, l = 0)),
        strip.placement = "outside",      # closer to panels
        panel.spacing = unit(0.1, "lines"),
        axis.text.y = element_text(size = 5),
        axis.text.x = element_text(size = 5, angle = 45, hjust = 1),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 5),
        legend.position = 'bottom',
        legend.key.size = unit(.5, 'lines'),
        legend.spacing.x = unit(0.1, "cm"),
        plot.margin = margin(0.1, 0.1, 0.1, 0.1, 'lines'))

htmp_plot

```

## Stitching figure

```{r}

legend_df <- data.frame(
  x = c(1,1),
  xend = c(2,2),
  y = c(1,2),
  database = c("KEGG","MetaCyc")
)


df_long <- ggplot(legend_df) +
  geom_segment(aes(x = x, xend = xend, y = y, yend = y, linetype = database),
               linewidth = 1.2) +
  scale_linetype_manual(
    name = "Database",
    values = c("KEGG" = "solid",
               "MetaCyc" = "dashed")
  ) +
  theme_classic()+
  theme(legend.position = 'bottom',
        panel.background = element_blank(),
        panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(linewidth = 0.3, colour = "black"),
        
        axis.title = element_text(size = 8),
        axis.text = element_text(size = 7),
        
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 7),
        
        legend.key = element_blank(),
        legend.key.width = unit(1.2, "cm"),  # keep line visible but compact
        legend.spacing.x = unit(0.2, "cm"),
        
        plot.margin = margin(2,2,2,2))

df_long

leg_long <- get_legend(df_long)

## Joining

layout <- "
ABC
DDD
EEE
"

final_fig <- line_c_plot + line_N_plot + line_S_plot + leg_long + free(htmp_plot) +
  plot_layout(design = layout, heights = c(1, 0.1, 4)) +
  plot_annotation(tag_levels = list(c('A', 'B', 'C', '', 'D')))


final_fig


final_fig <- final_fig +
  plot_annotation(title = "Metatranscriptomics")&
  theme(plot.title = element_text(face = 'bold', hjust = 0.5))

final_fig

ggsave('output_figures/Figure_3.png', 
       dpi = 300,
       units = 'mm', 
       height = 200, 
       width = 180)
# ggsave('output_figures/Figure_3.svg', 
#        dpi = 300,
#        units = 'mm', 
#        height = 200, 
#        width = 180)
```

## Other pathways - Supplementary heatmap

```{r}

ko_vip_heatmap_2 <- annot_class_vip %>% 
  filter(ko_id %in% top_kos,
         Category_2 %in% supp_ko)


top_class <- ko_vip_heatmap_2 %>% 
  filter(response_type != 'Pulse') %>% 
  group_by(ko_id, response_type, Class) %>% 
  summarise(mean_exp = mean(getmm_mean)) %>% 
  slice_max(order_by = mean_exp, prop = 0.2) %>% 
  pull(Class) %>% 
  unique()



class_vip_ko_vip <- ko_vip_heatmap_2 %>% 
  filter(Class %in% top_class) %>% 
  group_by(gene_abb, Category_2, Class, response_type) %>%
  mutate(zscore = (getmm_mean - mean(getmm_mean)) / sd(getmm_mean))

```

```{r}


supp_htmp_plot <-  class_vip_ko_vip %>% 
  filter(response_type != 'Pulse',
         !is.na(Category_2)) %>% 
  ggplot() +
  geom_tile(aes(x = Month,
                y = gene_abb,
                fill = zscore)) + 
  # scale_fill_viridis_c(option = 'A', direction = -1) +
  scale_fill_distiller(palette = 'RdBu', direction = -1) +
  facet_grid2(cols = vars(response_type, Class),
              rows = vars(Category_2),
              switch = 'y',
              scales = 'free',
              space = 'free',
              strip = strip_nested(
                text_x = elem_list_text(angle = c(0, 0, rep(0, 100))),
                text_y = elem_list_text(angle = c(rep(90, 8), rep(0, 100)),
                                        face = c(rep('bold', 8), 
                                                 rep('plain', 100))),
                size = 'variable',
                clip = 'off'
              )) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.title = element_blank(),
        strip.background = element_blank(),
        strip.text = element_text(size = 4, margin = margin(b = 0, t = 0)),
        strip.text.y.left = element_text(angle = 0, margin = margin(r = 0, l = 0)),
        strip.placement = "outside",      # closer to panels
        panel.spacing = unit(0.1, "lines"),
        axis.text.y = element_text(size = 5),
        axis.text.x = element_text(size = 5, angle = 45, hjust = 1),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 5),
        legend.position = 'bottom',
        legend.key.size = unit(.5, 'lines'),
        legend.spacing.x = unit(0.1, "cm"),
        plot.margin = margin(0.1, 0.1, 0.1, 0.1, 'lines'))

supp_htmp_plot

supp_htmp_plot <- supp_htmp_plot +
  plot_annotation(title = "Metatranscriptomics")&
  theme(plot.title = element_text(face = 'bold', hjust = 0.5))

supp_htmp_plot


ggsave('output_figures/Supp_Figure_8.png', 
       dpi = 300,
       units = 'mm', 
       height = 200, 
       width = 180)

```


## Expression at MAG level -- key genes

## Expression key genes - Heatmap


```{r}

annot_MAG_ko <- geTMM_long %>% 
  left_join(annot_cydbs, by = 'Geneid') %>%
  filter(!is.na(ko_id)) %>% 
  left_join(plot_groups, by = c('fasta' = 'genome')) %>% 
  group_by(sampleid, fasta, response_type, ko_id) %>% 
  summarise(getmm_sum = sum(geTMM)) %>%
  mutate(sampleid = str_replace_all(sampleid, '-', '_')) %>% 
  left_join(metadata_ready, by = c('sampleid' = 'treatment_code')) %>%
  group_by(fasta, ko_id, response_type, Month) %>% 
  summarise(getmm_mean = mean(getmm_sum)) %>% 
  left_join(mag_taxonomy, by = c('fasta' = 'bin')) %>% 
  filter(getmm_mean > 0) 


```


```{r}

annot_MAG_vip <- annot_MAG_ko %>% 
  inner_join(key_genes, by = 'ko_id')

write_csv(annot_MAG_vip, 'output_tables/Supplementary_data_3.csv')

## Plot only top expressed

top_kos <- annot_MAG_vip %>%
  group_by(fasta, Month, response_type) %>%
  slice_max(order_by = getmm_mean, prop = 0.1) %>% 
  select(ko_id) %>% 
  pull() %>% 
  unique()

supp_ko <- c("Gluconeogenesis", "EMC", "Glyoxylate bypass", "Glycine cleavage",
             "Arginine biosynthesis", "methionine sulfoxide reductases", "Other")

ko_vip_heatmap <- annot_MAG_vip %>% 
  filter(ko_id %in% top_kos,
         !Category_2 %in% supp_ko) %>%
  mutate(Category_2 = case_when(str_detect(Category_2, 'Carbon fixation') ~'C fixation',
                                str_detect(Category_2, 'Pentose phosphate pathway') ~'PPP',
                                str_detect(Category_2, 'Nitrogen metabolism') ~'N metabolism',
                                str_detect(Category_2, 'Sulfur metabolism') ~'S metabolism',
                                str_detect(Category_2, 'Phosphorous metabolism') ~'P metabolism',
                                .default = Category_2),
         Category_2 = str_replace(Category_2, ' ', '\n')) %>% 
  mutate(Category_2 = factor(Category_2, 
                             levels = c('C\nfixation',
                                        'TCA',
                                        'PPP',
                                        'Glycolisis',
                                        'Piruvate\noxidation',
                                        'N\nmetabolism',
                                        'S\nmetabolism',
                                        'P\nmetabolism'))) 

mag_vip_ko <- ko_vip_heatmap %>% 
  group_by(gene_abb, Category_2, fasta, response_type) %>%
  mutate(zscore = (getmm_mean - mean(getmm_mean)) / sd(getmm_mean)) %>% 
  split(.$Category_2)


```


