---
title: "Supplementary figure 1"
author: "Viviana Freire Zapata"
date: "12/3/2025"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: '2'
    collapsed: false
    highlight: tango
    css: style.css
editor_options: 
  chunk_output_type: console
---

# Data Pre-processing



# Importing libraries

```{r libraries, message=FALSE, warning=FALSE}

library(vegan)
library(matrixStats)
library(GGally)
library(ggpubr)
library(performance)
library(ggtext)
library(ggforce)
library(grid)
library(ggplotify)
library(ggh4x)
library(patchwork)
library(tidyverse)

source('figure_scripts/functions_cdis_norm_stats.R')
source('figure_scripts/functions_cdis_exploration.R')

```

# Loading data 

# Loading data RP

```{r}
## Raw matrix


RP_df <- read_csv("scripts/LC-MS-MS/1.Pre-processing/output/RP_matrix_clean_2025.csv")


## Loading annotation

annot_rp <- read_csv("scripts/LC-MS-MS/3.Cleanning_Annotation/output/RP_final_annotation_11_20_25.csv")

```


# Loading data HILIC

```{r}
## Raw matrix


HILIC_df <- read_csv("scripts/LC-MS-MS/1.Pre-processing/output/HILIC_matrix_clean_2025.csv")


## Loading annotation

annot_hilic <- read_csv("scripts/LC-MS-MS/3.Cleanning_Annotation/output/HILIC_final_annotation_11_20_25.csv")

```


## Loading significant features

```{r}

sig_features_rp <- read_csv("scripts/LC-MS-MS/5.Metabolite_model/output/RP_sig_features_model_fwd_selected_RP.csv")


sig_features_hilic <- read_csv("scripts/LC-MS-MS/5.Metabolite_model/output/HILIC_sig_features_model_fwd_selected.csv")


```


# Metadata

```{r}

## Metadata

metadata <- read_csv("input/metadata_omics.csv") %>% 
  filter(month != 'May_22')


```


## Prepare matrix RP

```{r}

RP_ready <- RP_df %>% 
  select(FeatureID, contains("RP")) %>% 
  filter(FeatureID %in% annot_rp$FeatureID) %>% 
  column_to_rownames(var = "FeatureID")

```


# Metabolite Normalization

## Median normalization

```{r}

norm_matrix_rp <- median.norm(RP_ready, transform_data = TRUE)

```


## Filtering full matrix 

```{r}

norm_matrix_sig_rp <- norm_matrix_rp %>% 
  rownames_to_column(var = 'FeatureID') %>% 
  filter(FeatureID %in% sig_features_rp$FeatureID) %>% 
  pivot_longer(!FeatureID, names_to = 'SampleID', values_to = 'norm_int')

```


## Prepare matrix HILIC

```{r}

hilic_ready <- HILIC_df %>% 
  select(FeatureID, contains("HILIC")) %>% 
  filter(FeatureID %in% annot_hilic$FeatureID) %>% 
  column_to_rownames(var = "FeatureID")

```


# Metabolite Normalization

## Median normalization

```{r}

norm_matrix_hilic <- median.norm(hilic_ready, transform_data = TRUE)

```



## Filtering full matrix HILIC

```{r}

norm_matrix_sig_hilic <- norm_matrix_hilic %>% 
  rownames_to_column(var = 'FeatureID') %>% 
  filter(FeatureID %in% sig_features_hilic$FeatureID) %>% 
  pivot_longer(!FeatureID, names_to = 'SampleID', values_to = 'norm_int')


```

## Checking similar sig features between HILIC and RP


```{r}

combined_sig <- rbind(norm_matrix_sig_rp, norm_matrix_sig_hilic)


```


## Adding class info

## HILIC

```{r}

## Loading class of compounds

name_class_hilic <- read_csv("scripts/LC-MS-MS/4.Get_inchis/output/HILIC_classyfire_from_name.csv") %>% 
  filter(FeatureID %in% sig_features_hilic$FeatureID) %>% 
  select(FeatureID, Name, superclass, class, subclass, level_5 = `level 5`) %>% 
  mutate(LC_mode = 'HILIC')

canopus_class_hilic <- read_csv("scripts/LC-MS-MS/4.Get_inchis/output/HILIC_classyfire_canopus.csv") %>% 
  filter(FeatureID %in% sig_features_hilic$FeatureID) %>% 
  select(FeatureID, Name,
         superclass = `ClassyFire#superclass`,
         class = `ClassyFire#class`, 
         subclass = `ClassyFire#subclass`, 
         level_5 = `ClassyFire#level 5`) %>% 
  mutate(LC_mode = 'HILIC')

```

## RP

```{r}

name_class_rp <- read_csv("scripts/LC-MS-MS/4.Get_inchis/output/RP_classyfire_from_name_ready.csv") %>% 
  filter(FeatureID %in% norm_matrix_sig_rp$FeatureID) %>% 
  dplyr::select(FeatureID, Name, superclass, class, subclass, level_5 = `level 5`) %>% 
  mutate(LC_mode = 'RP')

canopus_class_rp <- read_csv("scripts/LC-MS-MS/4.Get_inchis/output/RP_classyfire_canopus_ready.csv") %>% 
  filter(FeatureID %in% norm_matrix_sig_rp$FeatureID) %>% 
  dplyr::select(FeatureID, Name,
                superclass = `ClassyFire#superclass`,
                class = `ClassyFire#class`, 
                subclass = `ClassyFire#subclass`, 
                level_5 = `ClassyFire#level 5`) %>% 
  mutate(LC_mode = 'RP')


```


## Combine class info in a single df

```{r}

class_info <- rbind(name_class_hilic, canopus_class_hilic, 
                    name_class_rp, canopus_class_rp)

```


## Joining CLASS info to df

```{r}

metadata_com <- metadata %>% 
  mutate(sampleid_com = str_remove(SampleID_RP, 'RP_'))

combined_sig_ready <- combined_sig %>% 
  mutate(sampleid_com = str_remove(SampleID, 'RP_|HILIC_')) %>% 
  left_join(metadata_com, by ='sampleid_com') %>% 
  left_join(class_info, by = c('FeatureID')) %>%
  mutate(Name = case_when(str_detect(Name, 'melitose') ~ 'D-Raffinose',
                          .default = Name),
         Name = case_when(str_detect(Name, 'alpha-alpha-Trehalose') ~ 'Trehalose',
                          .default = Name),
         Name = str_remove(Name, "\\(\\+\\)-")) %>% 
  group_by(FeatureID, Name, month) %>% 
  summarise(mean_int = mean(norm_int),
            sd = sd(norm_int))
  

```


## Heatmap

```{r}

combined_sig_split <- combined_sig_ready %>% 
  left_join(class_info, by = 'FeatureID')%>%
  select(!c(Name.y)) %>% 
  rename(Name = Name.x) %>% 
  split(.$superclass)

heat_vip <- imap(combined_sig_split, function(df, code_id){
  
  superclass <- df %>% pull(superclass) %>% unique
  class <- df %>% pull(class) %>% unique
  subclass <- df %>% pull(subclass) %>% unique
  level_5 <- df %>% pull(level_5) %>% unique
  
  class_tax <- str_flatten(c(superclass, class, subclass, level_5),
                           collapse = '; ', na.rm = TRUE)
  
  comp_name <- df %>% pull(Name) %>% unique
  
  plot <- df %>% 
    mutate(month = factor(month, levels = c('May', 'June', 'July_1', 'July_2',
                                            'July_3', 'August', 'September', 
                                            'October'))) %>% 
    group_by(Name, month, LC_mode) %>% 
    group_by(Name, LC_mode) %>% 
    mutate(zscore = (mean_int - mean(mean_int))/ sd(mean_int)) %>% 
    ungroup() %>% 
    ggplot(aes(x = month,
               y = Name,
               fill = zscore)) +
    geom_tile() +
    facet_grid2(rows = vars(LC_mode),
                space = 'free',
                scales = 'free')+
    scale_fill_distiller(palette = 'RdBu', 
                         #limits = c(-3, 3)
                         )+
    labs(x = 'Month',
         title = subclass)+
    theme_bw() +
    theme(plot.title = element_text(face = 'bold', hjust = 0.5, size = 5),
          axis.text.x = element_text(angle = 90, hjust = 1, size = 5),
          axis.text.y = element_text(size = 5),
          axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          strip.text.x = element_text(face = 'bold'),
          strip.text.y.right = element_text(angle = 0, size = 5),
          legend.position = 'bottom',
          strip.background = element_rect(fill = 'white'),
          panel.grid = element_blank(),
          strip.text.x.top = element_text(size = 5),
          strip.background.x = element_blank(),
          strip.background.y = element_blank())
  
  return(plot)
  
})



heat_vip$`Organic acids and derivatives`


```



```{r}

keep_plot <- c(2, 4, 6, 7, 8, 9)

heat_filtered <- heat_vip[keep_plot]

des <- "
ABC
DEF
"

plot_filt <- purrr::reduce(heat_filtered, `+`)+
  plot_layout(design = des
              ) &
  theme(legend.position = 'none')


plot_filt

```


## Save final plot


```{r}

ggsave("output_figures/Supp_Figure_6.svg", plot_filt, dpi = 300,
       units = 'mm',
       width = 300)

ggsave("output_figures/Supp_Figure_6.png", plot_filt, dpi = 300,
       units = 'mm',
       width = 300)



```


## All metabolites heatmap

```{r}


norm_matrix_rp_ln <- norm_matrix_rp %>% 
  rownames_to_column(var = 'FeatureID') %>% 
  pivot_longer(!FeatureID, names_to = 'SampleID', values_to = 'norm_int') %>% 
  left_join(annot_rp, by = 'FeatureID') %>% 
  mutate(LC_mode = 'RP')


norm_matrix_hilic_ln <- norm_matrix_hilic %>% 
  rownames_to_column(var = 'FeatureID') %>% 
  pivot_longer(!FeatureID, names_to = 'SampleID', values_to = 'norm_int') %>% 
  left_join(annot_hilic, by = 'FeatureID') %>% 
  mutate(LC_mode = 'HILIC')


combined_all <- rbind(norm_matrix_rp_ln, norm_matrix_hilic_ln)

## Prepare matrix

combined_all_ready <- combined_all %>% 
  filter(Annotation_level %in% c('L1', 'L2')) %>% 
  mutate(sampleid_com = str_remove(SampleID, 'RP_|HILIC_')) %>% 
  left_join(metadata_com, by ='sampleid_com') %>% 
  mutate(Name = case_when(str_detect(Name, 'melitose') ~ 'D-Raffinose',
                          .default = Name),
         Name = case_when(str_detect(Name, 'alpha-alpha-Trehalose') ~ 'Trehalose',
                          .default = Name),
         Name = str_remove(Name, "\\(\\+\\)-")) %>% 
  group_by(FeatureID, Name, month, LC_mode) %>% 
  summarise(mean_int = mean(norm_int)) %>% 
  group_by(FeatureID, Name, LC_mode) %>% 
  mutate(zscore = (mean_int - mean(mean_int))/ sd(mean_int)) %>% 
  ungroup()


## Heatmap


plot_all <- combined_all_ready %>% 
  filter(!str_detect(Name, 'Similar'),
         !str_detect(Name, 'NP-')) %>% 
  group_by(Name, FeatureID, LC_mode) %>%
  mutate(for_order = mean(zscore)) %>%
  ungroup() %>%
  arrange(for_order) %>%
  mutate(order = n():1) %>% 
  mutate(month = factor(month, levels = c('May', 'June', 'July_1', 'July_2',
                                          'July_3', 'August', 'September', 
                                          'October'))) %>% 
  ggplot(aes(x = month,
             y = fct_reorder(Name, order),
             fill = zscore)) +
  
  geom_tile() +
  facet_grid2(rows = vars(LC_mode),
              space = 'free',
              scales = 'free')+
  scale_fill_distiller(palette = 'RdBu', 
                       #limits = c(-3, 3)
  )+
  labs(x = 'Month')+
  theme_bw() +
  theme(plot.title = element_text(face = 'bold', hjust = 0.5, size = 6),
        axis.text.x = element_text(angle = 90, hjust = 1, size = 5),
        #axis.text.y = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        strip.text.x = element_text(face = 'bold'),
        strip.text.y.right = element_text(angle = 0, size = 5),
        legend.position = 'bottom',
        strip.background = element_rect(fill = 'white'),
        panel.grid = element_blank(),
        strip.text.x.top = element_text(size = 6),
        strip.background.x = element_blank(),
        strip.background.y = element_blank())

plot_all

```


# HILIC NMDS

# Feature filtering

```{r}
### Calculating IQR to keep only traits that are changing more
## Filtering metabolites and genes that doesn't change much
## Idea decrease the number of features to have a more balance matrices 

iqr_test_matrix <- rowIQRs(as.matrix(hilic_ready))

iqr_res <- as.data.frame(as.matrix(iqr_test_matrix))

# Filtering according to metaboAnalyts

# 1) Less than 250 variables: 5% will be filtered; 
# 2) Between 250 - 500 variables: 10% will be filtered; 
# 3) Between 500 - 1000 variables: 25% will be filtered; and 
# 4) Over 1000 variables: 40% will be filtered.

## 10% of the features will be filtered

# Sorting from largest to smallest (higher ranking = low iqr)
rk <- rank(-iqr_test_matrix, ties.method = 'random')

# Determining the number of remaining features
remain.num <- nrow(hilic_ready)*.9

remain <- rk <= remain.num

features_keep <- rownames(hilic_ready)[remain] 
  
## Filter matrix

matrix_ft <- hilic_ready[features_keep,]

## Note: 816 features kept for downstream analysis
```


# Metabolite Normalization

## Median normalization

```{r}

norm_matrix <- median.norm(matrix_ft, transform_data = TRUE)


```


## Creating colors 

```{r}

month_list <- metadata_com %>% 
  mutate(month = factor(month, levels = c('May', 'June', 'July_1', 'July_2',
                                          'July_3', 'August', 'September', 
                                          'October'))) %>% 
  select(month) %>% 
  distinct() %>% 
  pull()

my_colors_lc <- set_names(get_palette('Spectral', 12)[c(1:2, 9:12, 4, 5)],
                          nm = levels(month_list))

stage_list <- metadata %>% 
  mutate(stage = factor(stage, levels = c('Pre_monsoon', 'During_monsoon', 
                                          'Post_monsoon'))) %>%
  select(stage) %>% 
  distinct() %>% 
  pull()

stage_colors <- set_names(c('#FEDD7F', '#9bc1bc', '#ed6a5a'),
                          nm = levels(stage_list))

```


### NMDS - LC-MS

Calculating distance matrix

```{r}
t_norm_matrix <- t(norm_matrix)


dm <- vegdist(t_norm_matrix[metadata$SampleID_HILIC,], method = 'bray')



```


Performing NMDS analysis

#NMDS stress
# > 0.2 is deemed suspect
# approaching 0.3 is arbitrary
# = < 0.1 fair
# = < 0.5 good


```{r}
set.seed(123)
nmds <- metaMDS(dm,
                k = 2,
                maxit = 999,
                trymax = 500,
                wascores = TRUE)
stressplot(nmds)
```


Extracting NMDS scores and plotting

```{r}

nmds_scores <- as.data.frame(scores(nmds, display = 'sites')) %>%
  rownames_to_column(var = 'SampleID') %>% 
  left_join(metadata_com, by = c('SampleID' = 'SampleID_HILIC'))

nmds_scores$month <- factor(nmds_scores$month, levels = c("May", "June", "July_1", "July_2", "July_3", "August", "September", "October"))

nmds_scores$stage <- factor(nmds_scores$stage, levels = c("Pre_monsoon", "During_monsoon", "Post_monsoon"))


```


```{r}
## Testing PERMANOVA

# distance matrix by Bray because relative abundance mode was selected

dm_all <- vegdist(t_norm_matrix[metadata_com$SampleID_HILIC,], method = "bray")

set.seed(456)

permanova <- adonis2(dm_all ~ month,
                     data=metadata_com, 
                     permutations=999,
                     strata = metadata_com$site,
                     method="bray",
                     by = 'terms')

permanova

```

## get Centroid

```{r}

centroid <- nmds_scores %>% 
  group_by(stage) %>% 
  summarise(axis1 = mean(NMDS1),
              axis2 = mean(NMDS2)) %>% 
  ungroup()

```

## extract convex hull

```{r}

stage.hull <- nmds_scores %>% 
    group_by(stage) %>%
    slice(chull(NMDS1, NMDS2))

```

```{r}

nmds_plot_hilic <- nmds_scores %>% 
  ggplot(aes(x = NMDS1,
             y = NMDS2))+
  # geom_mark_ellipse(aes(fill = stage),
  #                   alpha = 0.05,
  #                   show.legend = FALSE) +
  #scale_fill_manual(values = c('firebrick4', 'steelblue', 'orange')) +
  geom_point(aes(color = month),
             size = 3) +
  geom_point(data = centroid, 
             aes(x = axis1, y = axis2, 
                 color = stage), 
             size = 5, shape = 17)+
  geom_polygon(data = stage.hull, 
               aes(x = NMDS1, y = NMDS2, 
                   fill = stage, 
                   group = stage), 
               alpha = 0.30) +
  scale_color_manual(values = my_colors_lc)+
  scale_fill_manual(values = stage_colors)+
  annotate("text", x = -0.013, y = 0.007,
           label = paste(c('Stress: ', round(nmds$stress, 4), "\n",
                          'Month pvalue: ', 
                          round(permanova$`Pr(>F)`[1], 4),"\n",
                          'Month R²: ', round(permanova$R2[1], 2)),
                         collapse = ''),
           size = 2)+
  labs(fill = "Monsoon stage", shape = "Monsoon stage",
       title = "Metabolomics - HILIC mode")+
  theme_bw() +
  theme(plot.title = element_text(face = 'bold', hjust = 0.5),
        plot.subtitle = element_markdown(),
        legend.position = 'none',
        legend.box = 'vertical',
        panel.grid = element_blank(),
        legend.text = element_markdown()) 

nmds_plot_hilic


```


# Creating metadata for plot

```{r}

### Creating Month lengend

metadata_plot <- tibble(Month = c('May', 'June', 'July_1', 'July_2',
                                          'July_3', 'August', 'September', 
                                          'October', 'May_22'),
                        Stage = c('Pre monsoon', 'Pre monsoon', 'During monsoon', 
                                  'During monsoon', 'During monsoon', 'During monsoon',
                                  'Post monsoon','Post monsoon', 'Pre monsoon'))

final_colors <- set_names(get_palette('Spectral', 12)[c(1:2, 9:12, 4, 5, 3)],
                          nm = levels(metadata_plot$Month))

plot_label <- metadata_plot %>% 
  mutate(Month = factor(Month, levels = c('May', 'June', 'July_1', 'July_2',
                                          'July_3', 'August', 'September', 
                                          'October', 'May_22'))) %>% 
  ggplot()+
  geom_point(aes(x = Month,
                 y = 1,
                 color = Month))+
  scale_color_manual(values = final_colors)+
  theme_bw()+
  theme(legend.position = 'bottom')

plot_label

leg_month <- get_legend(plot_label)


### Creating monsoon stage

stage_list <- metadata_plot %>% 
  dplyr::select(Stage) %>% 
  distinct() %>% 
  mutate(Stage = factor(Stage, levels = c('Pre monsoon', 'During monsoon',
                                          'Post monsoon')))
                         

stage_colors2 <- set_names(c('#FEDD7F', '#9bc1bc', '#ed6a5a'),
                          nm = levels(stage_list))

plot_stage <- metadata_plot %>% 
  mutate(Stage = factor(Stage, levels = c('Pre monsoon', 'During monsoon',
                                          'Post monsoon'))) %>% 
   mutate(Month = factor(Month, levels = c('May', 'June', 'July_1', 'July_2',
                                          'July_3', 'August', 'September', 
                                          'October', 'May_22'))) %>% 
  ggplot()+
  geom_point(aes(x = Stage,
                 y = 1,
                 color = Stage),
             shape = 15)+
  scale_color_manual(values = stage_colors2) +
  ggnewscale::new_scale_color() +
  geom_point(aes(x = Month,
                 y = 1,
                 color = Month),
             shape = 16) +
  scale_color_manual(values = final_colors) +
  guides(color = guide_legend(nrow = 2)) +
  theme_bw()+
  theme(legend.position = 'bottom',
        legend.box = 'vertical',
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 6),
        legend.key.size = unit(.1, 'lines'),
        legend.key.spacing = unit(0.3, 'lines'),
        legend.spacing.y = unit(-0.1, 'lines'))

plot_stage

leg_stage <- get_legend(plot_stage)
leg_stage

```



## RDA - HILIC MODE

# ENVS 

```{r}

evs <- read_csv("input/env_data_ready_2025.csv") %>% 
  filter(!sample == c('SAMP1-S1-22', 'SAMP1-S2-22', 'SAMP1-S3-22', 'SAMP1-S4-22'))%>% 
  mutate(sample = str_replace_all(sample, "-", "_")) %>% 
  rename(SampleID = sample)

```

# Microbial diversity

```{r}

avs_shannon <- read_csv("input/ASV_shannon.csv") %>% 
  mutate(sample = str_replace_all(sample, "-", "_")) %>% 
  rename(SampleID = sample) %>% 
  select(!dataset)

```


# Preprare matrix - envs and microbial diversity

```{r}

envs_microbe <- evs %>% 
  mutate(C_N = total_C/total_N) %>% 
  left_join(avs_shannon, by = 'SampleID') %>% 
  left_join(metadata_com, by = c('SampleID' = 'treatment_code')) %>% 
  dplyr::select(!c(SampleID, sampling, stage, month, site)) %>%
  rename(SampleID = SampleID_RP) %>% 
  column_to_rownames(var = 'SampleID') %>% 
  select(total_N, ph_mean, moisture, C_N, `NO3-N (ppm)`, `NH4-N (ppm)`, Shannon)

```


## Standarization of envs matrix

```{r}

env_std <- decostand(envs_microbe, method = 'standardize')

```

## Checking for collinearity 



```{r}
## Transpose normalized matrix

rp_df2 <- as.data.frame(t(norm_matrix))

# Building model with all variables

rda_model <- rda(rp_df2 ~ ., data = env_std)

plot(performance::check_collinearity(rda_model))

# Quick plot to see how it is working

ordiplot(rda_model, scaling = 1, choices = c(1,2))

```


```{r}

summary(rda_model)

anova.cca(rda_model, step = 1000)

anova.cca(rda_model, step = 1000, by = "term")

```


```{r}

# Selecting variables ----

fwd.sel <- ordiR2step(rda(rp_df2 ~ 1, data = env_std), # simple model,
                      scope = rda_model, # full model
                      direction = 'forward',
                      R2scope = TRUE,
                      pstep = 1000,
                      trace = TRUE,
                      na.action = na.exclude) 

fwd.sel$call

```

## Model based on foward selection

```{r}

envs_fwd_sel <- envs_microbe %>% 
  select(moisture, ph_mean, C_N)

env_std3 <- decostand(envs_fwd_sel, method = 'standardize')


# Building model with all variables

rda_model_fwd <- rda(rp_df2 ~ ., data = env_std3)


```

## Check R2 of the new model
```{r}


RsquareAdj(rda_model_fwd)
## 0.2

```


## Testing the singnificace of the model


```{r}

anova.cca(rda_model_fwd, step = 1000)

anova.cca(rda_model_fwd, step = 1000, by = 'term')


total_variance <- sum(rda_model_fwd$CCA$eig)/(sum(rda_model_fwd$CCA$eig)+ sum(rda_model_fwd$CA$eig)) *100

```


## RDA plot 

```{r}

#### Scaling 2

# Custom triplot code!

## extract % explained by the first 2 axes
perc <- round(100*(summary(rda_model_fwd)$cont$importance[2, 1:2]), 2)

## extract scores - these are coordinates in the RDA space
sc_sites <- as.data.frame(scores(rda_model_fwd, display="sites", choices=c(1,2), 
                                 scaling=2)) %>%
  rownames_to_column(var = 'SampleID') %>% 
  left_join(metadata_com, by = c('SampleID' = 'SampleID_HILIC')) %>% 
  mutate(color = my_colors_lc[month]) 


sc_species <- as.data.frame(scores(rda_model_fwd, display="species", choices=c(1,2), scaling=2))

sc_bp <- as.data.frame(scores(rda_model_fwd, display="bp", choices=c(1, 2), scaling=2)) %>% 
  rownames_to_column(var = 'envs')



RDA_plot_hilic <- sc_sites %>% 
  ggplot()+
  geom_point(aes(x = RDA1,
                 y = RDA2,
                 color = month),
             size = 2) +
  scale_color_manual(values = my_colors_lc)+
  geom_point(data = sc_species,
             aes(x = RDA1,
                 y = RDA2),
             fill = 'gray80',
             size = 1,
             shape = 22) +
  geom_segment(data = sc_bp, 
               aes(x = 0, y = 0, 
                   xend = 3*RDA1, yend = 3*RDA2), 
               color = "black", 
               linewidth = 0.5,
               arrow = arrow(length = unit(.5, 'lines'))) +
  geom_text(data = sc_bp,
            aes(x = 3*RDA1+.3, y = 3*RDA2 +.3,
                label = envs),
            color = "black",
            fontface = "bold",
            size = 2) +
  labs(title = "Metabolomics - RDA - Scaling 2",
       x = paste0("RDA1 (", perc[1], "%)" ) ,
       y = paste0("RDA2 (", perc[2], "%)"))+
  theme_bw() +
  theme(plot.title = element_text(face = 'bold', hjust = 0.5),
        plot.subtitle = element_markdown(),
        panel.grid = element_blank(),
        legend.position = 'none',
        legend.box = 'vertical',
        legend.text = element_markdown()) 

RDA_plot_hilic

```

## Metabolite classes

## HILIC

```{r}

## Loading class of compounds

name_class_hilic_all <- read_csv("scripts/LC-MS-MS/4.Get_inchis/output/HILIC_classyfire_from_name.csv") %>% 
  select(FeatureID, Name, superclass, class, subclass, level_5 = `level 5`) %>% 
  mutate(LC_mode = 'HILIC')

canopus_class_hilic_all <- read_csv("scripts/LC-MS-MS/4.Get_inchis/output/HILIC_classyfire_canopus.csv") %>% 
  select(FeatureID, Name,
         superclass = `ClassyFire#superclass`,
         class = `ClassyFire#class`, 
         subclass = `ClassyFire#subclass`, 
         level_5 = `ClassyFire#level 5`) %>% 
  mutate(LC_mode = 'HILIC')

```

## RP

```{r}

name_class_rp_all <- read_csv("scripts/LC-MS-MS/4.Get_inchis/output/RP_classyfire_from_name_ready.csv") %>% 
  dplyr::select(FeatureID, Name, superclass, class, subclass, level_5 = `level 5`) %>% 
  mutate(LC_mode = 'RP')

canopus_class_rp_all <- read_csv("scripts/LC-MS-MS/4.Get_inchis/output/RP_classyfire_canopus_ready.csv") %>% 
  dplyr::select(FeatureID, Name,
                superclass = `ClassyFire#superclass`,
                class = `ClassyFire#class`, 
                subclass = `ClassyFire#subclass`, 
                level_5 = `ClassyFire#level 5`) %>% 
  mutate(LC_mode = 'RP')


```



## Combine class info in a single df

```{r}

class_info_rp_all <- rbind(name_class_rp_all, canopus_class_rp_all) %>% 
  mutate(superclass = ifelse(is.na(superclass), 'No assigned', superclass),
         class = ifelse(is.na(class), glue::glue('Unassigned class of {superclass}'),
                        class),
         subclass = ifelse(is.na(subclass), glue::glue('Unassigned subclass of {class}'),
                           subclass))


class_info_hilic_all <- rbind(name_class_hilic_all, canopus_class_hilic_all)%>%
  mutate(superclass = ifelse(is.na(superclass), 'No assigned', superclass),
         class = ifelse(is.na(class), glue::glue('Unassigned class of {superclass}'),
                        class),
         subclass = ifelse(is.na(subclass), glue::glue('Unassigned subclass of {class}'),
                           subclass))

```


## Piechart metabolite class - RP mode

```{r}


all_clusters <- class_info_rp_all %>% 
  select(FeatureID, superclass, class) %>% 
  distinct() %>% 
  group_by(superclass, class) %>% 
  summarise(class_n = n()) %>% 
  group_by(superclass) %>% 
  mutate(superclass_n = sum(class_n),
         superclass = factor(superclass),
         class = str_replace(class, 'Unassigned class of No assigned', 'No assigned')) %>% 
  arrange(superclass, class)

superclass_clusters <- all_clusters %>%
  select(superclass, superclass_n) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(perc = superclass_n / sum(superclass_n),
         label_pos = cumsum(perc) - 0.5 * perc)

class_clusters <- all_clusters %>%
  ungroup() %>% 
  select(superclass, class, class_n) %>%
  distinct() %>% 
  mutate(class = factor(class, levels = rev(all_clusters$class))) %>% 
  arrange(class) %>% 
  mutate(perc = class_n / sum(class_n),
         label_pos = cumsum(perc) - 0.5 * perc)
```


## Pie chart

```{r}

pie_plot_rp <- superclass_clusters %>% 
  ggplot() +
  geom_col(aes(x = 'A',
               y = perc,
               fill = superclass),
           color = 'white') +
  geom_col(data = class_clusters,
           aes(x = 'B',
               y = perc,
               fill = superclass,
               alpha = 0.7),
           color = 'white',
           show.legend = FALSE) +
  # geom_text(aes(x = 'A',
  #               y = 1 - label_pos,
  #               label = superclass,
  #           size = 2)) +
  geom_text(data = class_clusters %>% filter(perc > 0.03),
            (aes(x = 'B',
                 y = label_pos,
                 label = class)),
            size = 2) +
  scale_x_discrete(limits = c('', 'A', 'B')) +
  scale_fill_manual(values = ggpubr::get_palette('Paired', 15)) +
  theme_void() +
  guides(fill = guide_legend(ncol = 3)) +
  coord_polar('y')+
  theme(legend.position = 'bottom',
        legend.title = element_text(size = 7),
        legend.title.position = 'top',
        legend.text = element_text(size = 5),
        legend.key.size = unit(.5, 'lines'),
        legend.key.spacing = unit(0.3, 'lines'),
        legend.spacing.y = unit(-0.1, 'lines'))

pie_plot_rp

```


## Saving plot

```{r}

## Final figure
des <- "
ABE
CCE
DDE
"


final_fig <- nmds_plot_hilic + RDA_plot_hilic + leg_stage + free(pie_plot_rp)+ plot_all + 

  plot_layout(design = des,
              heights = c(1, 0.1, 1)) +
  plot_annotation(tag_levels = list(c('A', 'B', '', 'C', 'D'))) &
  theme(text = element_text(size = 6))

final_fig

```



## Save final plot


```{r}

ggsave("output_figures/Sup_figure_1.svg", final_fig, dpi = 300,
       units = 'mm', 
       height = 310, 
       width = 300)

ggsave("output_figures/Sup_figure_1.png", final_fig, dpi = 300,
       units = 'mm', 
       height = 310, 
       width = 300)


```



