---
title: "Figure 2"
author: "Viviana Freire Zapata"
date: "3/17/2025"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: '2'
    collapsed: false
    highlight: tango
    css: style.css
editor_options: 
  chunk_output_type: console
---

# Data Pre-processing

## 1. Setup

### Importing libraries

```{r libraries, message=FALSE, warning=FALSE}
library(easystats)
library(matrixStats)
library(vegan)
library(GGally)
library(ggpubr)
library(glmnet)
library(ggbeeswarm)
library(lme4)
library(lmerTest)
library(readxl)
library(pheatmap)
library(ggh4x)
library(dtwclust)
library(patchwork)
library(tidyverse)
source('functions_cdis_exploration.R')
source('functions_cdis_norm_stats.R')
source('functions_extra.R')
```


# Loading data RP

```{r}
## Raw matrix

RP_df <- read_csv("../../../../UPDATE_LC_2024/1.Pre-processing/RP_matrix_clean_2024.csv")


## Loading annotation

annot_rp <- read_csv("../../../../UPDATE_LC_2024/3.Cleanning_Annotation/output/RP_final_annotation_table_1_2_25.csv")


```

# Loading data HILIC

```{r}
## Raw matrix

HILIC_df <- read_csv("../../../../UPDATE_LC_2024/1.Pre-processing/HILIC_matrix_clean_2024.csv")


## Loading annotation

annot_hilic <- read_csv("../../../../UPDATE_LC_2024/3.Cleanning_Annotation/output/HILIC_final_annotation_table_12_3_24.csv")


```


## Loading significant features

```{r}

sig_features_rp <- read_csv("../../../../UPDATE_LC_2024/5.RP_stats/output/Sig_features_model_fwd_selected_RP.csv")


sig_features_hilic <- read_csv("../../../../UPDATE_LC_2024/5.HILIC_stats/output/Sig_features_model_fwd_selected.csv")


```


# Metadata

```{r}
metadata_RP <- read_csv("../../../../UPDATE_LC_2024/input/RP_metadata_arid_field.csv") %>% 
    mutate(month = str_replace(month, 'may', 'May'),
         month = str_replace(month, 'june', 'June'),
         month = str_replace(month, 'july_1', 'July_1'),
         month = str_replace(month, 'july_2', 'July_2'),
         month = str_replace(month, 'july_3', 'July_3'),
         month = str_replace(month, 'august', 'August'),
         month = str_replace(month, 'september','September'),
         month = str_replace(month, 'october', 'October'))

metadata_HILIC <- read_csv("../../../../UPDATE_LC_2024/input/HILIC_metadata_arid_field.csv") %>% 
    mutate(month = str_replace(month, 'may', 'May'),
         month = str_replace(month, 'june', 'June'),
         month = str_replace(month, 'july_1', 'July_1'),
         month = str_replace(month, 'july_2', 'July_2'),
         month = str_replace(month, 'july_3', 'July_3'),
         month = str_replace(month, 'august', 'August'),
         month = str_replace(month, 'september','September'),
         month = str_replace(month, 'october', 'October'))

```

## Prepare matrix RP

```{r}

RP_ready <- RP_df %>% 
  select(FeatureID, contains("RP")) %>% 
  filter(FeatureID %in% annot_rp$FeatureID) %>% 
  column_to_rownames(var = "FeatureID")

```


# Metabolite Normalization

## Median normalization

```{r}

norm_matrix_rp <- median.norm(RP_ready, transform_data = TRUE)

```


## Filtering full matrix 

```{r}

norm_matrix_sig_rp <- norm_matrix_rp %>% 
  rownames_to_column(var = 'FeatureID') %>% 
  filter(FeatureID %in% sig_features_rp$FeatureID) %>% 
  pivot_longer(!FeatureID, names_to = 'SampleID', values_to = 'norm_int')

```


## Prepare matrix HILIC

```{r}

hilic_ready <- HILIC_df %>% 
  select(FeatureID, contains("HILIC")) %>% 
  filter(FeatureID %in% annot_hilic$FeatureID) %>% 
  column_to_rownames(var = "FeatureID")

```


# Metabolite Normalization

## Median normalization

```{r}

norm_matrix_hilic <- median.norm(hilic_ready, transform_data = TRUE)

```



## Filtering full matrix HILIC

```{r}

norm_matrix_sig_hilic <- norm_matrix_hilic %>% 
  rownames_to_column(var = 'FeatureID') %>% 
  filter(FeatureID %in% sig_features_hilic$FeatureID) %>% 
  pivot_longer(!FeatureID, names_to = 'SampleID', values_to = 'norm_int')


```

## Checking similar sig features between HILIC and RP


```{r}

sig_hilic_2 <- sig_features_hilic %>% 
  select(FeatureID, Name, Formula, Annotation_level) %>% 
  distinct() %>% 
  mutate(LC_mode = 'HILIC')


sig_rp_2 <- sig_features_rp %>% 
  select(FeatureID, Name, Formula, Annotation_level) %>% 
  distinct() %>% 
  mutate(LC_mode = 'RP')

combined_sig <- rbind(sig_hilic_2, sig_rp_2)


```


## Filter features with same NAME found by both modes, priority features high annotation level


```{r}

delete_repeated <- c('FT_10.755_259.12997', 'FT_13.321_503.12848', 'FT_11.478_245.11431')


combined_sig_ft <- combined_sig %>% 
  filter(!FeatureID %in% delete_repeated)

```

## Adding class info

## HILIC

```{r}

## Loading class of compounds

name_class_hilic <- read_csv("../../UPDATE_LC_2024/4.Get_inchis/output/HILIC_classyfire_from_name.csv") %>% 
  filter(FeatureID %in% sig_features_hilic$FeatureID) %>% 
  select(FeatureID, Name, superclass, class, subclass, level_5 = `level 5`) %>% 
  mutate(LC_mode = 'HILIC')

canopus_class_hilic <- read_csv("../../UPDATE_LC_2024/4.Get_inchis/output/HILIC_classyfire_canopus.csv") %>% 
  filter(FeatureID %in% sig_features_hilic$FeatureID) %>% 
  select(FeatureID, Name,
         superclass = `ClassyFire#superclass`,
         class = `ClassyFire#class`, 
         subclass = `ClassyFire#subclass`, 
         level_5 = `ClassyFire#level 5`) %>% 
  mutate(LC_mode = 'HILIC')

```

## RP

```{r}

name_class_rp <- read_csv("../../UPDATE_LC_2024/4.Get_inchis/output/RP_classyfire_from_name_ready.csv") %>% 
  filter(FeatureID %in% sig_features_rp$FeatureID) %>% 
  select(FeatureID, Name, superclass, class, subclass, level_5 = `level 5`) %>% 
  mutate(LC_mode = 'RP')

canopus_class_rp <- read_csv("../../UPDATE_LC_2024/4.Get_inchis/output/RP_classyfire_canopus_ready.csv") %>% 
  filter(FeatureID %in% sig_features_rp$FeatureID) %>% 
  select(FeatureID, Name,
         superclass = `ClassyFire#superclass`,
         class = `ClassyFire#class`, 
         subclass = `ClassyFire#subclass`, 
         level_5 = `ClassyFire#level 5`) %>% 
  mutate(LC_mode = 'RP')

```


## Combine class info in a single df

```{r}

class_info <- rbind(name_class_hilic, canopus_class_hilic, 
                    name_class_rp, canopus_class_rp)

```


## Joining CLASS info to df

```{r}

combined_sig_ready <- combined_sig_ft %>% 
  left_join(class_info, by = c('FeatureID', 'Name', 'LC_mode')) %>% 
  distinct()

```


## Heatmap

```{r}

metadata_ready_rp <- metadata_RP %>% 
  mutate(month = factor(month, levels= c('May', 'June', 'July_1', 'July_2',
                                         'July_3', 'August', 'September','October')),
         stage = factor(stage, levels = c('Pre_monsoon', 'During_monsoon',
                                          'Post_monsoon')))


metadata_ready_hilic <- metadata_HILIC %>% 
  mutate(month = factor(month, levels= c('May', 'June', 'July_1', 'July_2',
                                         'July_3', 'August', 'September', 'October')),
         stage = factor(stage, levels = c('Pre_monsoon', 'During_monsoon',
                                          'Post_monsoon')))

## Joining both matrices

norm_matrix_sig_hilic_2 <- norm_matrix_sig_hilic %>% 
  left_join(metadata_ready_hilic, by = 'SampleID') 

norm_matrix_sig_rp_2 <- norm_matrix_sig_rp %>% 
  left_join(metadata_ready_rp, by = 'SampleID')

df_combined <- rbind(norm_matrix_sig_hilic_2, norm_matrix_sig_rp_2) %>% 
  group_by(FeatureID, month) %>% 
  summarise(norm_int = mean(norm_int)) %>% 
  left_join(combined_sig_ready, by = 'FeatureID')


## Heatmap COMBINED

color_strips <- strip_nested(
  background_y = elem_list_rect(fill = c("white")))


heat_df <- df_combined %>%
  group_by(FeatureID) %>%
  mutate(mean_order = mean(norm_int)) %>%
  ungroup() %>%
  group_by(superclass) %>% 
  arrange(superclass, mean_order) %>%
  mutate(order = n():1) %>%
  group_by(FeatureID) %>% 
  mutate(zscore = (norm_int - mean(norm_int)) / sd(norm_int))

## Setting color palette


my_colors_annot <- set_names(get_palette('Set2', length(unique(heat_df$Annotation_level))),
                             nm = unique(heat_df$Annotation_level))


my_colors_class <- set_names(get_palette('Paired', length(unique(heat_df$superclass))),
                             nm = c(sort(unique(heat_df$superclass)), NA))

###

names_fixed <- heat_df %>% 
  select(FeatureID, Name) %>% 
  mutate(Name = ifelse(is.na(Name), 'no name', Name)) %>% 
  distinct() %>% 
  group_by(Name) %>% 
  mutate(n = n():1,
         Name_fix = ifelse(n > 1, glue::glue('{Name}_{n}'), Name)) %>% 
  ungroup() %>% 
  select(FeatureID, Name_fix)


heat_mat <- heat_df %>% 
  ungroup() %>% 
  left_join(names_fixed, by = 'FeatureID') %>% 
  select(Name_fix, month, zscore) %>% 
  pivot_wider(names_from = month, values_from = zscore) %>% 
  column_to_rownames(var = 'Name_fix')

heat_dendo <- hclust(dist(heat_mat), 'ave')

## Adding class infor

factor_tile <- heat_df %>% 
  ggplot() +
  geom_tile(aes(x = 1,
                y = FeatureID,
                fill = superclass)) +
  scale_y_dendrogram(hclust = heat_dendo) +
  scale_fill_manual(values = my_colors_class)+
  guides(fill = guide_legend(ncol = 3)) +
  theme_void() +
  theme(legend.position = 'bottom',
        axis.ticks.length.y = unit(5, 'pt'),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 5),
        legend.title = element_text(size = 6),
        legend.box.spacing = unit(.05, 'lines'),
        legend.key.size = unit(.5, 'lines'),
        legend.title.position = 'top')

factor_tile



heat_plot <- heat_df %>% 
  left_join(names_fixed, by = 'FeatureID') %>% 
  ggplot(aes(x = month,
             y = Name_fix,
             fill = zscore)) +
  geom_tile()+
  scale_y_dendrogram(hclust = heat_dendo) +
  scale_fill_distiller(palette = 'RdBu', limits = c(-2.5, 2.5)) +
  labs(y = 'FeatureID',
       fill = 'Relative intensity')+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7),
        axis.text.y = element_text(size = 4),
        strip.text.y = element_text(angle = 0),
        axis.ticks = element_blank(),
        axis.ticks.length.y = unit(0, 'pt'),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = 'bottom',
        legend.text = element_text(size = 5),
        legend.title = element_text(size = 6),
        legend.box.spacing = unit(.05, 'lines'),
        legend.key.size = unit(.5, 'lines'),
        legend.title.position = 'top')

heat_plot


lay_heat <- "BA"

heat_ready <- heat_plot + factor_tile +
  plot_layout(guides = 'collect', widths = c(.05, 1), design = lay_heat) &
  theme(legend.position = 'bottom')


heat_ready

```


# Dynamic time warping - RP MODE


## Prepare matrix for DTW analysis 
## metabolites in rows, Month in columns - Time in order

## Working only with significant features

```{r}

data_dtw_RP <- norm_matrix_sig_rp %>% 
  left_join(metadata_RP, by = 'SampleID') %>% 
  group_by(FeatureID, month) %>% 
  summarise(mean_intensity = mean(norm_int)) %>% 
  pivot_wider(names_from = month, values_from = mean_intensity) %>% 
  column_to_rownames(var = "FeatureID")

data_ready_ord_RP <- data_dtw_RP[,c('May', 'June', 'July_1',
                             'July_2', 'July_3', 'August', 'September',
                             'October')]

```


## Dynamic time warping

```{r}

dwt_test_RP <- tsclust(data_ready_ord_RP, 
                    type = 'h', 
                    k = 5,
                    preproc = zscore,
                    distance = 'sbd',
                    centroid = shape_extraction,
                    control = hierarchical_control(method = 'average'))


dwt_test_RP@clusinfo

plot(dwt_test_RP, type = 'sc')


```


# Dynamic time warping - HILIC MODE

## Working only with significant features

```{r}

data_dtw_HILIC <- norm_matrix_sig_hilic %>% 
  left_join(metadata_HILIC, by = 'SampleID') %>% 
  group_by(FeatureID, month) %>% 
  summarise(mean_intensity = mean(norm_int)) %>% 
  pivot_wider(names_from = month, values_from = mean_intensity) %>% 
  column_to_rownames(var = "FeatureID")

data_ready_ord_HILIC <- data_dtw_HILIC[,c('May', 'June', 'July_1',
                             'July_2', 'July_3', 'August', 'September',
                             'October')]

```


## 10. Dynamic time warping

```{r}

dwt_test_HILIC <- tsclust(data_ready_ord_HILIC, 
                    type = 'h', 
                    k = 5,
                    preproc = zscore,
                    distance = 'sbd',
                    centroid = shape_extraction,
                    control = hierarchical_control(method = 'average'))


dwt_test_HILIC@clusinfo

plot(dwt_test_HILIC, type = 'sc')


```


### Classes of metabolites from Dwt

## Adding class info

## Final plot - Combining HILIC and RP

```{r}

clusts_rp <- tibble(FeatureID = names(dwt_test_RP@cluster),
                 cluster = paste0('Cluster ', dwt_test_RP@cluster))

clusts_hilic <- tibble(FeatureID = names(dwt_test_HILIC@cluster),
                 cluster = paste0('Cluster ', dwt_test_HILIC@cluster))


```

## RP

```{r}
feature_dwt_clusts_rp <- as_tibble(dwt_test_RP@datalist) %>% 
  rowid_to_column(var = 'time') %>% 
  mutate(time = case_when(time == 1 ~ 'May',
                          time == 2 ~ 'June',
                          time == 3 ~ 'July_1',
                          time == 4 ~ 'July_2',
                          time == 5 ~ 'July_3',
                          time == 6 ~ 'August',
                          time == 7 ~ 'September',
                          time == 8 ~ 'October')) %>% 
  mutate(time = factor(time, levels = c('May', 'June', 'July_1', 'July_2', 'July_3',
                                        'August', 'September', 'October'))) %>% 
  pivot_longer(!time, names_to = 'FeatureID', values_to = 'value') %>% 
  left_join(clusts_rp, by = 'FeatureID') %>% 
  mutate(cluster = factor(cluster, levels = paste0('Cluster ', 1:5)),
         LC_mode = 'RP')


rename_vec <- set_names(1:length(unique(dwt_test_RP@cluster)),
                        paste0('Cluster ', unique(dwt_test_RP@cluster)))


```

## Centroids RP
```{r}
dwt_centroids_RP <- as.data.frame(dwt_test_RP@centroids) %>% 
  rename(rename_vec) %>% 
  rowid_to_column(var = 'time') %>% 
  pivot_longer(!time, names_to = 'cluster', values_to = 'value') %>% 
  mutate(cluster = factor(cluster, levels = paste0('Cluster ', 1:5)),
         LC_mode = 'RP')

```

## HILIC

```{r}
feature_dwt_clusts_hilic <- as_tibble(dwt_test_HILIC@datalist) %>% 
  rowid_to_column(var = 'time') %>% 
  mutate(time = case_when(time == 1 ~ 'May',
                          time == 2 ~ 'June',
                          time == 3 ~ 'July_1',
                          time == 4 ~ 'July_2',
                          time == 5 ~ 'July_3',
                          time == 6 ~ 'August',
                          time == 7 ~ 'September',
                          time == 8 ~ 'October')) %>% 
  mutate(time = factor(time, levels = c('May', 'June', 'July_1', 'July_2', 'July_3',
                                        'August', 'September', 'October'))) %>% 
  pivot_longer(!time, names_to = 'FeatureID', values_to = 'value') %>% 
  left_join(clusts_hilic, by = 'FeatureID') %>% 
  mutate(cluster = factor(cluster, levels = paste0('Cluster ', 1:5)),
         LC_mode = 'HILIC')


rename_vec <- set_names(1:length(unique(dwt_test_HILIC@cluster)),
                        paste0('Cluster ', unique(dwt_test_HILIC@cluster)))


```

## Centroids HILIC
```{r}
dwt_centroids_HILIC <- as.data.frame(dwt_test_HILIC@centroids) %>% 
  rename(rename_vec) %>% 
  rowid_to_column(var = 'time') %>% 
  pivot_longer(!time, names_to = 'cluster', values_to = 'value') %>% 
  mutate(cluster = factor(cluster, levels = paste0('Cluster ', 1:5)),
         LC_mode = 'HILIC')

```

## Combined HILIC + RP DTW plot

```{r}

combined_dtw <- rbind(feature_dwt_clusts_rp, feature_dwt_clusts_hilic) %>% 
  left_join(class_info, by = c('FeatureID', 'LC_mode')) %>% 
  filter(!is.na(superclass))


dtw_centroids_combined <- rbind(dwt_centroids_HILIC, dwt_centroids_RP)


feature_dwt_plot <- combined_dtw %>% 
  ggplot(aes(x = time,
             y = value))+
  geom_line(aes(group = FeatureID,
                color = superclass),
            alpha = 0.5)+
  geom_line(data = . %>% 
              group_by(LC_mode, cluster, time) %>% summarise(cluster_mean = mean(value)), 
            aes(y = cluster_mean, 
                group = cluster), 
            color = "gray15", size = 1)+
  geom_vline(xintercept = c(1.5, 5.5), linetype = 'dashed')+
  # ggplot() +
  # geom_line(aes(x = time,
  #               y = value,
  #               group = FeatureID, 
  #               color = superclass
  #               ),
  #           size = 0.8) +
  # geom_line(data = dwt_centroids,
  #           aes(y = value,
  #               x = time,
  #               group = 1),
  #           linetype = 'dashed',
  #           color = 'black',
  #           linewidth = 1,
  #           alpha = 0.8) +
  scale_color_manual(values = my_colors_class)+
  #labs(title = 'Significant features')+
  facet_wrap2(~LC_mode + cluster,
              scales = 'free',
              nrow = 2, 
              strip = strip_nested())+
  labs(y = 'Zscore') +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 7),
        strip.text.x = element_text(face = 'bold'),
        legend.position = 'none',
        strip.background = element_rect(fill = 'white'),
        panel.grid = element_blank(),
        strip.text.x.top = element_text(size = 6),
        strip.background.x = element_blank())

feature_dwt_plot


```

## Metabolite bubble plot
```{r}

selected_clus <- c('Cluster 1_HILIC', 'Cluster 5_HILIC', 'Cluster 3_RP', 'Cluster 4_RP')

plot_test <- combined_dtw %>% 
  mutate(temp = glue::glue('{cluster}_{LC_mode}')) %>% 
  filter(temp %in% selected_clus) %>% 
  filter(!superclass == 'Alkaloids and derivatives') %>% 
  select(!c(time, value)) %>% 
  distinct() %>% 
  group_by(cluster, LC_mode, superclass, class, subclass) %>% 
  count() %>% 
  ggplot() +
  geom_point(aes(x = cluster,
                 y = subclass,
                 size = n,
                 fill = superclass),
             shape = 21) +
  #labs(title = 'Significant features')+
  facet_grid2(cols = vars(LC_mode),
              rows = vars(superclass),
              scales = 'free')+
  scale_fill_manual(values = my_colors_class)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5),
        #axis.text.y = element_blank(),
        axis.title.x = element_blank(),
        strip.text.x = element_text(face = 'bold'),
        strip.text.y = element_text(angle = 0, face = 'bold'),
        legend.position = 'none',
        strip.background = element_rect(fill = 'white'),
        panel.grid = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_text(size = 7),
        strip.text.y.right = element_text(size = 6),
        strip.text.x.top = element_text(size = 6),
        strip.background.x = element_blank(),
        strip.background.y = element_blank())

plot_test


ggsave("output/class_metabolites_ft.png", plot_test, dpi = 300,
       units = 'mm',
       height = 220, 
       width = 180)


plot_test_name <- combined_dtw %>% 
  mutate(temp = glue::glue('{cluster}_{LC_mode}')) %>% 
  filter(temp %in% selected_clus) %>% 
  filter(!Name == 'no name') %>% 
  filter(subclass == 'Carbohydrates and carbohydrate conjugates') %>% 
  filter(str_detect(temp, 'HILIC')) %>% 
  select(Name, temp) %>% 
  distinct()

```


## Final plot

```{r}

lay <- "
BAC
BAC
BAD
BAD
"


final <- heat_ready + free(feature_dwt_plot) + plot_test + 
  plot_layout(design = lay, widths = c(.1, 1, 2)) +
  plot_annotation(tag_levels = list(c('', 'A', 'B', 'C')))

final


```

## Saving plot

```{r}

ggsave("output/Figure_2.svg", final, dpi = 300,
       units = 'mm', 
       height = 220, 
       width = 180)


ggsave("output/Figure_2.png", final, dpi = 300,
       units = 'mm',
       height = 220, 
       width = 180)

```

