---
title: "Omic integration"
author: "Viviana Freire Zapata"
output: html_notebook
editor_options: 
  chunk_output_type: console
---


```{r}
library(DESeq2)
library(edgeR)
library(limma)
library(patchwork)
library(ggh4x)
library(readxl)
library(ggpubr)
library(vegan)
library(ggrepel)
library(clusterProfiler)
library(VennDiagram)
library(mixOmics)
library(ggtext)
library(devtools)
library(SLIDE)
library(pROC)
library(caret)
library(ROCR)
library(tidyverse)
source('scripts/LC-MS-MS/5.Metabolite_model/functions_cdis_norm_stats.R')

```

# Loading data

## Metadata

```{r}
metadata <- read_csv('input/metadata_omics.csv') 

metadata_ready <- metadata %>% 
  mutate(Month = factor(month, levels = c("May", "July_1", "July_2", "July_3",
                                          "August", "October", "May_22"))) 

metatadata_metaT <- metadata_ready %>% 
  filter(!is.na(SampleID_metaT))
```

## Taxonomy

```{r}

mag_taxonomy <- read_csv('input/mag_taxonomy_denovo.csv') %>% 
  mutate(Class = ifelse(is.na(Class), 
                        glue::glue('Unassigned Class \nof {Phylum}'),
                        Class))

```


```{r}
## annotation

annot <- read_csv("input/annot_with_cydbs_2.23.25.csv") %>% 
  rename(geneid = `...1`)
```

# Metagenomics

## Laoding metaG TMM noralized counts

```{r}
# 
# tmm_mag <- read_csv("output/TMM_gene_counts_mags_2025.csv") %>% 
#   column_to_rownames(var = 'mag_geneid')

```

## Keep more abundant features and Scale

```{r}
# Calculate variance per feature
#  MAG-KO matrix, rows = features, columns = samples
# 
# var_genes <- apply(tmm_mag, 1, var)
# 
# # keep top 1% most variable features
# 
# tmm_mag_ft <- tmm_mag[var_genes > quantile(var_genes, 0.995), ]
# 
# # Scale
# 
# tmm_mag_scaled <- t(scale(t(tmm_mag_ft)))

```


# Metatranscriptomics

## Loading metaT data

```{r}

## Raw counts

rnaseq_counts <- read_tsv('input/rnaseq_counts_final_2025.txt', skip = 1) %>%
  filter(!str_detect(Geneid, 'RNA'))

## Sequences names

sra_table <- read_csv("input/SraRunTable.csv") %>%
  dplyr::select(Run, `Sample Name`) %>%
  mutate(treatment_code = str_replace_all(`Sample Name`, '-', '_'))


```


```{r}
# Fixing names

raw_matrix <- rnaseq_counts %>%
  dplyr::select(!c(Chr, Start, End, Strand, Length)) %>%
  column_to_rownames(var = "Geneid") %>%
  rename_with(.fn = function(x) str_remove(x, '/xdisk/tfaily/vfreirezapata/metat_2025_final/bowtie_results/')) %>%
  rename_with(.fn = function(x) str_remove(x, '_mapped_sorted.bam'))

# Summing counts per sample

matrix_sum <- raw_matrix %>%
  rownames_to_column(var = 'Geneid') %>%
  pivot_longer(!Geneid, names_to = 'Run', values_to = 'counts') %>%
  left_join(sra_table, by = 'Run') %>%
  group_by(treatment_code, Geneid) %>%
  summarise(counts_sum = sum(counts))


```

## Creating matrix per KO (sum expression of genes with same annotation)

```{r}

countData_long <- matrix_sum %>%
  left_join(annot, by = c('Geneid' = 'geneid')) %>%
  filter(!is.na(ko_id)) %>%
  group_by(fasta, ko_id, treatment_code) %>%
  summarise(sum_counts = sum(counts_sum))

## Filter low-count KOs

countData_filter <- countData_long %>%
  filter(sum_counts > 20) ## Filtering low-count KOs

countData_temp <- countData_long %>%
  pivot_wider(names_from = 'treatment_code', values_from = 'sum_counts', values_fill = 0) %>%
  ungroup() %>%
  unite(fasta_ko, fasta, ko_id, sep = '__') %>%
  column_to_rownames(var = 'fasta_ko')

countData_ready <- countData_temp[rowSums(countData_temp != 0) >= 10,
                                  metatadata_metaT$treatment_code]

```

## Normalize counts - edgeR + voom

```{r}

# KO_counts: rows = KO, cols = samples

dge <- DGEList(counts = countData_ready)

keep <- filterByExpr(dge, group = metatadata_metaT$month)

dge <- dge[keep, keep.lib.sizes=FALSE]

dge <- calcNormFactors(dge, method = "TMM")

design <- model.matrix(~0 + metatadata_metaT$month)

v <- voom(dge, design=design, plot=TRUE)   # v$E is log2-CPM (voom)

expr <- v$E

expr_scaled <- t(scale(t(expr)))# scale KOs (rows) before integration

# Save v$weights for diagnostics if needed

```

## Keep only the top 5% more variant genes

```{r}

var_genes_metat <- apply(expr, 1, var)

expr_filt <- expr[var_genes_metat > quantile(var_genes_metat, 0.95), ]

```

## Sanitycheck outliers / batch effects

```{r}

# PCA before normalization and scaling

# 1. Transpose (PCA expects samples as rows)
count_t <- t(countData_ready)

# 2. Run PCA
pca_res <- prcomp(count_t, scale. = TRUE)

# Build data frame for ggplot
pca_df <- as.data.frame(pca_res$x)
pca_df2 <- pca_df %>% 
  rownames_to_column(var = 'treatment_code') %>% 
  left_join(metatadata_metaT, by = 'treatment_code')

ggplot(pca_df2, aes(x = PC1, y = PC2, color = month)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal(base_size = 14) +
  labs(x = paste0("PC1 (", round(100*summary(pca_res)$importance[2,1], 1), "%)"),
       y = paste0("PC2 (", round(100*summary(pca_res)$importance[2,2], 1), "%)")) +
  scale_color_brewer(palette = "Set2")

# PCA after normalization and scaling

# 1. Transpose (PCA expects samples as rows)
expr_t <- t(expr)

# 2. Run PCA
pca_res <- prcomp(expr_t, scale. = TRUE)

# Build data frame for ggplot
pca_df <- as.data.frame(pca_res$x)

pca_df2 <- pca_df %>% 
  rownames_to_column(var = 'treatment_code') %>% 
  left_join(metatadata_metaT, by = 'treatment_code')

ggplot(pca_df2, aes(x = PC1, y = PC2, color = month)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal(base_size = 14) +
  labs(x = paste0("PC1 (", round(100*summary(pca_res)$importance[2,1], 1), "%)"),
       y = paste0("PC2 (", round(100*summary(pca_res)$importance[2,2], 1), "%)")) +
  scale_color_brewer(palette = "Set2")

```


## Metabolomics


# Loading data 

```{r}
## Raw matrix

RP_df <- read_csv("scripts/LC-MS-MS/1.Pre-processing/output/RP_matrix_clean_2025.csv")


## Loading annotation

annot_metabo <- read_csv("scripts/LC-MS-MS/3.Cleanning_Annotation/output/RP_final_annotation_checked.csv")


```

## Prepare matrix

```{r}

RP_ready <- RP_df %>% 
  dplyr::select(FeatureID, contains("RP")) %>% 
  filter(FeatureID %in% annot_metabo$FeatureID) %>% 
  column_to_rownames(var = "FeatureID")

```

# Feature filtering

```{r}
### Calculating IQR to keep only traits that are changing more
## Filtering metabolites and genes that doesn't change much
## Idea decrease the number of features to have a more balance matrices 

iqr_test_matrix <- rowIQRs(as.matrix(RP_ready))

iqr_res <- as.data.frame(as.matrix(iqr_test_matrix))

# Filtering according to metaboAnalyts

# 1) Less than 250 variables: 5% will be filtered; 
# 2) Between 250 - 500 variables: 10% will be filtered; 
# 3) Between 500 - 1000 variables: 25% will be filtered; and 
# 4) Over 1000 variables: 40% will be filtered.

## 10% of the features will be filtered

# Sorting from largest to smallest (higher ranking = low iqr)
rk <- rank(-iqr_test_matrix, ties.method = 'random')

# Determining the number of remaining features
remain.num <- nrow(RP_ready)*.9

remain <- rk <= remain.num

features_keep <- rownames(RP_ready)[remain] 

## Filter matrix

matrix_LC_ft <- RP_ready[features_keep,]

## Note: 919 features kept for downstream analysis
```


# Metabolite Normalization and scaling

## Median normalization

```{r}

LC_matrix_ready <- median.norm(matrix_LC_ft, transform_data = TRUE)

```


# DIABLO

## Step 1 - PLS-DA
## Following - mixOMics tutorial 
## https://mixomicsteam.github.io/mixOmics-Vignette/id_05.html

## Running NON scal

```{r}
## X matrix - samples in rows, metabolites in columns (24 samples x 1624 metabolites)

## Matching sample names

# Metabolomics

metabolomics_mx <- as.data.frame(t(LC_matrix_ready)) %>% 
  rownames_to_column(var = 'sampleid') %>% 
  left_join(metadata_ready, by = c('sampleid' = 'SampleID_RP')) %>% 
  filter(!is.na(SampleID_metaT)) %>% 
  dplyr::select(!c(SampleID_metaT, sampleid, SampleID_HILIC, site, sampling,
                   month, stage, Month)) %>%
  column_to_rownames(var = 'treatment_code')

# Metagenomics

# metag_mx <- as.data.frame(t(tmm_mag_scaled)) %>% 
#   rownames_to_column(var = 'sampleid') %>% 
#   mutate(sampleid = str_replace_all(sampleid, '-', '_')) %>% 
#   left_join(metadata_ready, by = c('sampleid' = 'treatment_code')) %>% 
#   filter(month != 'May_22') %>% 
#   dplyr::select(!c(SampleID_metaT, SampleID_RP, SampleID_HILIC, site, sampling,
#                    month, stage, Month)) %>%
#   column_to_rownames(var = 'sampleid')
#   

# Metatranscriptomics

expression_matrix <- as.data.frame(t(expr_filt)) %>%
  rownames_to_column(var = 'sampleid') %>%
  left_join(metadata_ready, by = c('sampleid' = 'treatment_code')) %>%
  filter(month != 'May_22') %>%
  dplyr::select(!c(SampleID_metaT, SampleID_RP, SampleID_HILIC, site, sampling,
                   month, stage, Month)) %>%
  column_to_rownames(var = 'sampleid')


```


########################
# Model by dry and wet
########################

```{r}

## Creating multiomic matrix

X <- list(metabolomics = metabolomics_mx,
          metaT = expression_matrix)

## Y matrix - metadata, using month

metadata_int <- metadata_ready %>% 
  filter(month != 'May_22') %>% 
  filter(!is.na(SampleID_metaT)) %>% 
  dplyr::select(treatment_code, month, stage) %>% 
  mutate(moisture = if_else(stage %in% c("During_monsoon"), 'wet', 'dry'))

Y <- factor(metadata_int$moisture)


```

```{r}
# Matrix design - data driven

# Regression analysis with PLS to understand correlation bt omics

## RP vs metaG

# res1.pls <- pls(X$metabolomics, X$metaG, ncomp = 1)
# 
# cor(res1.pls$variates$X, res1.pls$variates$Y)

## RP vs metaT

res2.pls <- pls(X$metabolomics, X$metaT, ncomp = 1)

cor(res2.pls$variates$X, res2.pls$variates$Y)

## metaT vs metaG

# res3.pls <- pls(X$metaG, X$metaT, ncomp = 1)
# 
# cor(res3.pls$variates$X, res3.pls$variates$Y)

## RP vs metaG
#           comp1
# comp1 0.7204116

## RP vs metaT

#           comp1
# comp1 0.8783546

## metaG vs metaT
#           comp1
# comp1 0.5597737


## Matrix design


design <- matrix(0.8, ncol = length(X), nrow = length(X), 
                 dimnames = list(names(X), names(X)))

diag(design) <- 0

design 


```

# Number of components

```{r}

diablo <- block.plsda(X, Y, ncomp = 5, design = design)

set.seed(123) # For reproducibility, remove for your analyses

perf.diablo = perf(diablo, validation = c('Mfold'), folds = 5, nrepeat = 50)

perf.diablo$error.rate  # Lists the different types of error rates

# Plot of the error rates based on weighted vote
plot(perf.diablo)

## Component

perf.diablo$choice.ncomp$WeightedVote

ncomp <- 3 #

```



## Tunning

```{r}


set.seed(456) # for reproducibility


list.test.keepX <- list(
  metaT = seq(200, 250, 10),
  metabolomics = seq(100, 250, 10)
)


tune.diablo <- tune.block.splsda(X, Y, ncomp = ncomp, 
                                 test.keepX = list.test.keepX, 
                                 design = design,
                                 validation = 'Mfold', folds = 5, 
                                 nrepeat = 50, 
                                 BPPARAM = BiocParallel::SnowParam(workers = 12),
                                 dist = "centroids.dist",
                                 measure = 'BER')

list.keepX_res <- tune.diablo$choice.keepX

list.keepX_res


```

# Final model

```{r}

diablo_moisture_final <- block.splsda(X, Y, ncomp = 3, 
                                      keepX = list.keepX_res, design = design)

```

# Model performance

```{r}

## Mfold test 
set.seed(456) # For reproducibility


perf.diablo_final = perf(diablo_moisture_final, validation = 'Mfold', 
                         folds = 5, nrepeat = 100, dist = "centroids.dist")

# Performance with Majority vote
perf.diablo_final$MajorityVote.error.rate

# Performance with Weighted vote
perf.diablo_final$WeightedVote.error.rate

## LOO test 
set.seed(456) # For reproducibility


perf.diablo_final = perf(diablo_moisture_final, validation = 'loo', 
                         nrepeat = 10, dist = "centroids.dist")

# Performance with Majority vote
perf.diablo_final$MajorityVote.error.rate

# Performance with Weighted vote
perf.diablo_final$WeightedVote.error.rate

```


```{r}
plot(perf.diablo_final)

perf.diablo_final$WeightedVote.error.rate

auc <- auroc(diablo_moisture_final, roc.block = "metaT", roc.comp = 3,
             print = FALSE)

auc

plotArrow(diablo_moisture_final, ind.names = FALSE, legend = TRUE, 
          title = 'DIABLO comp 1 - 2')

```

## Plots

```{r}

plotDiablo(diablo_moisture_final, ncomp = 1)

scores <- plotIndiv(diablo_moisture_final, ind.names = TRUE, legend = TRUE,
                    ellipse = TRUE,
                    block = 'weighted.average',
                    title = 'OMICs, DIABLO')

## Extracting PCA values

scores_df <- scores$df

write_csv(scores_df, "output/Diablo_scores_11_10_25.csv")

saveRDS(diablo_moisture_final, "output/diablo_moisture_final_11_10_25.rds")

## Check plots

plotVar(diablo_moisture_final)

```

# Test for overfitting

```{r}

nperm <- 100
perm_errors <- data.frame(iteration = 1:nperm,
                          metaT_comp1 = rep(NA, nperm),
                          metaT_comp2 = rep(NA, nperm),
                          metaT_comp3 = rep(NA, nperm),
                          metaB_comp1 = rep(NA, nperm),
                          metaB_comp2 = rep(NA, nperm),
                          metaB_comp3 = rep(NA, nperm))

for(i in 1:nperm){
  Y_perm <- sample(Y)  # shuffle labels
  diablo_perm <- block.splsda(X, Y_perm, ncomp = 3,
                              keepX = list.keepX_res, 
                              design = design)
  
  perf_perm <- perf(diablo_perm, validation = 'Mfold', folds = 5, 
                    nrepeat = 10, dist = 'centroids.dist',
                    cpus = 12)
  
  perm_errors$metaT_comp1[i] <- perf_perm$error.rate$metaT[1]
  perm_errors$metaT_comp2[i] <- perf_perm$error.rate$metaT[2]
  perm_errors$metaT_comp3[i] <- perf_perm$error.rate$metaT[3]
  perm_errors$metaB_comp1[i] <- perf_perm$error.rate$metabolomics[1]
  perm_errors$metaB_comp2[i] <- perf_perm$error.rate$metabolomics[2]
  perm_errors$metaB_comp3[i] <- perf_perm$error.rate$metabolomics[3]
}


perf_perm$WeightedVote.error.rate

## Error rate null model plot

error_null <- perm_errors %>% 
  pivot_longer(!iteration, names_to = 'omic_component', values_to = 'error rate') %>% 
  ggplot() +
  ggridges::geom_density_ridges(aes(x= `error rate`,
                                    y = omic_component,
                                    fill = omic_component),
                                alpha = 0.5) +
  geom_vline(xintercept = 0.4, color = "red", lty = 2)+
  scale_y_discrete(limits = rev) +
  #scale_color_manual(values = assembly_color) +
  labs(title = 'Permutation test for DIABLO',
       y = 'Omics - component') +
  theme_bw() +
  theme(plot.title = element_text(face = 'bold', hjust = 0.5)) 

error_null



```

# Test for stability
## Check if the same features are consistently selected across folds

```{r}

## Increasing nrepeat to 50

perf.diablo_final = perf(diablo_moisture_final, validation = 'Mfold', 
                         folds = 5, nrepeat = 50, dist = "centroids.dist")

stability <- imap(perf.diablo_final$features$stable, function(obj_rep, rep){
  
  dataset_level <- imap(obj_rep, function(obj_dataset, dataset){
    
    component_level <- imap(obj_dataset, function(obj_comp, comp){
      
      res <- as.data.frame(obj_comp) %>% 
        rename(FeatureID = Var1) %>% 
        mutate(component = comp)
      
    }) %>% 
      reduce(rbind) %>% 
      mutate(dataset = dataset)
    
  }) %>% 
    reduce(rbind) %>% 
    mutate(rep = rep)
  
}) %>% reduce(rbind)

```

## Extracting variables

```{r}

# Metabolomics

loadings_df_comp1 <- plotLoadings(diablo_moisture_final, contrib = 'max', comp = 1) %>% 
  reduce(rbind) %>% 
  rownames_to_column(var = 'FeatureID') %>% 
  mutate(component = 'comp1')

loadings_df_comp2 <- plotLoadings(diablo_moisture_final, contrib = 'max', comp = 2) %>% 
  reduce(rbind) %>% 
  rownames_to_column(var = 'FeatureID') %>% 
  mutate(component = 'comp2')

loadings_df_comp3 <- plotLoadings(diablo_moisture_final, contrib = 'max', comp = 3) %>% 
  reduce(rbind) %>% 
  rownames_to_column(var = 'FeatureID') %>% 
  mutate(component = 'comp3')

loadings_all <- rbind(loadings_df_comp1, loadings_df_comp2, loadings_df_comp3)

loadings_unique <- loadings_all %>% 
  select(FeatureID, component) %>% 
  mutate(selected = TRUE)


stability_summarized <- stability %>% 
  filter(Freq > 0.5) %>% 
  left_join(loadings_unique, by = c('FeatureID', 'component')) %>% 
  mutate(selected = ifelse(is.na(selected), FALSE, selected),
         FeatureID_check = ifelse(selected, 
                                  glue::glue("<i style='color:red'>{FeatureID}</i>"),
                                  FeatureID)) %>% 
  ggplot() +
  geom_tile(aes(y = FeatureID_check,
                x = rep),
            fill = 'steelblue') +
  facet_grid2(rows = vars(dataset, component),
              space = 'free',
              scale = 'free') +
  theme_bw() +
  theme(axis.text.y = element_markdown(size = 3),
        axis.text.x = element_text(angle = 45, hjust = 1))


stability_summarized


write_csv(loadings_all, "output/DIABLO_selected_features_11_10_25.csv")

```


## Joining with annotations

# Metabolomics

```{r}

metabolomics_features <- loadings_all %>% 
  filter(str_detect(FeatureID, 'FT_')) %>% 
  left_join(annot_metabo, by = 'FeatureID')

ko_pathway <- read_csv("input/ko_pathways_UPDATE_10_23_24.csv")

metag_features <- loadings_all %>% 
  filter(!str_detect(FeatureID, 'FT_')) %>% 
  separate_wider_delim(FeatureID, names = c('MAG', 'KO'), delim = '__') %>% 
  left_join(mag_taxonomy, by = c('MAG' = 'bin')) %>% 
  left_join(ko_pathway, by = c('KO' = 'ko_id'))

write_csv(metabolomics_features, 'output/DIABLO_metabolomics_features_11-10-2025.csv')

write_csv(metag_features, 'output/DIABLO_metaG_features_11-10-2025.csv')

```


